### 2024-11-30 background.tex
init commit

\chapter{Background and Related Work}
\label{background}

Background 123

% \begin{figure*}
% \centering
% \includegraphics[width=0.98\textwidth]{Chapter1/Images/Fig1.pdf}
% \caption{caption}
% \label{figCh1:figure}
% \end{figure*}

\renewcommand\thefigure{\thechapter .\arabic{figure}}

### 2024-12-01 introduction.tex
init commit
Stage work on drafting Introduction, Motivation parts

--- 2024-11-30 background.tex
@@
-\chapter{Background and Related Work}
-\label{background}
+\chapter[Introduction]{Chapter 1 Introduction}
+\label{chap:1}
 
-Background 123
+Formal verification has become an essential approach in modern mathematics and computer science; it provides a strong framework for proving mathematical proofs and algorithms.
+Traditional techniques of formal verification include the following: Model Checking \cite{clarke2018handbook}, Testing \cite{broy2005model}, and Theorem Proving \cite{bertot2013interactive}.
 
-% \begin{figure*}
-% \centering
-% \includegraphics[width=0.98\textwidth]{Chapter1/Images/Fig1.pdf}
-% \caption{caption}
-% \label{figCh1:figure}
-% \end{figure*}
+Model Checking (MC) is a well-valued static formal verification method since, when applied correctly, it provides complete assurance that something satisfies certain specifications.
+Model checking is a completely automated procedure and is considered easier compared to other techniques such as theorem proving.
+However, because of the state explosion problem \cite{burch1992symbolic}, application of model checking in large systems is quite a challenge due to the immense state space size of the systems analyzed.
+Also, the success of model checking strongly depends on the quality of the analyzed models.
+In contrast, testing, despite being highly valued for revealing defects in systems, is inherently limited, since it can only indicate the presence of errors, but lacks the ability to verify the correctness of a system.
 
-\renewcommand\thefigure{\thechapter .\arabic{figure}}
+Theorem proving is another widely used formal analysis technique.
+It does not suffer from the limitations of state space size as model checking does, which allows the analysis of larger and more complex systems.
+Furthermore, theorem provers use very expressive logics, like first-order or higher-order logics, which enable the study of a wider range of systems without the restrictions that often come with modeling.
+The most prominent provers are HOL4 \cite{slind2008brief}, HOL Light \cite{hol_light}, Coq \cite{bertot2013interactive}, and Isabelle/HOL \cite{isabelle_hol}, all broadly used in the community.
+
+Formalization plays a vital role in probability theory to ensure that basic results are rigorous mathematically and computable in automated reasoning systems.
+This project formalizes the Central Limit Theorem using the HOL4 proof assistant, hence enriching the domain of probability in formalized mathematics.
+
+\section{Motivation}
+The Central Limit Theorem is one of the most fundamental results in the Probability Theory.
+This theory finds broad applications in statistics, data science, finance, and engineering.
+The theorem underlines that under certain conditions, the sum of many independent random variables will, as the number goes large, converge in distribution to a normal distribution irrespective of the individual distributions of those variables \cite{chung2000course}.
+This property justifies many real-world applications and models.
+
+Although HOL4 already contains formal proofs of some fundamental probabilistic results, such as the Law of Large Numbers, a formalized proof of the Central Limit Theorem is still absent.
+Completing a proof would realize several important milestones, including the following:
+
+\begin{itemize}
+    \item \textbf{Enriching the HOL4 Theory Library}:
+    \begin{itemize}
+        \item The gap in the library would be filled by adding some advanced results from probability theory.
+        \item A more general foundation would be laid for future processes of formal verification.
+    \end{itemize}
+    \item \textbf{Supporting Future Applications}:
+    \begin{itemize}
+        \item Strengthening the framework for developers and researchers to model or verify real-world problems involving probabilistic reasoning.
+    \end{itemize}
+    \item \textbf{Providing Analytical Tools}:        
+        \begin{itemize}
+            \item Enable efficient analysis of problems dependent on the normal distribution and its characteristics.
+        \end{itemize}   
+\end{itemize}
+
+Formalizing the Central Limit Theorem also aligns with ongoing efforts to bridge traditional mathematical insights with computational tools.
+As a consequence, it paves the way for applications in areas such as artificial intelligence, machine learning, and quantitative modeling, where probabilistic reasoning is increasingly critical.
+
+This project will contribute to a stronger, more versatile foundation for probabilistic formalization, empowering researchers and developers with better tools to tackle complex, real-world challenges.

### 2024-12-01 background.tex
Stage work on drafting Background parts, minor change in introduction.tex

@@
 
-Background 123
+The history of the Central Limit Theorem, hereinafter referred to as CLT, is quite a fascinating journey through the gradual development of probability theory.
+Over the years, various versions of CLT have been formulated, reflecting its adaptability and foundational importance.
+Each such variation addresses certain mathematical challenges and applications, therefore enriching the broader understanding of convergence in probability.
+In this chapter, we explore some theorem-proving tools and their methodologies to formalize the CLT, emphasizing ...
 
-% \begin{figure*}
-% \centering
-% \includegraphics[width=0.98\textwidth]{Chapter1/Images/Fig1.pdf}
-% \caption{caption}
-% \label{figCh1:figure}
-% \end{figure*}
+\section{Background}
+Let us brief the few evolvement of CLT, mainly noticing its critical features and variants developed further.
+The first steps towards CLT took place during the 18th century when Abraham de Moivre showed that, under summing many independent, identically distributed random variables-large number of rolled dice or coin flips-results in the normal distribution-a distribution with the shape of a bell \cite{de1733approximatio}.
+De Moivre laid down the very background for some approximations in the Binomial Distribution.
 
-\renewcommand\thefigure{\thechapter .\arabic{figure}}
+In 1810, Pierre-Simon Laplace expanded de Moivre’s insights by proving that the sum of independent random variables converges to a normal distribution under broader conditions \cite{laplace1835oeuvres}.
+His work established the CLT as a universal principle for analyzing aggregate phenomena, such as population averages and measurement errors.
+
+In the 19th century, mathematicians such as Pafnuty Chebyshev formalized the conditions of the theorem, including those of variance and expectation, so that the theorem became more precise and mathematically sound \cite{chebyshev1890deux}.
+At the beginning of the 20th century, Lyapunov generalized CLT by introducing specific criteria-Lyapunov's condition-which explained when the theorem was applicable \cite{lyapunov1895pafnutii}.
+William Feller \cite{feller1945} later refined it to address discrete random variables.
+
+Developments in the modern era since World War II have broadened the scope of the CLT beyond normal distributions to stable distributions and applications involving stochastic processes and high-dimensional data.
+These extensions illustrate how the theorem can be adapted to a wide range of probabilistic and statistical contexts.
+
+Today, the CLT comes in a variety of variants each suited to particular situations:
+\begin{itemize}
+    \item \textbf{Classical CLT:} For sums of independent, identically distributed random variables possessing finite variance and mean.
+    \item \textbf{Generalized CLT:} When variables are weakly dependent or not identically distributed.
+    \item \textbf{Triangular Arrays:} For sums of random variables arranged in arrays where the conditions vary across rows.
+    \item \textbf{Local and Integral Versions:} While the local CLT concerns pointwise probabilities, the integral version deals with cumulative distributions.
+\end{itemize}
+From about 1810 to 1935, most of the efforts were devoted to proving the CLT for sums of independent random variables, with more recent generalizations involving only weakly dependent variables.
+Modern formulations neatly distinguish between normed sums, triangular arrays, and local versus integral theorems.

### 2024-12-03 introduction.tex
Stage work on drafting Background parts, minor change in introduction.tex
Fixed pdf generation, citation

@@
-\chapter[Introduction]{Chapter 1 Introduction}
-\label{chap:1}
+\chapter{Introduction}
+\label{intro}
 
@@
 Furthermore, theorem provers use very expressive logics, like first-order or higher-order logics, which enable the study of a wider range of systems without the restrictions that often come with modeling.
-The most prominent provers are HOL4 \cite{slind2008brief}, HOL Light \cite{hol_light}, Coq \cite{bertot2013interactive}, and Isabelle/HOL \cite{isabelle_hol}, all broadly used in the community.
+The most prominent provers are HOL4 \cite{slind2008brief}, HOL Light \cite{hollight}, Coq \cite{bertot2013interactive}, and Isabelle/HOL \cite{isabelle}, all broadly used in the community.
 

### 2024-12-03 draft.tex
Stage work on draft ideas

\section*{Draft ideas}
\subsection{Introduction}
\cite{tian2022assumption}
\cite{qasim2016formalization}

\subsubsection{Motivation}
\begin{itemize}
    \item Enriching the HOL4 Theory Library
    \item Supporting Future Applications
    \item Providing Analytical Tools
\end{itemize}

\subsubsection{Contributions}
\begin{itemize}
    \item This thesis proves the Lyapunov version of the Central Limit Theorem, a fundamental result in probability theory. The CLT of Lyapunov extends the classical CLT by relaxing the identical distribution assumption to the case of independent but not necessarily identically distributed random variables.
    \item The proof makes use of Taylor expansions, moment-based bounds, and asymptotic error analysis to rigorously establish convergence to the normal distribution.
    \item The main insights involve the Lyapunov condition, which guarantees that the third absolute moments become small compared to the variance, and a new use of Big-O notation for the effective control of error terms.
\end{itemize}

\subsubsection{Results}
\begin{itemize}
    \item \textbf{Lyapunov CLT}: For independent random variables \(X_1, X_2, \ldots, X_n \) with finite means, variances, and third moments, if the Lyapunov condition is satisfied:
$$
\frac{\Gamma_n}{s_n^3} \to 0 \quad \text{as } n \to \infty,
$$
then the normalized sum \( \frac{S_n - \mu}{s_n} \) converges in distribution to the standard normal distribution \( N(0, 1) \).
    \item \textbf{Proof}: It establishes that
    \begin{itemize}
        \item Error bounds of the test function \(f\) by Taylor expansion and moment inequalities.
        \item Asymptotic vanishing of higher-order terms under the Lyapunov condition.  \item A rigorous comparison between the distribution of \( \frac{S_n}{s_n} \) and \( N(0, 1)\), bounded by \( O\left(\frac{\Gamma_n}{s_n^3}\right) \).
    \end{itemize}

These results illustrate the robustness of the Lyapunov CLT against heterogeneous distributions and, in fact, form a bridge between classical formulations and modern applications.
\end{itemize}

\subsection{Background and Related Works}
\subsubsection{Background}
Different versions of CLT \cite{fischer2011history}
\begin{itemize}
    \item \textbf{Classical Central Limit Theorem} \cite{ross2019first}
    \begin{itemize}
        \item \textbf{Proven by:} De Moivre, Laplace
        \item \textbf{Statement:} If \( X_1, X_2, \ldots, X_n \) are independent and identically distributed (i.i.d.) random variables with finite mean \( \mu \) and variance \( \sigma^2 \), then:
  \[
  Z_n = \frac{\sum_{i=1}^n X_i - n\mu}{\sigma\sqrt{n} } \xrightarrow{d} N(0, 1).
  \]
  \item \textbf{Significance:} Foundation of probability theory and statistics, used in inferential statistics, e.g., hypothesis testing and confidence intervals.
    \end{itemize}

    \item \textbf{Lyapunov’s Central Limit Theorem} \cite{chung2000course}
    \item \textbf{Lindeberg’s Central Limit Theorem}
    \item \textbf{The Generalized Central Limit Theorem} 
    \begin{itemize}
        \item https://en.wikipedia.org/wiki/Central_limit_theorem
    \end{itemize}

\end{itemize}

\subsubsection{Related Work}
The CLT is a fundamental result in probability theory, and its formalization in theorem provers has been a highly active area of interest in the formal methods community. Several different ways of proving the CLT have been tried, and tools like Isabelle/HOL and HOL4 provide different approaches reflecting their different frameworks and mathematical libraries.

\textbf{Isabelle/HOL's Formalization of the CLT}
In Isabelle/HOL, the CLT has been formalized in a powerful and especially elegant way using the \textbf{characteristic function approach}, which relies on the fact that characteristic functions uniquely determine probability distributions and their pointwise convergence implies convergence in distribution.

\begin{itemize}
    \item \textbf{Methodology}
    \item \textbf{Characteristic Function Approach}
    \begin{itemize}
        \item It provides a direct and compact route to proving the CLT, leveraging the algebraic simplicity of characteristic functions.
        \item ...
    \end{itemize}
\end{itemize}

\subsection{Premilaires}
\subsubsection{HOL4}
\cite{tian2022assumption}
\subsubsection{Measure Theory}
\cite{qasim2016formalization}
\subsubsection{Lebesgue Integration Theory}
\cite{qasim2016formalization}
\subsubsection{Probability Theory}
\cite{qasim2016formalization}

\subsection{Central Limit Theorem}
Informal proof of Lyapounov's theorem for a single sequence by the method of Lindeberg \cite{chung2000course}.
\begin{itemize}
    \item \textbf{Notation and Statement}
    \begin{enumerate}
    \item \( \{X_j\}_{j=1}^n \) be a sequence of independent random variables with:
    \begin{itemize}
        \item[] \( \mathbb{E}[X_j] = 0 \), 
        \item[] \( \text{Var}(X_j) = \sigma_j^2 \) (finite variances, \( \sigma_j^2 < \infty \)),
        \item[] \( \mathbb{E}[|X_j|^3] = \gamma_j \) (finite third moments, \( \gamma_j < \infty \)).
    \end{itemize}

        \item Define:
        \begin{itemize}
            \item[] \( S_n = \sum_{j=1}^n X_j \),
            \item[] \( s_n^2 = \sum_{j=1}^n \sigma_j^2 \) (total variance),
            \item[] \( \Gamma_n = \sum_{j=1}^n \gamma_j \) (total third moment).
        \end{itemize}

        \item Statement:
    If 
    \[
   \frac{\Gamma_n}{s_n^3} \to 0 \quad \text{as } n \to \infty.
   \]
   then \( \frac{S_n}{s_n} \xrightarrow{d} \mathcal{N}(0,1) \),
   where \( \mathcal{N}(0,1) \) denotes the standard normal distribution.

    \end{enumerate}
    \item \textbf{Step-by-Step Proof}
    \begin{enumerate}
        \item \textbf{Normalize the Random Variables}

        Define the normalized variables:
\[
X_{n,j} = \frac{X_j}{s_n}, \quad \text{for } j = 1, \dots, n.
\]
The sum \( S_n \) is now expressed as:
\[
\frac{S_n}{s_n} = \sum_{j=1}^n X_{n,j}.
\]

\item \textbf{Use the Lindeberg Replacement to approximate}

Replace each \( X_j \) with a corresponding normal random variable \( Y_j \sim \mathcal{N}(0, \sigma_j^2) \), where \( \{Y_j\}_{j=1}^n \) are independent and have the same mean and variance as \( X_j \), then satisfy:
\[
\mathbb{E}[Y_j] = 0 \quad \text{and} \quad \text{Var}(Y_j) = \sigma_j^2.
\]

Let all the \( X\)'s and \( Y\)'s be totally independent.

Now, construct the sequence:
\[
Z_j = Y_1 + \dots + Y_{j-1} + X_j + \dots + X_n, \quad 1 \leq j \leq n.
\]
Thus:
\begin{itemize}
    \item[] \( Z_1 = X_2 + X_3 + \dots + X_n \),
    \item[] \( Z_2 = Y_1 + X_3 + \dots + X_n \),
    \item[] \( Z_n = Y_1 + Y_2 + \dots + Y_{n-1} \).
\end{itemize}

Each \( Z_j \) represents a sum that is partially replaced. All variables before \( X_j \) are replaced by \( Y_j \), but the variable after \( X_j \) remains as \( X_j \).

\item \textbf{Compare Distributions}

To show that \( \frac{S_n}{s_n} \xrightarrow{d} \mathcal{N}(0, 1) \), we compare the expectations of \( \frac{S_n}{s_n} \) with those of \( \frac{Z_j}{s_n} \).

***** cont. *****

    \end{enumerate}
\end{itemize}

\subsection{Conclusions}
The CLT is a fundamental result in probability theory, ranging from mathematics and statistics to finance and data science. This work described the formal proof of the CLT using Lyapunov's criterion in the HOL4 theorem prover. The proof showed that the normalized sum of independent (but not necessarily identically distributed) random variables converges in distribution to the standard normal distribution by exploiting a moment-based approach along with Lyapunov's condition.

\textbf{Conclusions}
\begin{itemize}
    \item \textbf{Formalization of Lyapunov's CLT}
    \begin{itemize}
        \item The proof formalized Lyapunov's version of the CLT by providing rigorous bounds on error terms and establishing convergence under the Lyapunov condition. The use of Big-O notation and asymptotic error analysis was instrumental in managing higher-order terms.
    \end{itemize}

    \item \textbf{Theoretical Insights}
    \begin{itemize}
        \item In addition, a constructive, moment-based methodology used in its proof gave a natural perspective on finite variances and higher-order moments, each playing an important part in ensuring convergence according to CLT.
    \end{itemize}

    \item \textbf{Comparison with Previous Works}
    \begin{itemize}
        \item The approach represents a complementary alternative to existing formalizations, such as \textbf{Isabelle/HOL's characteristic function-based proof}, in that it represents a constructive proof based on explicit bounds and moment inequalities. It is also an addition to the variety of formalizations and expands the reach of theorem provers regarding probabilistic results.
    \end{itemize}

\end{itemize}
Formalization of CLT in HOL4 showcases the power of theorem provers in the formal verification of fundamental results in probability theory. It fills the gap between theoretical results and their verification by computation, showing the potential of formal methods in academic and applied domains.

\subsubsection{Future Directions}
...

In closing, this thesis provides not only a sound formalization of Lyapunov's CLT but also the whole range of insights into what happens when formal methods are integrated with probability theory. It lays the ground for further investigations into probabilistic results within theorem provers, opening perspectives to theoretical advances as well as practical applications.

### 2024-12-04 draft.tex
Done informal proof of CLT

@@
 
+\newpage
 \subsection{Central Limit Theorem}
@@
     \[
-   \frac{\Gamma_n}{s_n^3} \to 0 \quad \text{as } n \to \infty.
+   \frac{\Gamma_n}{s_n^3} \to 0 \quad \text{as } n \to \infty
    \]
-   then \( \frac{S_n}{s_n} \xrightarrow{d} \mathcal{N}(0,1) \),
-   where \( \mathcal{N}(0,1) \) denotes the standard normal distribution.
+   then 
+   \[
+   \frac{S_n}{s_n} \xrightarrow{d} \Phi
+   \]
+   where \( \Phi\) denotes the standard normal distribution \( \mathcal{N}(0,1) \).
 
@@
 
-Now, construct the sequence:
+Since  \( Y_j \sim \mathcal{N}(0, \sigma_j^2) \), the normalized random variable:
 \[
-Z_j = Y_1 + \dots + Y_{j-1} + X_j + \dots + X_n, \quad 1 \leq j \leq n.
+\frac{Y_j}{s_n} \sim \mathcal{N}\left(0, \frac{\sigma_j^2}{s_n^2}\right)
+\]
+
+Adding these normalized variables gives:
+\[
+\frac{1}{s_n} \sum_{j=1}^n Y_j \sim \mathcal{N}\left(0, \sum_{j=1}^n \frac{\sigma_j^2}{s_n^2}\right).
+\]
+
+Since \( s_n^2 = \sum_{j=1}^n \sigma_j^2 \), this simplifies to:
+\[
+\frac{1}{s_n} \sum_{j=1}^n Y_j \sim \mathcal{N}(0, 1).
+\]
+
+Now, construct the sequence \( Z\):
+\[
+Z_j = Y_1 + \dots + Y_{j-1} + X_j + \dots + X_n, \quad 1 \leq j \leq n
 \]
@@
     \item[] \( Z_2 = Y_1 + X_3 + \dots + X_n \),
-    \item[] \( Z_n = Y_1 + Y_2 + \dots + Y_{n-1} \).
+    \item[] \( Z_n = Y_1 + Y_2 + \dots + Y_{n-1} \)
 \end{itemize}
@@
 
-To show that \( \frac{S_n}{s_n} \xrightarrow{d} \mathcal{N}(0, 1) \), we compare the expectations of \( \frac{S_n}{s_n} \) with those of \( \frac{Z_j}{s_n} \).
+To show that \( \frac{S_n}{s_n} \xrightarrow{d} \mathcal{N}(0, 1) \), we compare their expectations.
 
-***** cont. *****
+\textbf{Using Test Funtions}
+
+Consider a test function \( f \) from \( C^3_B \), the class of bounded continuous functions with three bounded derivatives.
+
+Consider:
+\[
+\mathbb{E}\left[f\left(\frac{S_n}{s_n}\right)\right] \quad \text{and} \quad \mathbb{E}\left[f\left(\mathcal{N}\right)\right],
+\]
+
+By introducing the replacement sequence \( Z_j \), we rewrite:
+
+\[
+\mathbb{E}\left[f\left(\frac{S_n}{s_n}\right)\right] - \mathbb{E}\left[f\left(\frac{Z_n}{s_n}\right)\right].
+\]
+
+By telescoping:
+\[
+\mathbb{E}\left[f\left(\frac{S_n}{s_n}\right)\right] - \mathbb{E}\left[f\left(\frac{Z_n}{s_n}\right)\right] = \sum_{j=1}^n \left[\mathbb{E}\left[f\left(\frac{X_j + Z_j}{s_n}\right)\right] - \mathbb{E}\left[f\left(\frac{Y_j + Z_{j+1}}{s_n}\right)\right]\right]
+\]
+
+\item \textbf{Taylor Expansion of \( f \)}
+
+By Taylor's theorem for \( f \in C^3_B \):
+\[
+\left|f(x + y) - \left[f(x) + f'(x)y + \frac{f''(x)}{2}y^2\right]\right| \leq \frac{M}{6}|y|^3,
+\]
+where \( M = \sup_{x \in \mathbb{R}} |f^{(3)}(x)| \) is the maximum bound on the third derivative of \( f \).
+
+Apply this to \( f(\xi + \eta) \) and take Expectations, we write:
+\[
+\left| \mathbb{E}[f(\xi + \eta)] - \mathbb{E}[f(\xi)] - \mathbb{E}[f'(\xi)]\mathbb{E}[\eta] - \frac{1}{2}\mathbb{E}[f''(\xi)]\mathbb{E}[\eta^2] \right| \leq \frac{M}{6} \mathbb{E}[|\eta|^3],
+\]
+where \( \xi \) and \( \eta \) are independent random variables such that \(\mathbb{E}[|\eta|^3] < \infty \).
+
+Then, let \( \zeta \) be another independent random variable, having the same mean and variance as \( \eta \). Now, consider:
+     \[
+     \mathbb{E}[f(\xi + \eta)] - \mathbb{E}[f(\xi + \zeta)].
+     \]
+
+Using Taylor's theorem again, this difference can be bounded as:
+     \[
+     \left| \mathbb{E}[f(\xi + \eta)] - \mathbb{E}[f(\xi + \zeta)] \right| \leq \frac{M}{6} (\mathbb{E}[|\eta|^3] + \mathbb{E}[|\zeta|^3]),
+     \]
+     where the sum \( \mathbb{E}[|\eta|^3] + \mathbb{E}[|\zeta|^3] \) accounts for the third absolute moments of both random variables.
+
+Then, substitute with:
+\begin{itemize}
+    \item \( \xi = Z_j / s_n \),
+    \item \( \eta = X_j / s_n \),
+    \item \( \zeta = Y_j / s_n \).
+\end{itemize}
+We have:
+   \begin{equation} \label{eq:1}
+       \left| \mathbb{E}\left[f\left(\frac{Z_j}{s_n} + \frac{X_j}{s_n}\right)\right] - \mathbb{E}\left[f\left(\frac{Z_j}{s_n} + \frac{Y_j}{s_n}\right)\right] \right| \leq \frac{M}{6} \left( \mathbb{E}\left[\left| \frac{X_j}{s_n} \right|^3\right] + \mathbb{E}\left[\left| \frac{Y_j}{s_n} \right|^3\right] \right)
+   \end{equation}
+
+\item \textbf{Error Bound Using Lyapunov's Condition}
+
+From the definition of \( X_j / s_n \) and \( Y_j / s_n \), we know:
+     \[
+     \mathbb{E}\left[\left|\frac{X_j}{s_n}\right|^3\right] = \frac{\mathbb{E}[|X_j|^3]}{s_n^3}, \quad \mathbb{E}\left[\left|\frac{Y_j}{s_n}\right|^3\right] = \frac{\mathbb{E}[|Y_j|^3]}{s_n^3}.
+     \]
+
+For \( Y_j \sim \mathcal{N}(0, \sigma_j^2) \), the third absolute moment is:
+     \[
+     \mathbb{E}[|Y_j|^3] = c \sigma_j^3,
+     \]
+     where \( c = \sqrt{8 / \pi} \) (a constant depending on the normal distribution).
+
+Since \( \mathbb{E}[|X_j|^3] = \gamma_j \) and \(\mathbb{E}[|Y_j|^3] = c \sigma_j^3\), we wrire:
+     \[
+   \mathbb{E}\left[\left|\frac{X_j}{s_n}\right|^3\right] + \mathbb{E}\left[\left|\frac{Y_j}{s_n}\right|^3\right] = \frac{\gamma_j}{s_n^3} + \frac{c \sigma_j^3}{s_n^3}
+   \]
+
+By summing over \(j = 1, \dots , n\), \eqref{eq:1} becomes:
+\[
+\sum_{j=1}^n \left| \mathbb{E}\left[f\left(\frac{Z_j}{s_n} + \frac{X_j}{s_n}\right)\right] - \mathbb{E}\left[f\left(\frac{Z_j}{s_n} + \frac{Y_j}{s_n}\right)\right] \right| \leq \frac{M}{6} \sum_{j=1}^n \left( \frac{\gamma_j}{s_n^3} + \frac{c \sigma_j^3}{s_n^3} \right)
+\]
+
+Lyapunov's inequality states that:
+\[
+\sigma_j^3 \leq \gamma_j.
+\]
+Thus:
+\[
+\frac{\gamma_j}{s_n^3} + \frac{c \cdot \sigma_j^3}{s_n^3} \leq \frac{\gamma_j}{s_n^3} + \frac{c \cdot \gamma_j}{s_n^3} = \frac{(1 + c) \cdot \gamma_j}{s_n^3}.
+\]
+Simplify the Sum, we have:
+
+\begin{equation} \label{eq:2}
+\frac{M}{6} \sum_{j=1}^n \left( \frac{\gamma_j}{s_n^3} + \frac{c \cdot \sigma_j^3}{s_n^3} \right) \leq \frac{M}{6} \cdot \frac{1}{s_n^3} \sum_{j=1}^n (1 + c) \gamma_j
+\end{equation}
+
+Let \( K = \frac{M}{6}(1 + c) \), and the total third absolute moment of \( X_j \):
+\[
+\Gamma_n = \sum_{j=1}^n \gamma_j ( \space as \space define \space above)
+\]
+We substitute these into the inequality \eqref{eq:2}:
+\begin{equation} \label{eq:3}
+\frac{M}{6} \sum_{j=1}^n \left( \frac{\gamma_j}{s_n^3} + \frac{c \cdot \sigma_j^3}{s_n^3} \right) \leq K \cdot \frac{\Gamma_n}{s_n^3}
+\end{equation}
+
+By the definition of Big-O Notation, we have 
+\begin{displayquote}
+For a function \( f(n) \), we say that \( f(n) = O(g(n)) \) if there exist positive constants \( C > 0 \) and \( n_0 \geq 1 \) such that:
+\[
+|f(n)| \leq C |g(n)|, \quad \text{for all } n \geq n_0
+\]
+\end{displayquote}
+
+Here:
+\begin{itemize}
+    \item \( f(n) = \frac{\Gamma_n}{s_n^3} \),
+    \item \( g(n) = \frac{\Gamma_n}{s_n^3} \)
+\end{itemize}
+
+Clearly:
+\[
+|f(n)| = \left| \frac{\Gamma_n}{s_n^3} \right| \quad \text{and} \quad |g(n)| = \left| \frac{\Gamma_n}{s_n^3} \right|
+\]
+
+Since \( f(n) \) and \( g(n) \) are identical, we have:
+\[
+|f(n)| \leq 1 \cdot |g(n)| \quad \text{for all } n \geq 1
+\]
+
+Here, the constant \( C = 1 \). Then, the inequality \( |f(n)| \leq C |g(n)| \) holds for all \( n \geq 1 \), so we can choose \( n_0 = 1 \).
+
+Since we have found constants \( C = 1 \) and \( n_0 = 1 \) such that:
+\[
+|f(n)| \leq C |g(n)| \quad \text{for all } n \geq n_0,
+\]
+it follows by the definition of Big-O notation that:
+\begin{equation} \label{eq:4}
+\frac{\Gamma_n}{s_n^3} = O\left(\frac{\Gamma_n}{s_n^3}\right)
+\end{equation}
+
+The Big O Notation property called \textbf{Multiplication by a constant} states that:
+Let \( k \) be a nonzero constant. Then:
+\[
+O(|k| \cdot g) = O(g)
+\]
+Apply this to \eqref{eq:4}, we have:
+   \[
+   K \cdot \frac{\Gamma_n}{s_n^3} = O\left(\frac{\Gamma_n}{s_n^3}\right)
+   \]
+Then \eqref{eq:3} becomes:
+\[
+\frac{M}{6} \sum_{j=1}^n \left( \frac{\gamma_j}{s_n^3} + \frac{c \cdot \sigma_j^3}{s_n^3} \right) \leq O\left(\frac{\Gamma_n}{s_n^3}\right)
+\]
+
+We have thus obtained the following estimate:
+\[
+\forall f \in C^3: \left| \space  \mathbb{E}\left[f\left(\frac{S_n}{s_n}\right)\right] - \mathbb{E}\left[f\left(N\right)\right] \right| \leq O\left(\frac{\Gamma_n}{s_n^3}\right) 
+\]
+
+\item \textbf{Convergence to Normal}
+
+By Lyapunov's condition:
+\[
+\frac{\Gamma_n}{s_n^3} \to 0 \quad \text{as } n \to \infty
+\]
+This directly implies that:
+\[
+\left| \mathbb{E}\left[f\left(\frac{S_n}{s_n}\right)\right] - \mathbb{E}\left[f(N)\right] \right| \to 0.
+\]
+
+Since \( \mathbb{E}[f(S_n / s_n)] \to \mathbb{E}[f(N)] \) for all bounded and continuous test functions \( f \), by the general criterion for vague convergence, we conclude that:
+\[
+\frac{S_n}{s_n} \xrightarrow{d} \mathcal{N}(0, 1)
+\]
 

### 2024-12-05 draft.tex
Done draft ideas

@@
 \section*{Draft ideas}
+\subsection{Abstract}
+\begin{itemize}
+    \item What is the CLT? Why is formal verification important?
+    \item What limitations or gaps does this work address?
+    \item Briefly summarize the methods, results, and significance.
+\end{itemize}
+
 \subsection{Introduction}
@@
 \cite{qasim2016formalization}
+\begin{itemize}
+    \item What is CLT
+    \item Define the CLT in its simplest form (e.g., i.i.d. random variables)
+    \item Highlight its importance in statistics, machine learning, and other fields as a foundation for normal approximation.
+\end{itemize}
 
 \subsubsection{Motivation}
+Why formalize the CLT in HOL4? 
+The challenges with probabilistic reasoning, the need for rigor, and the importance of verifying a key theorem like the CLT.
 \begin{itemize}
@@
 \subsubsection{Results}
+This form of CLT is more general than the i.i.d case but not the most general (not the Lindeberg-Feller version).
+
 \begin{itemize}
@@
 These results illustrate the robustness of the Lyapunov CLT against heterogeneous distributions and, in fact, form a bridge between classical formulations and modern applications.
+
+Also, highlight key formal results of other lemmas and theorems to support (e.g., bounding errors using Big-O, weak convergence).
 \end{itemize}
@@
 \subsection{Background and Related Works}
-\subsubsection{Background}
-Different versions of CLT \cite{fischer2011history}
+\subsubsection{Historical Background}
+Introduce the histrory of CLT and mention different versions of CLT \cite{fischer2011history}
+\begin{itemize}
+    \item Classical versions: De Moivre-Laplace (i.i.d.), Lindeberg-Feller, Lyapunov’s theorem.
+    \item Explain why Lyapunov’s version is chosen (finite moments, no characteristic functions).
+\end{itemize}
+
 \begin{itemize}
@@
 In Isabelle/HOL, the CLT has been formalized in a powerful and especially elegant way using the \textbf{characteristic function approach}, which relies on the fact that characteristic functions uniquely determine probability distributions and their pointwise convergence implies convergence in distribution.
+\cite{serafin2015formally}
 
@@
 
-\subsection{Premilaires}
+\subsubsection{Overview of Informal Proof}
+\begin{itemize}
+    \item Highlight the direct and real-valued approach of Lyapunov's proof.
+    \item Discuss how it avoids characteristic functions and instead uses bounding error terms (Taylor expansion, Big-O).
+\end{itemize}
+
+\subsection{Preliminaries}
 \subsubsection{HOL4}
@@
 \newpage
-\subsection{Central Limit Theorem}
+\subsection{The proof of Central Limit Theorem}
+\subsubsection{Informal Proof}
 Informal proof of Lyapounov's theorem for a single sequence by the method of Lindeberg \cite{chung2000course}.
@@
     \[
-   \frac{\Gamma_n}{s_n^3} \to 0 \quad \text{as } n \to \infty
+   \frac{\Gamma_n}{s_n^3} \to 0 \quad \text{as } n \to \infty.
    \]
@@
 \[
-\frac{Y_j}{s_n} \sim \mathcal{N}\left(0, \frac{\sigma_j^2}{s_n^2}\right)
+\frac{Y_j}{s_n} \sim \mathcal{N}\left(0, \frac{\sigma_j^2}{s_n^2}\right).
 \]
@@
 \[
-Z_j = Y_1 + \dots + Y_{j-1} + X_j + \dots + X_n, \quad 1 \leq j \leq n
+Z_j = Y_1 + \dots + Y_{j-1} + X_{j + 1} + \dots + X_n, \quad 1 \leq j \leq n.
 \]
@@
     \item[] \( Z_2 = Y_1 + X_3 + \dots + X_n \),
-    \item[] \( Z_n = Y_1 + Y_2 + \dots + Y_{n-1} \)
+    \item[] \( Z_n = Y_1 + Y_2 + \dots + Y_{n-1} \).
 \end{itemize}
 
-Each \( Z_j \) represents a sum that is partially replaced. All variables before \( X_j \) are replaced by \( Y_j \), but the variable after \( X_j \) remains as \( X_j \).
+In general, each \( Z_j \) represents a sum where:
+\begin{itemize}
+    \item All variables before \( X_j \) are replaced by \( Y_1, \dots, Y_{j-1} \),
+    \item \( X_j \) is excluded from the summation,
+    \item All variables after \( X_j \) remain as \( X_{j+1}, X_{j+2}, \dots, X_n \).
+\end{itemize}
+
+Thus, by telescoping property, we have:
+\[
+Y_j + Z_j = X_{j+1} + Z_{j+1}.
+\]
 
@@
 \[
-\mathbb{E}\left[f\left(\frac{S_n}{s_n}\right)\right] - \mathbb{E}\left[f\left(\frac{Z_n}{s_n}\right)\right] = \sum_{j=1}^n \left[\mathbb{E}\left[f\left(\frac{X_j + Z_j}{s_n}\right)\right] - \mathbb{E}\left[f\left(\frac{Y_j + Z_{j+1}}{s_n}\right)\right]\right]
+\mathbb{E}\left[f\left(\frac{S_n}{s_n}\right)\right] - \mathbb{E}\left[f\left(\frac{Z_n}{s_n}\right)\right].
 \]
 
-\item \textbf{Taylor Expansion of \( f \)}
+\[
+= \mathbb{E}\left[f\left(\frac{X_1 + \cdots + X_n}{s_n}\right)\right] - \mathbb{E}\left[f\left(\frac{Y_1 + \cdots + Y_n}{s_n}\right)\right].
+\]
+
+\[
+= \sum_{j=1}^n \left[\mathbb{E}\left[f\left(\frac{X_j + Z_j}{s_n}\right)\right] - \mathbb{E}\left[f\left(\frac{Y_j + Z_j}{s_n}\right)\right]\right].
+\]
+
+\item \textbf{Expand expectations with Taylor’s theorem}
 
@@
    \begin{equation} \label{eq:1}
-       \left| \mathbb{E}\left[f\left(\frac{Z_j}{s_n} + \frac{X_j}{s_n}\right)\right] - \mathbb{E}\left[f\left(\frac{Z_j}{s_n} + \frac{Y_j}{s_n}\right)\right] \right| \leq \frac{M}{6} \left( \mathbb{E}\left[\left| \frac{X_j}{s_n} \right|^3\right] + \mathbb{E}\left[\left| \frac{Y_j}{s_n} \right|^3\right] \right)
+       \left| \mathbb{E}\left[f\left(\frac{Z_j}{s_n} + \frac{X_j}{s_n}\right)\right] - \mathbb{E}\left[f\left(\frac{Z_j}{s_n} + \frac{Y_j}{s_n}\right)\right] \right| \leq \frac{M}{6} \left( \mathbb{E}\left[\left| \frac{X_j}{s_n} \right|^3\right] + \mathbb{E}\left[\left| \frac{Y_j}{s_n} \right|^3\right] \right).
    \end{equation}
 
-\item \textbf{Error Bound Using Lyapunov's Condition}
+\item \textbf{Bound error terms using Big-O notation}
 
@@
      \[
-   \mathbb{E}\left[\left|\frac{X_j}{s_n}\right|^3\right] + \mathbb{E}\left[\left|\frac{Y_j}{s_n}\right|^3\right] = \frac{\gamma_j}{s_n^3} + \frac{c \sigma_j^3}{s_n^3}
+   \mathbb{E}\left[\left|\frac{X_j}{s_n}\right|^3\right] + \mathbb{E}\left[\left|\frac{Y_j}{s_n}\right|^3\right] = \frac{\gamma_j}{s_n^3} + \frac{c \sigma_j^3}{s_n^3}.
    \]
@@
 \[
-\sum_{j=1}^n \left| \mathbb{E}\left[f\left(\frac{Z_j}{s_n} + \frac{X_j}{s_n}\right)\right] - \mathbb{E}\left[f\left(\frac{Z_j}{s_n} + \frac{Y_j}{s_n}\right)\right] \right| \leq \frac{M}{6} \sum_{j=1}^n \left( \frac{\gamma_j}{s_n^3} + \frac{c \sigma_j^3}{s_n^3} \right)
+\sum_{j=1}^n \left| \mathbb{E}\left[f\left(\frac{Z_j}{s_n} + \frac{X_j}{s_n}\right)\right] - \mathbb{E}\left[f\left(\frac{Z_j}{s_n} + \frac{Y_j}{s_n}\right)\right] \right| \leq \frac{M}{6} \sum_{j=1}^n \left( \frac{\gamma_j}{s_n^3} + \frac{c \sigma_j^3}{s_n^3} \right).
 \]
@@
 \begin{equation} \label{eq:2}
-\frac{M}{6} \sum_{j=1}^n \left( \frac{\gamma_j}{s_n^3} + \frac{c \cdot \sigma_j^3}{s_n^3} \right) \leq \frac{M}{6} \cdot \frac{1}{s_n^3} \sum_{j=1}^n (1 + c) \gamma_j
+\frac{M}{6} \sum_{j=1}^n \left( \frac{\gamma_j}{s_n^3} + \frac{c \cdot \sigma_j^3}{s_n^3} \right) \leq \frac{M}{6} \cdot \frac{1}{s_n^3} \sum_{j=1}^n (1 + c) \gamma_j.
 \end{equation}
@@
 \[
-\Gamma_n = \sum_{j=1}^n \gamma_j ( \space as \space define \space above)
+\Gamma_n = \sum_{j=1}^n \gamma_j ( \space as \space define \space above).
 \]
@@
 \begin{equation} \label{eq:3}
-\frac{M}{6} \sum_{j=1}^n \left( \frac{\gamma_j}{s_n^3} + \frac{c \cdot \sigma_j^3}{s_n^3} \right) \leq K \cdot \frac{\Gamma_n}{s_n^3}
+\frac{M}{6} \sum_{j=1}^n \left( \frac{\gamma_j}{s_n^3} + \frac{c \cdot \sigma_j^3}{s_n^3} \right) \leq K \cdot \frac{\Gamma_n}{s_n^3}.
 \end{equation}
@@
 \[
-|f(n)| \leq C |g(n)|, \quad \text{for all } n \geq n_0
+|f(n)| \leq C |g(n)|, \quad \text{for all } n \geq n_0.
 \]
@@
     \item \( f(n) = \frac{\Gamma_n}{s_n^3} \),
-    \item \( g(n) = \frac{\Gamma_n}{s_n^3} \)
+    \item \( g(n) = \frac{\Gamma_n}{s_n^3} \).
 \end{itemize}
@@
 \[
-|f(n)| = \left| \frac{\Gamma_n}{s_n^3} \right| \quad \text{and} \quad |g(n)| = \left| \frac{\Gamma_n}{s_n^3} \right|
+|f(n)| = \left| \frac{\Gamma_n}{s_n^3} \right| \quad \text{and} \quad |g(n)| = \left| \frac{\Gamma_n}{s_n^3} \right|.
 \]
@@
 \[
-|f(n)| \leq 1 \cdot |g(n)| \quad \text{for all } n \geq 1
+|f(n)| \leq 1 \cdot |g(n)| \quad \text{for all } n \geq 1.
 \]
@@
 \begin{equation} \label{eq:4}
-\frac{\Gamma_n}{s_n^3} = O\left(\frac{\Gamma_n}{s_n^3}\right)
+\frac{\Gamma_n}{s_n^3} = O\left(\frac{\Gamma_n}{s_n^3}\right).
 \end{equation}
@@
    \[
-   K \cdot \frac{\Gamma_n}{s_n^3} = O\left(\frac{\Gamma_n}{s_n^3}\right)
+   K \cdot \frac{\Gamma_n}{s_n^3} = O\left(\frac{\Gamma_n}{s_n^3}\right).
    \]
@@
 \[
-\frac{M}{6} \sum_{j=1}^n \left( \frac{\gamma_j}{s_n^3} + \frac{c \cdot \sigma_j^3}{s_n^3} \right) \leq O\left(\frac{\Gamma_n}{s_n^3}\right)
+\frac{M}{6} \sum_{j=1}^n \left( \frac{\gamma_j}{s_n^3} + \frac{c \cdot \sigma_j^3}{s_n^3} \right) \leq O\left(\frac{\Gamma_n}{s_n^3}\right).
 \]
@@
 \[
-\forall f \in C^3: \left| \space  \mathbb{E}\left[f\left(\frac{S_n}{s_n}\right)\right] - \mathbb{E}\left[f\left(N\right)\right] \right| \leq O\left(\frac{\Gamma_n}{s_n^3}\right) 
+\forall f \in C^3: \left| \space  \mathbb{E}\left[f\left(\frac{S_n}{s_n}\right)\right] - \mathbb{E}\left[f\left(N\right)\right] \right| \leq O\left(\frac{\Gamma_n}{s_n^3}\right).
 \]
@@
 \[
-\frac{\Gamma_n}{s_n^3} \to 0 \quad \text{as } n \to \infty
+\frac{\Gamma_n}{s_n^3} \to 0 \quad \text{as } n \to \infty.
 \]
@@
 
-Since \( \mathbb{E}[f(S_n / s_n)] \to \mathbb{E}[f(N)] \) for all bounded and continuous test functions \( f \), by the general criterion for vague convergence, we conclude that:
+Since \( \mathbb{E}[f(S_n / s_n)] \to \mathbb{E}[f(N)] \) for all bounded and continuous test functions \( f \), by the general criterion for vague convergence (in Theorem 6.1.6 in \cite{chung2000course}), we conclude that:
 \[
-\frac{S_n}{s_n} \xrightarrow{d} \mathcal{N}(0, 1)
+\frac{S_n}{s_n} \xrightarrow{d} \mathcal{N}(0, 1).
 \]
@@
     \end{enumerate}
+\end{itemize}
+
+\subsubsection{Formal Proof in HOL4}
+Key lemmas and theorems:
+\begin{itemize}
+    \item Weak convergence
+    \item Taylor expansion of expectations
+    \item Big-O bounds for error terms
+\end{itemize}
+
+\subsection{Limitations and Future Directions}
+\subsubsection{Limitations}
+\begin{itemize}
+    \item Mention limitations in HOL4’s current libraries: Incomplete ComplexTheory, which prevents characteristic function-based proofs, Lack of direct support for third moments of normal random variables, making certain bounds harder to verify formally.
+    \item This work is for proof of Lyapunov’s version, which is more general than the i.i.d. case but not the most general form (e.g., Lindeberg-Feller).
+\end{itemize}
+
+\subsubsection{Future Directions}
+\begin{itemize}
+    \item Explore the possibility of extending HOL4 to include: Characteristic functions and ComplexTheory, and A general Lindeberg-Feller version of the CLT.    
 \end{itemize}
@@
 
-\subsubsection{Future Directions}
-...
-
 In closing, this thesis provides not only a sound formalization of Lyapunov's CLT but also the whole range of insights into what happens when formal methods are integrated with probability theory.

### 2024-12-06 background.tex
Keep working on Historical Background

@@
-\chapter{Background and Related Work}
-\label{background}
+\chapter[Background and Related Work]{Background and Related Work}
+\label{chap:2}
 
@@
 
-\section{Background}
+\section{Historical Background}
 Let us brief the few evolvement of CLT, mainly noticing its critical features and variants developed further.
@@
 These extensions illustrate how the theorem can be adapted to a wide range of probabilistic and statistical contexts.
+The Central Limit Theorem (CLT) has undergone significant evolution since its inception, transitioning from specific approximations to a cornerstone of modern probability theory.
+Among the key milestones in this journey are the foundational papers (1900/01) by Lyapunov, which marked a pivotal step toward abstraction and a more formalized understanding of the theorem.
+The works of Lyapunov not only reconstructed the earlier approaches but also brought in a degree of mathematical rigor consistent with the analytical standards established by Weierstrass.
+
+Lyapunov's work differed from predecessors like Poisson and Chebyshev in that he gave an explicit, uniform bound for the difference between the distribution of the sum of random variables and the limiting normal distribution.
+This degree of precision met the requirement of a more precise probabilistic result, as called for by Chebyshev, and at the same time made the proof simpler.
+Lyapunov managed to do this by an especially elementary and at the same time subtle rewriting of Poisson's argument, including standards of rigor that had become standard in mathematics in the late 19th century.
+
+Lyapunov's goals for the work also distinguish it.
+He remarks in his 1900 publication that his two main aims were to give a direct proof of the CLT - one that did not depend so heavily on the specific theories about moments developed by Chebyshev and Markov - and to relax the conditions under which the theorem was known to hold.
+This "direct" and "elementary" method was supposed to make the internal relationships within the theorem itself clearer and the theorem more accessible and self-contained.
+Indeed, Lyapunov's work represents these attributes: internally mathematically clear, abstract, yet attached to practical applicability-a sign of probability theorists of the time.
+His contemporaries, including von Mises, Lévy, Cramér, Khinchin, and Kolmogorov, similarly pursued abstract relationships in probability theory while eschewing purely formalistic approaches.
+
+Yet Lyapunov himself was far from neglecting practical applicability as a criterion of mathematical achievement.
+The balance between working within mathematics and responding to external criteria of utility is typical of the broader development of probability theory in the modern period.
+This tension between purely abstract formalization and practical relevance is a defining feature of mathematical work to this day.
+
+The generalization of CLT in Lyapunov's version is a balancing act, an extension of the classical case of independence and identical distribution.
+It accommodates independent random variables with only finite moments, giving way for a more flexible framework yet rigid enough to allow for a mathematical generalization.
+The present version forms the basis necessary for formalization in this thesis, whereby emphasis has been laid on directly obtaining real-valued methods to establish convergence under relaxed conditions.
+
+\textbf{Why Lyapunov's Version of the CLT is Chosen}
+
+Lyapunov’s version of the Central Limit Theorem is particularly well-suited for formalization due to its elegant balance between generality and simplicity.
+Unlike earlier versions of the CLT that required independent and identically distributed (i.i.d.) random variables, Lyapunov’s theorem weakens these conditions by accommodating independent random variables with potentially different variances and finite higher moments.
+This generalization broadens its applicability while maintaining the rigor needed for precise mathematical analysis.
+
+One of the main reasons for selecting Lyapunov's version is that it is based on finite moments, whereas most other proofs of the CLT are based on the characteristic function approach.
+Though the method of characteristic functions is powerful and general, it necessarily works within the realm of complex-valued functions and makes use of Fourier transforms, which adds extra layers of abstraction and dependencies.
+In contrast, Lyapunov's proof stays within the realm of real-valued analysis, using explicit bounds and direct error estimates.
+This makes it both more accessible and more amenable to formalization in systems like HOL4, which currently has limited support for complex-valued analysis.
+
+Moreover, Lyapunov's theorem explicitly provides a uniform bound on the error between the distribution of the normalized sum of random variables and the limiting normal distribution.
+The explicit control over error terms, obtained by using finite moments, is in harmony with the constructive nature of formal proofs.
+By employing real-valued tools such as Taylor expansions, bounding techniques, and Big-O notation, Lyapunov's proof sidesteps the intricacies of characteristic functions while retaining mathematical rigor.
+
+In the context of this thesis, Lyapunov's is the ideal candidate for formalisation: Its direct approach relies on ﬁnite moments, focusing on real-valued methods, which perfectly suits the strengths and current capabilities of the HOL4 theorem prover.
+Moreover, compared to the classical i.i.d.
+case, Lyapunov's theorem is general  and provides a stronger basis for subsequent extensions and applications.
 

### 2024-12-10 draft.tex
Added Draft Outline of Thesis and Introduction

@@
 \end{itemize}
+\section{Introduction}
+\textbf{Purpose:} Set the stage for this thesis, establish its relevance, and provide an overview of the challenges and contributions.
 
-\subsection{Introduction}
-\cite{tian2022assumption}
-\cite{qasim2016formalization}
+\subsection{Background of the Central Limit Theorem (CLT)}
 \begin{itemize}
-    \item What is CLT
-    \item Define the CLT in its simplest form (e.g., i.i.d. random variables)
-    \item Highlight its importance in statistics, machine learning, and other fields as a foundation for normal approximation.
+    \item Briefly describe the CLT's significance in probability theory and its foundational role in various fields.
+    \item Highlight the diversity of proofs for the CLT, focusing on Lyapunov’s direction as a generalization of the i.i.d. case.
 \end{itemize}
 
-\subsubsection{Motivation}
-Why formalize the CLT in HOL4? 
-The challenges with probabilistic reasoning, the need for rigor, and the importance of verifying a key theorem like the CLT.
+\subsection{Formal Verification of Mathematical Theorems}   
 \begin{itemize}
-    \item Enriching the HOL4 Theory Library
-    \item Supporting Future Applications
-    \item Providing Analytical Tools
+    \item Discuss the importance of formalizing theorems like the CLT for computational mathematics.
+    \item Introduce HOL4 and its potential for advancing probability theory through rigorous formalization.
 \end{itemize}
 
-\subsubsection{Contributions}
+\subsection{Challenges and Motivation}
 \begin{itemize}
-    \item This thesis proves the Lyapunov version of the Central Limit Theorem, a fundamental result in probability theory.
-The CLT of Lyapunov extends the classical CLT by relaxing the identical distribution assumption to the case of independent but not necessarily identically distributed random variables.
-    \item The proof makes use of Taylor expansions, moment-based bounds, and asymptotic error analysis to rigorously establish convergence to the normal distribution.
-    \item The main insights involve the Lyapunov condition, which guarantees that the third absolute moments become small compared to the variance, and a new use of Big-O notation for the effective control of error terms.
+    \item Outline the unique challenges of formalizing Lyapunov’s approach in HOL4:
+    \item HOL4 lacks a robust library for complex analysis and probability.
+    \item The proof depends on Taylor expansions, Big-O notation, and Lyapunov’s inequality, which needed to be formalized from scratch.
+    \item Describe the failed attempt to use moment generating functions (MGFs) and how it inspired the shift to Lyapunov’s approach.
 \end{itemize}
 
-\subsubsection{Results}
-This form of CLT is more general than the i.i.d case but not the most general (not the Lindeberg-Feller version).
-
+\subsection{Thesis Contributions}
 \begin{itemize}
-    \item \textbf{Lyapunov CLT}: For independent random variables \(X_1, X_2, \ldots, X_n \) with finite means, variances, and third moments, if the Lyapunov condition is satisfied:
-$$
-\frac{\Gamma_n}{s_n^3} \to 0 \quad \text{as } n \to \infty,
-$$
-then the normalized sum \( \frac{S_n - \mu}{s_n} \) converges in distribution to the standard normal distribution \( N(0, 1) \).
-    \item \textbf{Proof}: It establishes that
-    \begin{itemize}
-        \item Error bounds of the test function \(f\) by Taylor expansion and moment inequalities.
-        \item Asymptotic vanishing of higher-order terms under the Lyapunov condition.
-\item A rigorous comparison between the distribution of \( \frac{S_n}{s_n} \) and \( N(0, 1)\), bounded by \( O\left(\frac{\Gamma_n}{s_n^3}\right) \).
+    \item Summarize key contributions:
+    \begin{enumerate}
+        \item Formalization of Lyapunov’s inequality in HOL4.
+        \item Development of supporting libraries for Taylor expansions, Big-O notation, and moment analysis.
+        \item Complete formalization of the CLT based on Lyapunov’s theorem.
+    \end{enumerate}
 \end{itemize}
 
-These results illustrate the robustness of the Lyapunov CLT against heterogeneous distributions and, in fact, form a bridge between classical formulations and modern applications.
+\section{Background and Related Work}
+\textbf{Purpose:} Position this work in the context of previous efforts and provide foundational concepts.
+\subsection{Overview of the Central Limit Theorem}
+\begin{itemize}
+    \item Explain the statement of the CLT and its generalizations (e.g., Lyapunov’s and Lindeberg’s approaches).
+    \item Focus on Lyapunov’s version as the basis of this proof:
 
-Also, highlight key formal results of other lemmas and theorems to support (e.g., bounding errors using Big-O, weak convergence).
+     \begin{equation}
+     \text{If } \frac{\Gamma_n}{s_n^3} \to 0 \text{ as } \space \space n \to \infty, 
+     \end{equation}
+     then 
+     \[
+     \frac{S_n}{s_n} \xrightarrow{d} \Phi. 
+     \] 
+     Define key terms: \( \Gamma_n, s_n, S_n \), and Lyapunov’s condition.
 \end{itemize}
 
-\subsection{Background and Related Works}
-\subsubsection{Historical Background}
-Introduce the histrory of CLT and mention different versions of CLT \cite{fischer2011history}
+\subsection{Formal Proof Assistants in Probability Theory}
 \begin{itemize}
-    \item Classical versions: De Moivre-Laplace (i.i.d.), Lindeberg-Feller, Lyapunov’s theorem.
-    \item Explain why Lyapunov’s version is chosen (finite moments, no characteristic functions).
+    \item Discuss Isabelle/HOL and its use of characteristic functions to prove the CLT (mention Luke Serafin’s work briefly here).
+    \item Highlight HOL4’s potential and this thesis's motivation to enrich its probability theory libraries.
 \end{itemize}
 
+\subsection{Lyapunov’s Direction and Proof Techniques}   
 \begin{itemize}
-    \item \textbf{Classical Central Limit Theorem} \cite{ross2019first}
+    \item Provide an informal overview of Lyapunov’s proof steps:
     \begin{itemize}
-        \item \textbf{Proven by:} De Moivre, Laplace
-        \item \textbf{Statement:} If \( X_1, X_2, \ldots, X_n \) are independent and identically distributed (i.i.d.) random variables with finite mean \( \mu \) and variance \( \sigma^2 \), then:
-  \[
-  Z_n = \frac{\sum_{i=1}^n X_i - n\mu}{\sigma\sqrt{n} } \xrightarrow{d} N(0, 1).
-  \]
-  \item \textbf{Significance:} Foundation of probability theory and statistics, used in inferential statistics, e.g., hypothesis testing and confidence intervals.
+        \item Construction of normal approximations (\( Y_j \)).
+        \item Use of Taylor expansions to bound differences.
+        \item Application of Lyapunov’s inequality to ensure convergence.
     \end{itemize}
 
-    \item \textbf{Lyapunov’s Central Limit Theorem} \cite{chung2000course}
-    \item \textbf{Lindeberg’s Central Limit Theorem}
-    \item \textbf{The Generalized Central Limit Theorem} 
-    \begin{itemize}
-        \item https://en.wikipedia.org/wiki/Central_limit_theorem
+   \item Contrast Lyapunov’s approach with other CLT proofs, emphasizing why it’s more suitable for HOL4’s existing framework.
 \end{itemize}
 
+\section{The Formalization Process}
+\textbf{Purpose: Detail the technical challenges and solutions in formalizing Lyapunov’s proof in HOL4.}
+\subsection{Overview of Lyapunov’s Proof Structure}
+\begin{itemize}
+    \item Provide a roadmap of the proof components you formalized:
+    \begin{enumerate}
+        \item Taylor expansions and approximation of test functions.
+        \item Telescoping sums and bounding higher-order moments.
+        \item Lyapunov’s inequality and its application to ensure convergence.
+    \end{enumerate}
 \end{itemize}
 
-\subsubsection{Related Work}
-The CLT is a fundamental result in probability theory, and its formalization in theorem provers has been a highly active area of interest in the formal methods community.
-Several different ways of proving the CLT have been tried, and tools like Isabelle/HOL and HOL4 provide different approaches reflecting their different frameworks and mathematical libraries.
+\subsection{Formalizing the Supporting Theorems in HOL4}
+\begin{itemize}
+    \item Taylor Expansion
+    \item Big-O Notation
+    \item Lyapunov’s Inequality
+\end{itemize}
 
-\textbf{Isabelle/HOL's Formalization of the CLT}
-In Isabelle/HOL, the CLT has been formalized in a powerful and especially elegant way using the \textbf{characteristic function approach}, which relies on the fact that characteristic functions uniquely determine probability distributions and their pointwise convergence implies convergence in distribution.
-\cite{serafin2015formally}
+\subsection{Technical Challenges and Innovations}
+\begin{itemize}
+    \item Share how HOL4’s lack of complex analysis tools required alternative approaches (e.g., using Lyapunov’s proof instead of characteristic functions).
+    \item ...
+\end{itemize}
 
+\section{The Formal Proof of the CLT in HOL4}
+\textbf{Purpose:} Present the completed proof and validate its correctness.
+\subsection{Step-by-Step Proof}
 \begin{itemize}
-    \item \textbf{Methodology}
-    \item \textbf{Characteristic Function Approach}
+    \item Present the formalized proof of Lyapunov’s theorem in HOL4.
+    \item Break it into logical steps, mirroring the informal proof:
+    \begin{enumerate}
+        \item Setup and normalization of \( S_n \).
+        \item Construction of \( Z_j \) and comparison with \( X_j \).
+        \item Application of Taylor expansions and bounding error terms.
+        \item Application of Lyapunov’s inequality to complete the proof.
+    \end{enumerate}
+\end{itemize}
+
+\subsection{Validation and Soundness}  
 \begin{itemize}
-        \item It provides a direct and compact route to proving the CLT, leveraging the algebraic simplicity of characteristic functions.
-        \item ...
+    \item Discuss how HOL4 ensures the proof’s correctness and soundness.
+    \item Highlight key intermediate results that confirm the integrity of this thesis's approach.
+\end{itemize}
+
+\section{Results and Comparisons}
+\textbf{Purpose:} Showcase the impact of this work and compare it to related efforts.
+\subsection{Achievements in HOL4}
+\begin{itemize}
+    \item Summarize the completed formalization of Lyapunov’s CLT proof.
+    \item  Highlight the additional libraries (e.g., Taylor expansions, Big-O notation) and their potential for future work.
+\end{itemize}
+
+\subsection{Comparison with Isabelle/HOL}
+\begin{itemize}
+    \item Briefly compare this work with Luke Serafin’s formalization of the CLT in Isabelle:
+    \begin{itemize}
+        \item HOL4 uses Lyapunov’s approach instead of characteristic functions.
+        \item Discuss the trade-offs: HOL4’s modularity and flexibility versus Isabelle’s existing tools for complex analysis.
     \end{itemize}
@@
 
-\subsubsection{Overview of Informal Proof}
+\section{Conclusion}
+\textbf{Purpose:} Reflect on the thesis’s contributions and look ahead.
+\subsection{Summary of Contributions}
+Restate achievements:
+\begin{enumerate}
+     \item Formalized Lyapunov’s theorem and the CLT in HOL4.
+     \item  Extended HOL4’s libraries for probability theory.
+     \item  Overcame challenges related to HOL4’s lack of complex analysis tools.
+\end{enumerate}
+
+Also, discuss the significance of enriching HOL4 for future formalizations in probability and functional analysis.
+
+\subsection{Future Work}
+Suggest potential extensions, such as:
 \begin{itemize}
-    \item Highlight the direct and real-valued approach of Lyapunov's proof.
-    \item Discuss how it avoids characteristic functions and instead uses bounding error terms (Taylor expansion, Big-O).
+    \item Formalizing multivariate versions of the CLT.
+    \item Extending HOL4’s libraries for advanced probabilistic theorems.
 \end{itemize}
 
-\subsection{Preliminaries}
-\subsubsection{HOL4}
-\cite{tian2022assumption}
-\subsubsection{Measure Theory}
-\cite{qasim2016formalization}
-\subsubsection{Lebesgue Integration Theory}
-\cite{qasim2016formalization}
-\subsubsection{Probability Theory}
-\cite{qasim2016formalization}
+\subsection{Appendices and Supplementary Material}
+Code listings, etc.
 
-\newpage
 \subsection{The proof of Central Limit Theorem}
@@
 \end{itemize}
-
-\subsubsection{Formal Proof in HOL4}
-Key lemmas and theorems:
-\begin{itemize}
-    \item Weak convergence
-    \item Taylor expansion of expectations
-    \item Big-O bounds for error terms
-\end{itemize}
-
-\subsection{Limitations and Future Directions}
-\subsubsection{Limitations}
-\begin{itemize}
-    \item Mention limitations in HOL4’s current libraries: Incomplete ComplexTheory, which prevents characteristic function-based proofs, Lack of direct support for third moments of normal random variables, making certain bounds harder to verify formally.
-    \item This work is for proof of Lyapunov’s version, which is more general than the i.i.d. case but not the most general form (e.g., Lindeberg-Feller).
-\end{itemize}
-
-\subsubsection{Future Directions}
-\begin{itemize}
-    \item Explore the possibility of extending HOL4 to include: Characteristic functions and ComplexTheory, and A general Lindeberg-Feller version of the CLT.    
-\end{itemize}
-
-\subsection{Conclusions}
-The CLT is a fundamental result in probability theory, ranging from mathematics and statistics to finance and data science.
-This work described the formal proof of the CLT using Lyapunov's criterion in the HOL4 theorem prover.
-The proof showed that the normalized sum of independent (but not necessarily identically distributed) random variables converges in distribution to the standard normal distribution by exploiting a moment-based approach along with Lyapunov's condition.
-
-\textbf{Conclusions}
-\begin{itemize}
-    \item \textbf{Formalization of Lyapunov's CLT}
-    \begin{itemize}
-        \item The proof formalized Lyapunov's version of the CLT by providing rigorous bounds on error terms and establishing convergence under the Lyapunov condition.
-The use of Big-O notation and asymptotic error analysis was instrumental in managing higher-order terms.
-    \end{itemize}
-
-    \item \textbf{Theoretical Insights}
-    \begin{itemize}
-        \item In addition, a constructive, moment-based methodology used in its proof gave a natural perspective on finite variances and higher-order moments, each playing an important part in ensuring convergence according to CLT.
-    \end{itemize}
-
-    \item \textbf{Comparison with Previous Works}
-    \begin{itemize}
-        \item The approach represents a complementary alternative to existing formalizations, such as \textbf{Isabelle/HOL's characteristic function-based proof}, in that it represents a constructive proof based on explicit bounds and moment inequalities.
-It is also an addition to the variety of formalizations and expands the reach of theorem provers regarding probabilistic results.
-    \end{itemize}
-
-\end{itemize}
-Formalization of CLT in HOL4 showcases the power of theorem provers in the formal verification of fundamental results in probability theory.
-It fills the gap between theoretical results and their verification by computation, showing the potential of formal methods in academic and applied domains.
-
-In closing, this thesis provides not only a sound formalization of Lyapunov's CLT but also the whole range of insights into what happens when formal methods are integrated with probability theory.
-It lays the ground for further investigations into probabilistic results within theorem provers, opening perspectives to theoretical advances as well as practical applications.

### 2024-12-10 introduction.tex
Added Draft Outline of Thesis and Introduction

--- 2024-11-30 background.tex
@@
-\chapter{Background and Related Work}
-\label{background}
+\chapter[Introduction]{Introduction}
+\label{chap:1}
 
-Background 123
+% \section{Introduction}
+% \textbf{Purpose:} Set the stage for this thesis, establish its relevance, and provide an overview of the challenges and contributions.
 
-% \begin{figure*}
-% \centering
-% \includegraphics[width=0.98\textwidth]{Chapter1/Images/Fig1.pdf}
-% \caption{caption}
-% \label{figCh1:figure}
-% \end{figure*}
+% \subsection{Background of the Central Limit Theorem (CLT)}
+% \begin{itemize}
+%     \item Briefly describe the CLT's significance in probability theory and its foundational role in various fields.
+%     \item Highlight the diversity of proofs for the CLT, focusing on Lyapunov’s direction as a generalization of the i.i.d. case.
+% \end{itemize}
 
-\renewcommand\thefigure{\thechapter .\arabic{figure}}
+% \subsection{Formal Verification of Mathematical Theorems}   
+% \begin{itemize}
+%     \item Discuss the importance of formalizing theorems like the CLT for computational mathematics.
+%     \item Introduce HOL4 and its potential for advancing probability theory through rigorous formalization.
+% \end{itemize}
+
+% \subsection{Challenges and Motivation}
+% \begin{itemize}
+%     \item Outline the unique challenges of formalizing Lyapunov’s approach in HOL4:
+%     \item HOL4 lacks a robust library for complex analysis and probability.
+%     \item The proof depends on Taylor expansions, Big-O notation, and Lyapunov’s inequality, which needed to be formalized from scratch.
+%     \item Describe the failed attempt to use moment generating functions (MGFs) and how it inspired the shift to Lyapunov’s approach.
+% \end{itemize}
+
+% \subsection{Thesis Contributions}
+% \begin{itemize}
+%     \item Summarize key contributions:
+%     \begin{enumerate}
+%         \item Formalization of Lyapunov’s inequality in HOL4.
+%         \item Development of supporting libraries for Taylor expansions, Big-O notation, and moment analysis.
+%         \item Complete formalization of the CLT based on Lyapunov’s theorem.
+%     \end{enumerate}
+% \end{itemize}
+
+Formal verification has continued revolutionizing the understanding and application of foundational theorems in mathematics: the surety of accompanying rigor and its implementation in computational systems.
+Thus have built chances for innovative mathematical services.
+Probability theory is a domain with both important theoretical results and practical applications, and it has increasingly become a focus of formalization efforts.
+The most important from such results is the Central Limit Theorem (CLT), which carries very strong implications in theory and practice.
+
+The CLT is perhaps the most prominent result in probability theory describing how the sum of independent random variables, if suitably normalized, approximates a standard normal distribution when the number of terms increases indefinitely.
+It closes the gap between random single events and predictable behavior in the aggregate, forms the basis of inferential statistics, and underpins models in a vast variety of fields such as physics, finance, and artificial intelligence.
+Given this importance, formal verification of the CLT using proof assistants will be an important step toward proving its correctness in automated reasoning systems.
+
+However, formalizing CLT is far from straightforward.
+There exist numerous variants of this theorem, together with their numerous proofs, ranging from classical independent and identically distributed (i.i.d.) to some more general versions, such as Lyapunov's theorem \cite{billingsley2017probability}.
+Each of those proofs uses advanced mathematical tools, including characteristic functions, moment generating functions, or even higher-order moments, according to the situation.
+These proofs require advanced mathematical infrastructure, which is not available with all proof assistants.
+In particular, the case of HOL4 \cite{slind2008brief} has found it impossible to use some of the approaches, particularly the one based on characteristic functions, because the strong complex analysis library was lacking.
+
+Instead, this thesis follows another path: formalizing in the HOL4 proof assistant a generalized version of the CLT by Lyapunov's theorem.
+Lyapunov's approach is particularly appealing, as it has no direct use of either complex analysis or characteristic functions.
+Instead, it pays attention to moment-based convergence criteria, trying to provide its conditions for convergence using Taylor expansions, Big-O notation of error bounds, and Lyapunov's inequality.
+This approach aligns well with the existing mathematical infrastructure of HOL4 while simultaneously expanding its capabilities.
+
+This project formalizes Lyapunov's proof in HOL4 and demonstrates the feasibility of verifying advanced probabilistic results in HOL4. More specifically, it has added:
+\begin{itemize}
+    \item Formally defined Taylor expansions in function bounds and error terms.
+    \item Big-O, allowing accurate representations of asymptotic behavior.
+    \item Lyapunov's inequality, the key condition for the convergence between the CLT.
+\end{itemize}
+
+\section{Motivation}
+The formalization of the Central Limit Theorem in HOL4 involved interesting challenges because of the highly mathematical nature of the theorem and the specific limitations of existing libraries in HOL4.
+Another difference between the proof assistants, Isabelle/HOL, which has a very mature library for complex analysis and HOL4, is that there are no built-in key concepts for working with characteristic functions and others associated with the common classical proofs of CLT.
+
+Lack of a well-developed library has been one of the major setbacks of HOL4.
+Proving the central limit theorem using the more traditional kinds of proofs, such as the one formalized in Isabelle/HOL, involves the use of characteristic functions for convergence proof.
+Characteristic functions are essentially Fourier transforms and thereby hinge a lot on complex analysis.
+It would have been very difficult to follow this approach in HOL4, since the proof assistant currently lacks support for these advanced tools.
+
+An alternative approach was explored using moment generating functions (MGFs).
+MGFs are much more popular for proving the CLT, wherein their specialized property is utilized to show convergence to the standard normal distribution.
+However, it soon became evident that formally establishing the uniqueness property of MGFs would become a serious blocker.
+While this property is a triviality in classical analysis, the formalization work involved was highly intricate and beyond the scope of what is currently practical within HOL4's libraries.
+As a result, this approach was abandoned, leading to the need for a new direction.
+
+The breakthrough came with the adoption of Lyapunov's theorem as the basis for formalizing the CLT.
+Lyapunov's argument has a very direct and elementary approach to the theorem without using complex analysis or moment-generating functions but rather basing on real-valued, moment-based convergence criteria; hence, it being particularly fitted into the existing framework of HOL4.
+Lyapunov's proof is a milestone of abstraction in the probability theory, but it is still beautifully simple and thus met the practical and strict requirements of formal verification.
+
+However, Lyapunov’s method introduced its own set of challenges for formalization in HOL4:
+\begin{itemize}
+    \item \textbf{Taylor Expansions:} The proof relies heavy on Taylor expansions to approximate test functions and bound errors. These had to be rigorously formalized in HOL4 from first principles.
+    \item \textbf{Big-O Notation:} Reasoning about the asymptotic behavior of error terms involved the formalization of Big-O notation, which did not exist in HOL4 until now.
+    \item \textbf{Lyapunov's Inequality:} An important part of the proof was the formalization of Lyapunov's inequality, which gives sufficient conditions for convergence.
+It was carefully adapted to fit into HOL4's logical framework.
+\end{itemize}
+
+Following the Lyapunov's approach, with its valuation on real techniques and elementary methods, proved most important: it served to encapsulate the CLT under the constraints of HOL4 and was entirely consistent toward building reusable mathematics libraries.
+Taylor series developments, Big-O notion, and Lyapunov's inequality are indispensable not only for this proof but also as a basis of further formalizations in the realm of probability theory and elsewhere.
+
+This shift from characteristic functions and MGFs to the more concrete, moment-based approach of Lyapunov underlines flexibility and problem-solving ingenuity necessary in formal verification projects.
+It thus shows how advanced mathematical results can be formalized in HOL4, even when serious deficiencies in its infrastructure must first be overcome.
+Moreover, it emphasizes the role of Lyapunov's approach as a turning point in the history of probability theory and as one of the bases for modern formalization.
+
+\section{Structure of the Thesis}
+The rest of the thesis is organized as follows.
+Chapter 2 provides background on the Central Limit Theorem, its history, and the setting of Lyapunov's proof.
+Related work, including other formalizations, using proof assistants like Isabelle/HOL is also covered.
+Chapter 3 then details the formalization, discussing the challenges faced along with solutions developed, in particular the formalization of Taylor expansions, Big-O notation and Lyapunov's inequality.
+The key components of the proof, its completion, and its validation of the CLT are discussed in Chapter 4.
+Chapter 5 presents a discussion of results, and positioning of this work with respect to related efforts on formal verification.
+Finally, the thesis is concluded in Chapter 6 by summarizing the contributions made by this work and presenting the vision on future directions that probability theory in HOL4 may take.
+
+This thesis not only shows that a formalization of the CLT in HOL4 is possible but also opens up further opportunities for work on the formal probability theory.
+It fills in some foundational gaps and provides reusable tools, which illustrates more general potential of HOL4 to make serious contributions to the fully verified mathematics corpus.

### 2024-12-11 background.tex
Added Background and Related Work, citation problems

--- 2024-11-30 background.tex
@@
-\chapter{Background and Related Work}
-\label{background}
+\chapter[Background and Related Work]{Background and Related Work}
+\label{chap:2}
 
-Background 123
+% \section{Background and Related Work}
+% \textbf{Purpose:} Position this work in the context of previous efforts and provide foundational concepts.
+% \subsection{Overview of the Central Limit Theorem}
+% \begin{itemize}
+%     \item Explain the statement of the CLT and its generalizations (e.g., Lyapunov’s and Lindeberg’s approaches).
+%     \item Focus on Lyapunov’s version as the basis of this proof:
 
-% \begin{figure*}
-% \centering
-% \includegraphics[width=0.98\textwidth]{Chapter1/Images/Fig1.pdf}
-% \caption{caption}
-% \label{figCh1:figure}
-% \end{figure*}
+%      \begin{equation}
+%      \text{If } \frac{\Gamma_n}{s_n^3} \to 0 \text{ as } \space \space n \to \infty, 
+%      \end{equation}
+%      then 
+%      \[
+%      \frac{S_n}{s_n} \xrightarrow{d} \Phi. 
+%      \] 
+%      Define key terms: \( \Gamma_n, s_n, S_n \), and Lyapunov’s condition.
+% \end{itemize}
 
-\renewcommand\thefigure{\thechapter .\arabic{figure}}
+The history of the Central Limit Theorem, hereinafter referred to as CLT, is quite a fascinating journey through the gradual development of probability theory.
+Over the years, various versions of CLT have been formulated, reflecting its adaptability and foundational importance.
+Each such variation addresses certain mathematical challenges and applications, therefore enriching the broader understanding of convergence in probability.
+In this chapter, we explore some theorem-proving tools and their methodologies to formalize the CLT, emphasizing ...
+
+\section{Historical Background}
+Let us just trace briefly the historical evolution of the CLT, critical milestones, key contributors, and the gradual refinement of its conditions and scope.
+The roots of the CLT go back to the dispute of the eighteenth century, specifically to the time Abraham de Moivre laid the foundation for this theorem.
+In 1733, de Moivre showed that the sum of a large number of independent and identically distributed random variables converges to the normal distribution-an elegant approximation to problems such as the rolling of dice or repeated coin flips \cite{de1733approximatio}.
+His work initiated the interplay between discrete distributions-such as the Binomial-and their continuous approximations, and laid the ground for developments to come.
+
+In 1810, Pierre-Simon Laplace independently extended de Moivre's insights by proving a much more general form of this theorem.
+Under broader conditions, Laplace demonstrated convergence of the sum of independent random variables to a normal distribution, thereby establishing CLT as a universal principle.
+His work was used in applying the CLT to population statistics and measurement errors \cite{laplace1835oeuvres} and cemented its usefulness for aggregate phenomena analysis.
+
+The 19th century would see the formalization of the central limit theorem (CLT) by Pafnuty Chebyshev, who also gave the theorem a rigorous foundation by the introduction of variance and expectation conditions.
+With that work, Chebyshev rendered the CLT mathematically precise, linking it to the developing area of probability theory \cite{chebyshev1890deux}.
+Building on this work was further advancing the generalization of the CLT by Alexander Lyapunov to sequences of independent random variables that are not identically distributed.
+His version of the theorem brought forth Lyapunov's condition, which contained definite criteria for convergence \cite{lyapunov1895pafnutii}.
+
+In the early years of the 20th century, contributions by Lyapunov represented such a change in the evolution of the CLT.
+Unlike his predecessors, Lyapunov sought a “direct” and “elementary” proof.
+It would try to make the theorem internal relations simpler and relax the theorem's assumptions.
+His papers published in 1900 and 1901 reconstructed previous methods with a rigor of Weierstrass, and introduced new precision to the theorem by establishing explicit bounds on the normalized sum and the limiting normal distribution.
+This specificity answers Chebyshev's earlier demand for sharper probabilistic results although it should have simplified the proof through some clever but much deeper refinements of Poisson's arguments \cite{fischer2011history}.
+
+Lyapunov’s goals extended beyond mere generalization; he aimed to illuminate the theorem’s internal structure and make it more accessible.
+Avoiding the complexities of characteristic functions and complex analysis, he focused on moment-based criteria, relying on real-valued tools like Taylor expansions, bounding techniques, and Lyapunov's inequality.
+Thus the balance here is theoretical rigor and practical applicability, two characteristics that still resonate well with modern efforts at formal verification as that undertaken in this thesis.
+
+In contemporary times, further refinements of the CLT were provided by William Feller bringing it to discrete random variables \cite{feller1945}, and all subsequent developments that are probabilistic have been geared towards this common extension to stable distributions and higher dimensional data.
+These extensions demonstrate the theorem’s adaptability to various contexts, from stochastic processes to machine learning and data analysis.
+
+Lyapunov’s version of the CLT represents a key milestone in this journey.
+It generalizes the theorem for independent random variables with finite moments so that it becomes an observed but pliant frame of reference for convergence.
+It is this balancing between abstractness and concreteness that is a characteristic of the whole development of probability during the modern period.
+Contemporaries of Lyapunov such as von Mises, Lévy, Cramér, Khinchin, and Kolmogorov likewise pursued abstract relations in probability theory while keeping an eye on practical relevance.
+This tension is paradigmatically present in Lyapunov's work, which illustrates the way in which results which are purely formal mathematics may nonetheless satisfy external criteria of applicability and utility.
+For more information about the history and the proofs of the Central Limit Theorem, see \cite{fischer2011history, adams2009life}.
+
+This thesis is based on the generalization of CLT as expressed in Lyapunov's version.
+Because of its focus on real-valued methods and moment-based criteria, it avoids any complex tools like characteristic functions or complex analysis, hence being particularly suitable for formalization in HOL4.
+In this respect, Lyapunov's approach not only advances the mathematical generalization but also provides a practical framework to meet the rigor and accessibility requirements of automated reasoning systems by directly addressing convergence under relaxed conditions.
+
+\subsection{Related Work}
+
+Formalization of mathematical theorems as proofs has been a singular research focus in modern computational mathematics.
+The systems such as Isabelle/HOL \cite{isabelle}, Coq \cite{bertot2013interactive}, and HOL4 \cite{slind2008brief} have been heavily relied on over the years to encode and prove basic properties but offer several features and challenges.
+
+One of the most significant achievements in this area was the formalization of CLT within the Isabelle/HOL proof assistant.
+This was made possible mainly by the rich library related to complex analysis in Isabelle/HOL, which provided the means to make use of characteristic functions within the proof.
+Characteristic functions are Fourier transforms of probability distributions that make convergence easier to analyze by drawing on their special mathematical properties.
+Central to the approach are two key steps: first, pointwise convergence of the characteristic function for the normalized sum of independent random variables to that of the standard normal distribution should be demonstrated; and second, by the application of the Lévy Continuity Theorem, one obtains the desired convergence in distribution.
+These two steps are actually the core of the proof that eventually leads to the verification of CLT \cite{billingsley2017probability}.
+
+The characteristic function approach is beautifully complemented by the infrastructure provided by Isabelle/HOL, which supports careful manipulation of complex-valued functions, differentiation, and limits.
+This thereby enables Isabelle/HOL to achieve the rigorous performance that is demanded by Fourier analysis and probability theory.
+Among the striking personalities behind this effort is Luke Serafin, who has produced a giant leap forward in the formalization of the CLT \cite{serafin2015formally}.
+Serafin's systematic encoding of the characteristic function method has ensured that it was entirely generic and modular.
+His contributions involved the formalization of key intermediate results such as the for sums of independent random variables and the fact that components of his work are reusable, enabling extension into applications that relate to probabilistic theorems.
+
+The strength of Isabelle/HOL is to borrow from its excellent complex analysis library.
+The characteristic function-based proofs are formalized smoothly, and the modular and extensible framework allows researchers to always build on previous work, gradually increasing the scope of formalized mathematics.
+The CLT formalization within Isabelle/HOL is exemplary in showing how even the most intricate probabilistic results can be encoded rigorously with the help of proof assistants.
+But this also points to some of the challenges with this approach: it is very much based on Fourier transforms and complex-valued methods, an infrastructure which may not always be as easily available when moving to other proof systems.
+Also, characteristic functions are powerful but often complicate proofs in situations where simpler real-valued methods, such as those based on moments, would suffice.
+
+The formalization of CLT in Isabelle/HOL serves as an inspiration and comparison for the research presented in this thesis.
+The Example of Isabelle/HOL shows that the CLT can be formalized with the help of characteristic functions but also brings into relief the limitations of such a strategy-if proof assistants like HOL4 lack libraries comparable for complex analysis with it.
+This thesis takes a different site following the approach with the theorem of Lyapunov, which involves real-valued moment conditions rather than characteristic functions.
+Following this path, the work in HOL4 demonstrates how easier and less convoluted tools enable formulating the CLT while considering the inherent challenges of formal verification.

### 2024-12-12 background.tex
Modified Background section, got problem with citing

@@
 
-The history of the Central Limit Theorem, hereinafter referred to as CLT, is quite a fascinating journey through the gradual development of probability theory.
-Over the years, various versions of CLT have been formulated, reflecting its adaptability and foundational importance.
-Each such variation addresses certain mathematical challenges and applications, therefore enriching the broader understanding of convergence in probability.
-In this chapter, we explore some theorem-proving tools and their methodologies to formalize the CLT, emphasizing ...
+The Central Limit Theory (CLT) is one of the main theoretical results from the  Probability theory, a bridge from individual randomness to aggregate predictability.
+It marks the gradual approach of mathematical rigor and upgrading of the theorem used by addressing more and more difficult problems.
+Over centuries, different forms of CLTs have been derived, each of which has added to its understanding convergence and applications.
+This chapter will present the historical development of the CLT that describes its milestones and contributions.
+It will also examine the modern theorem proving tools and ways of formalization, focusing on Lyapunov's approach.
 

### 2024-12-14 background.tex
Added HOL section in Preliminaries (minor changes in Introduction, Background), citation bugs

@@
 
-The Central Limit Theory (CLT) is one of the main theoretical results from the  Probability theory, a bridge from individual randomness to aggregate predictability.
+The Central Limit Theory is one of the main theoretical results from the  Probability theory, a bridge from individual randomness to aggregate predictability.
 It marks the gradual approach of mathematical rigor and upgrading of the theorem used by addressing more and more difficult problems.
@@
 Characteristic functions are Fourier transforms of probability distributions that make convergence easier to analyze by drawing on their special mathematical properties.
-Central to the approach are two key steps: first, pointwise convergence of the characteristic function for the normalized sum of independent random variables to that of the standard normal distribution should be demonstrated; and second, by the application of the Lévy Continuity Theorem, one obtains the desired convergence in distribution.
+Central to the approach are two key steps: first, pointwise convergence of the characteristic function for the normalized sum of independent random variables to that of the standard normal distribution should be demonstrated; and second, by the application of the L\'evy Continuity Theorem, one obtains the desired convergence in distribution.
 These two steps are actually the core of the proof that eventually leads to the verification of CLT \cite{billingsley2017probability}.

### 2024-12-14 introduction.tex
Added HOL section in Preliminaries (minor changes in Introduction, Background), citation bugs

@@
 However, formalizing CLT is far from straightforward.
-There exist numerous variants of this theorem, together with their numerous proofs, ranging from classical independent and identically distributed (i.i.d.) to some more general versions, such as Lyapunov's theorem \cite{billingsley2017probability}.
+There exist numerous variants of this theorem, together with their numerous proofs, ranging from classical independent and identically distributed (i.i.d.) to some more general versions.
+Notably, Lyapunov's theorem extends the CLT to independent, non-identically distributed (i.n.i.d.) random variables, offering a broader applicability than the classical i.i.d.
+case \cite{billingsley2017probability}.
 Each of those proofs uses advanced mathematical tools, including characteristic functions, moment generating functions, or even higher-order moments, according to the situation.
@@
 
-Instead, this thesis follows another path: formalizing in the HOL4 proof assistant a generalized version of the CLT by Lyapunov's theorem.
-Lyapunov's approach is particularly appealing, as it has no direct use of either complex analysis or characteristic functions.
-Instead, it pays attention to moment-based convergence criteria, trying to provide its conditions for convergence using Taylor expansions, Big-O notation of error bounds, and Lyapunov's inequality.
+Instead, this thesis follows another path: leveraging Lyapunov's theorem and the Lindeberg approach, which avoids characteristic functions in favor of real-valued, moment-based techniques.
+Additionally, it pays attention to moment-based convergence criteria, trying to provide its conditions for convergence using Taylor expansions, Big-O notation of error bounds, and Lyapunov's inequality.
 This approach aligns well with the existing mathematical infrastructure of HOL4 while simultaneously expanding its capabilities.
-
-This project formalizes Lyapunov's proof in HOL4 and demonstrates the feasibility of verifying advanced probabilistic results in HOL4. More specifically, it has added:
-\begin{itemize}
-    \item Formally defined Taylor expansions in function bounds and error terms.
-    \item Big-O, allowing accurate representations of asymptotic behavior.
-    \item Lyapunov's inequality, the key condition for the convergence between the CLT.
-\end{itemize}
 
@@
 
-However, Lyapunov’s method introduced its own set of challenges for formalization in HOL4:
+However, Linderberg’s method introduced its own set of challenges for formalization in HOL4:
 \begin{itemize}
@@
 
-Following the Lyapunov's approach, with its valuation on real techniques and elementary methods, proved most important: it served to encapsulate the CLT under the constraints of HOL4 and was entirely consistent toward building reusable mathematics libraries.
+Following the Linderberg's approach, with its valuation on real techniques and elementary methods, proved most important: it served to encapsulate the CLT under the constraints of HOL4 and was entirely consistent toward building reusable mathematics libraries.
 Taylor series developments, Big-O notion, and Lyapunov's inequality are indispensable not only for this proof but also as a basis of further formalizations in the realm of probability theory and elsewhere.
@@
 
+\section{Thesis Contributions}
+This thesis formalizes Lyapunov's inequality in HOL4, hence establishing a rigorous framework for convergence in probability proofs.
+In particular, the necessary mathematical theories and lemmas for the proof of CLT: Taylor expansions to bound errors, Big-O notation for asymptotic behavior analysis, and telescoping techniques for incremental replacements.
+These lemmas are used to formalize the Lyapunov's theorem and Lindeberg's approach, with a subsequent complete proof of CLT within HOL4.
+Unlike other approaches using characterization or moment-generating functions, this work does not use complex analysis, but works with the methods compatible with the real-valued nature of HOL4.
+
+This thesis, beyond formalizing CLT, enriches libraries for probabilistic analysis that can be reused beyond, such as statistical inference and machine learning.
+This addresses some foundational gaps in HOL4 and demonstrates the use of HOL4 in advancing formalized mathematics.
+
 \section{Structure of the Thesis}
@@
 Related work, including other formalizations, using proof assistants like Isabelle/HOL is also covered.
-Chapter 3 then details the formalization, discussing the challenges faced along with solutions developed, in particular the formalization of Taylor expansions, Big-O notation and Lyapunov's inequality.
-The key components of the proof, its completion, and its validation of the CLT are discussed in Chapter 4.
+Chapter 3 then details the preliminaries, including HOL4 as proof engineering and related libraries.
+The key components of the proof, its completion, and its validation of the CLT are given in Chapter 4.
 Chapter 5 presents a discussion of results, and positioning of this work with respect to related efforts on formal verification.

### 2024-12-14 preliminaries.tex
init commit
Added HOL section in Preliminaries (minor changes in Introduction, Background), citation bugs

--- 2024-11-30 background.tex
@@
-\chapter{Background and Related Work}
-\label{background}
+\chapter[Preliminaries]{Preliminaries}
+\label{chap:3}
 
-Background 123
+This chapter describes an overview of the theoretical and formal theories required for the formalization of the Central Limit Theorem .
+This includes HOL Formalization, Measure Theory, Lebesgue Integration, and Probability Theory.
+\section{HOL Formalization}
+Higher Order Logic (HOL) \cite{hol4, slind2008brief} is derived from the Logic of Computable Functions (LCF) \cite{gordon1979edinburgh, milner1972logic} created by Robin Milner and colleagues in 1972.
+HOL is an adaptation of Church's Simple Theory of Types (STT) \cite{church1940formulation}, where a higher-order version of Hilbert's choice operator \(\varepsilon \), Axiom of Infinity, and Rank-1 polymorphism have been added.
+HOL4 implements the original HOL framework, while other theorem provers in the HOL family, such as Isabelle/HOL, include important extensions.
+Such a simple logical basis makes HOL more accessible than those systems founded on much more advanced dependent type theories, such such the Calculus of Inductive and Co-Inductive Constructions constructed by Coq.
+Therefore, theories and proofs founded on HOL are easier for a layman to comprehend rather than being lost in a complicated type theory.
 
-% \begin{figure*}
-% \centering
-% \includegraphics[width=0.98\textwidth]{Chapter1/Images/Fig1.pdf}
-% \caption{caption}
-% \label{figCh1:figure}
-% \end{figure*}
+HOL refers both to the logical system and the software implementing it.
+HOL4 is the latest version of this software and written in Standard ML (SML), a general-purpose functional programming language.
+SML has played the most vital role in the HOL4 for implementing its core engine, enabled automation due to which proof tactics have been written in that and also for interaction, whether it is through a proof script or in direct correspondence with the user.
+Integrated SML gives a way in which HOL4 is versatile and can easily be extended such that complex verification tools are provided to develop the management of proofs by a user efficiently.
 
-\renewcommand\thefigure{\thechapter .\arabic{figure}}
+% HOL logic is based on a formal system of typed logical terms, where types are set-theoretic interpretation sets in a universe \(U\).
+The HOL type system is, relative to dependent types, considerably less complex yet still suitable for the formalization of various disciplines.
+HOL comes in with four type categories.
+As part of its Backus-Naur Form grammar it allows for:
+% \begin{itemize}
+%     \item \textbf{Primitive types:} For example, \texttt{bool} (representing a two-element set \( \textbf{2} \)) and \texttt{ind} (representing an infinite set \( I \)).
+%     \item \textbf{Type constructors:} Operators creating new types, such as functions (`->`) and Cartesian products (`*`).
+%     \item \textbf{Type variables:} Enable polymorphism; it is thus possible to write generic theorems and functions.
+%     \item \textbf{Defined types:} Derived types, either as subsets or as bijections of basic types.
+% \end{itemize}
+
+HOL terms are representatives of such things and the grammar includes constants, variables, applications (function calls) and lambda abstractions.
+Quantifiers, for instance universal (!x.
+P(x)) and existential (?x.
+P(x)), are also provided in HOL - they are defined as specific lambda functions.
+
+\section{Measure Theory}
+
+\section{Lebesgue Integration Theory}
+
+\section{Probability Theory}

### 2024-12-15 draft.tex
Added new informal proof without introducing Y

@@
 
+\newpage
 \subsection{The proof of Central Limit Theorem}
+\subsubsection{Informal Proof without introducing Y ((the replacement Gaussian random variables))}
+\begin{itemize}
+        \item[] \( \mathbb{E}[X_j] = 0 \), 
+        \item[] \( \text{Var}(X_j) = \sigma_j^2 \) (finite variances, \( \sigma_j^2 < \infty \)),
+        \item[] \( \mathbb{E}[|X_j|^3] = \gamma_j \) (finite third moments, \( \gamma_j < \infty \)).
+    \end{itemize}
+
+Define:
+        \begin{itemize}
+            \item[] \( S_n = \sum_{j=1}^n X_j \),
+            \item[] \( s_n^2 = \sum_{j=1}^n \sigma_j^2 \) (total variance),
+            \item[] \( \Gamma_n = \sum_{j=1}^n \gamma_j \) (total third moment).
+\end{itemize}
+
+Statement:
+
+    If 
+    \[
+   \frac{\Gamma_n}{s_n^3} \to 0 \quad \text{as } n \to \infty.
+   \]
+   then 
+   \[
+   \frac{S_n}{s_n} \xrightarrow{d} \Phi
+   \]
+   where \( \Phi\) denotes the standard normal distribution \( \mathcal{N}(0,1) \).
+
+Let $\boldsymbol{Z_n = S_n / s_n }$, to prove $\boldsymbol{Z_n \xrightarrow{d} \mathcal{N}(0,1)}$, it suffices to show that for any bounded, continuous test function \( f \in C_b^3(\mathbb{R}) \) (bounded continuous functions with three bounded derivatives):
+\[
+\mathbb{E}[f(Z_n)] \to \mathbb{E}[f(\mathcal{N})],
+\]
+
+Using Taylor’s theorem, we expand \( f(Z_n) \) around \( 0 \):
+\[
+f(Z_n) = f(0) + f'(0)Z_n + \frac{f''(0)}{2}Z_n^2 + R(Z_n),
+\]
+where \( R(Z_n) \) is the remainder:
+\[
+R(Z_n) = \frac{f^{(3)}(\xi)}{6}Z_n^3,
+\]
+for some \( \xi \) between \( 0 \) and \( Z_n \).
+
+Taking expectations:
+\[
+\mathbb{E}[f(Z_n)] = f(0) + f'(0)\mathbb{E}[Z_n] + \frac{f''(0)}{2}\mathbb{E}[Z_n^2] + \mathbb{E}[R(Z_n)].
+\]
+
+Similarly, for \( \mathcal{N}(0,1) \), we expand \( f(\mathcal{N}) \):
+\[
+\mathbb{E}[f(\mathcal{N})] = f(0) + f'(0)\mathbb{E}[\mathcal{N}] + \frac{f''(0)}{2}\mathbb{E}[\mathcal{N}^2] + \mathbb{E}[R(\mathcal{N})].
+\]
+
+Thus, the difference becomes:
+\[
+\mathbb{E}[f(Z_n)] - \mathbb{E}[f(\mathcal{N})] = \underbrace{f'(0) (\mathbb{E}[Z_n] - \mathbb{E}[\mathcal{N}])}_{(a)} + \underbrace{\frac{f''(0)}{2} (\mathbb{E}[Z_n^2] - \mathbb{E}[\mathcal{N}^2])}_{(b)} + \underbrace{(\mathbb{E}[R(Z_n)] - \mathbb{E}[R(\mathcal{N})])}_{(c)}.
+\]
+
+By analyzing each term, we have:
+
+\textbf{(a)} $\boldsymbol{f'(0) (\mathbb{E}[Z_n] - \mathbb{E}[\mathcal{N}])}$
+\begin{itemize}
+    \item By assumption, \( \mathbb{E}[X_j] = 0 \), so \( \mathbb{E}[S_n] = 0 \), and:
+  \[
+  \mathbb{E}[Z_n] = \frac{\mathbb{E}[S_n]}{s_n} = 0.
+  \]
+  \item For \(\mathcal{N}(0, 1)\), \( \mathbb{E}[\mathcal{N}] = 0 \).
+\end{itemize}
+Therefore:
+  \[
+  f'(0) (\mathbb{E}[Z_n] - \mathbb{E}[\mathcal{N}]) = 0.
+  \]
+
+(b) $\boldsymbol{\frac{f''(0)}{2} (\mathbb{E}[Z_n^2] - \mathbb{E}[\mathcal{N}^2])}$
+\begin{itemize}
+    \item The variance of \( S_n \) is \( s_n^2 = \sum_{j=1}^n \sigma_j^2 \), so:
+  \[
+  \text{Var}(Z_n) = \frac{\text{Var}(S_n)}{s_n^2} = \frac{s_n^2}{s_n^2} = 1.
+  \]
+    \item Thus:
+  \[
+  \mathbb{E}[Z_n^2] = 1.
+  \]
+    \item For \( \mathcal{N}(0,1) \), \( \mathbb{E}[\mathcal{N}^2] = 1 \).
+\end{itemize}
+Therefore:
+  \[
+  \frac{f''(0)}{2} (\mathbb{E}[Z_n^2] - \mathbb{E}[\mathcal{N}^2]) = 0.
+  \]
+
+(c) $\boldsymbol{\mathbb{E}[R(Z_n)] - \mathbb{E}[R(\mathcal{N})]}$
+
+By Taylor's theorem, We have:
+\[
+\mathbb{E}[R(Z_n)] \leq \frac{M}{6} \mathbb{E}[Z_n^3], 
+\]
+and
+
+\[
+\mathbb{E}[R(\mathcal{N})] \leq \frac{M}{6} \mathbb{E}[\mathcal{N}^3]
+\]
+
+where \( M = \sup_{x \in \mathbb{R}} |f^{(3)}(x)| \) is the maximum bound on the third derivative of \( f \).
+
+Thus, 
+\begin{equation}{\label{final_eq}}
+    \left| \mathbb{E}[R(Z_n)] - \mathbb{E}[R(\mathcal{N})] \right| \leq \frac{M}{6} \left|\mathbb{E}[Z_n^3] - \mathbb{E}[\mathcal{N}^3] \right| 
+\end{equation}
+
+\textbf{For} $\boldsymbol{Z_n}$:
+
+The normalized third moment of \( Z_n = S_n / s_n \) is:
+\begin{equation}{\label{Z_cube}}
+    \mathbb{E}[Z_n^3] = \frac{\mathbb{E}[|S_n|^3]}{s_n^3}.
+\end{equation}
+
+Using Triangle inequality for sums, which states:
+\begin{equation}{\label{lya_Z}}
+    \mathbb{E}[|S_n|^3] = \mathbb{E}\left[\left| \sum_{j=1}^n X_j \right|^3\right] \leq \sum_{j=1}^n \mathbb{E}[|X_j|^3],
+\end{equation}
+
+Combining \eqref{Z_cube} and \eqref{lya_Z} we get:
+\[
+\mathbb{E}[|Z_n|^3] \leq \frac{\sum_{j=1}^n \gamma_j}{s_n^3},
+\]
+where \( \gamma_j = \mathbb{E}[|X_j|^3] \).
+
+Thus:
+\[
+\mathbb{E}[R(Z_n)] \leq \frac{M}{6} \frac{\sum_{j=1}^n \gamma_j}{s_n^3}.
+\]
+
+\textbf{For} $\boldsymbol{\mathcal{N}(0,1)}$:
+The third moment of \( \mathcal{N} \) is constant:
+\begin{equation}{\label{third_moment}}
+    \mathbb{E}[\mathcal{N}^3] = 0,
+\end{equation}
+
+To prove \eqref{third_moment}, we use moment generating function (MGF). The moment generating function (MGF) of a random variable \( X \) is defined as:
+\[
+M_X(t) = \mathbb{E}[e^{tX}].
+\]
+
+For \( X \sim N(0, 1) \), the MGF is known to be:
+\[
+M_X(t) = e^{t^2 / 2}.
+\]
+This follows from integrating the product of \( e^{tx} \) and the PDF of \( N(0, 1) \):
+\[
+M_X(t) = \int_{-\infty}^\infty e^{tx} \frac{1}{\sqrt{2\pi}} e^{-x^2 / 2} \, dx = e^{t^2 / 2}.
+\]
+
+Then, third derivative:
+   \[
+   M_X^{(3)}(t) = \frac{d}{dt} \left( e^{t^2 / 2} + t^2 e^{t^2 / 2} \right).
+   \]
+   Differentiate term by term:
+   \[
+   M_X^{(3)}(t) = t^3 e^{t^2 / 2} + 3t e^{t^2 / 2}.
+   \]
+Now evaluate \( M_X^{(3)}(t) \) at \( t = 0 \):
+\[
+M_X^{(3)}(0) = (0^3 + 3 \cdot 0) e^{0} = 0.
+\]
+
+Thus:
+\[
+\mathbb{E}[X^3] = M_X^{(3)}(0) = 0.
+\]
+
+Hence, the difference simplifies to:
+\[
+\left| \mathbb{E}[Z_n^3] - \mathbb{E}[N^3] \right| = \left| \mathbb{E}[Z_n^3] \right| \leq \frac{\sum_{j=1}^n \gamma_j}{s_n^3} = \frac{\Gamma_n}{s_n^3}.
+\]
+
+Since \( \mathbb{E}[\mathcal{N}^3] = 0 \), the \eqref{final_eq} simplifies to:
+\[
+\left| \mathbb{E}[R(Z_n)] - \mathbb{E}[R(\mathcal{N})] \right| \leq \frac{M}{6} \mathbb{E}[Z_n^3].
+\]
+
+Substituting \( \mathbb{E}[Z_n^3] \leq \frac{\Gamma_n}{s_n^3} \), we get:
+\[
+\mathbb{E}[R(Z_n)] \leq \frac{M}{6} \frac{\Gamma_n}{s_n^3}.
+\]
+
+Finally, by the assumption:
+\[
+\frac{\Gamma_n}{s_n^3} \to 0 \quad \text{as } n \to \infty,
+\]
+it follows that:
+\[
+\mathbb{E}[f(Z_n)] \to \mathbb{E}[f(\mathcal{N})].
+
+   \newpage
+
 \subsubsection{Informal Proof}

### 2024-12-21 preliminaries.tex
Added Chapter 3, including HOL4 and Measure Theory

@@
 
-% HOL logic is based on a formal system of typed logical terms, where types are set-theoretic interpretation sets in a universe \(U\).
-The HOL type system is, relative to dependent types, considerably less complex yet still suitable for the formalization of various disciplines.
-HOL comes in with four type categories.
-As part of its Backus-Naur Form grammar it allows for:
-% \begin{itemize}
-%     \item \textbf{Primitive types:} For example, \texttt{bool} (representing a two-element set \( \textbf{2} \)) and \texttt{ind} (representing an infinite set \( I \)).
-%     \item \textbf{Type constructors:} Operators creating new types, such as functions (`->`) and Cartesian products (`*`).
-%     \item \textbf{Type variables:} Enable polymorphism; it is thus possible to write generic theorems and functions.
-%     \item \textbf{Defined types:} Derived types, either as subsets or as bijections of basic types.
-% \end{itemize}
-
 HOL terms are representatives of such things and the grammar includes constants, variables, applications (function calls) and lambda abstractions.
@@
 
+The type system of HOL establishes the structural framework within which all terms and expressions are guaranteed to be well-defined and logically consistent.
+Types in HOL denote sets within the universe \( U \), and every term bears a certain type.
+The type grammar is simple and very expressive, and thus able to construct a wide variety of mathematical and logical objects.
+
+The type grammar is defined as:
+\[
+\sigma ::= \alpha \ | \ c \ | \ (\sigma_1, \ldots, \sigma_n)\text{op} \ | \ \sigma_1 \to \sigma_2
+\]
+where:
+\begin{enumerate}
+    \item \textbf{Type Variables (\(\alpha, \beta, \ldots\)):} Generic placeholders that allow polymorphism to provide functions and predicates over different types.
+    \begin{itemize}
+        \item Example: The type variable \(\alpha\) could indicate integers, Booleans, or functions.
+    \end{itemize}
+
+    \item \textbf{Atomic Types (\(c\)):} Fixed and pre-defined types within HOL. The two initial atomic types are:
+    \begin{itemize}
+        \item \texttt{bool}: The set of Boolean values \(\{T, F\}\).
+        \item \texttt{ind}: The set composed by individuals (an infinite set).
+
+    \end{itemize}
+
+    \item \textbf{Compound Types (\((\sigma_1, \ldots, \sigma_n)\text{op}\)):} Formed by applying type operators to other types.
+Their examples include Cartesian products, which designate the tuples over the elements.
+    \begin{itemize}
+        \item Example: The type \((\texttt{bool}, \texttt{ind}) \times\) represents pairs of a Boolean and an individual.
+    \end{itemize}
+
+    \item \textbf{Function Types (\(\sigma_1 \to \sigma_2\)):} Represent total functions mapping elements from a domain (\(\sigma_1\)) to a codomain (\(\sigma_2\)).
+    \begin{itemize}
+        \item Example: The type \(\texttt{bool} \to \texttt{ind}\) indicates a function mapping both Boolean-values to individual-elements.
+    \end{itemize}
+\end{enumerate}
+
+For example, consider the following types:
+\begin{enumerate}
+    \item A function from integers to Booleans:
+   \[
+   f : \texttt{int} \to \texttt{bool}
+   \]
+   This type indicates that \(f(x)\) is a function taking an integer \(x\) and returning a Boolean.
+
+    \item A tuple containing a Boolean and a function:
+       \[
+       p : (\texttt{bool}, (\texttt{int} \to \texttt{bool}))
+       \]
+       This is a pair type \(p = (b, f)\), where \(b\) is a Boolean, and \(f\) is a function mapping integers to Booleans.
+       \item The type system guarantees the consistency by making sure all terms are properly typed.
+So if \(g : \texttt{int} \to \texttt{bool}\), then \(g(5)\) as 5 is an integer, but, \(g(T)\) would be invalid since \(T\) is a Boolean, not an integer.
+Such stringent typing is at the level of terms to avoid self-contradictory values and assure that proofs built up in HOL are sound.
+\end{enumerate}
+
+In HOL, terms are representatives for elements of sets represented by their types.
+The grammar of the term deﬁnes the syntax and structure for the logical expressions that can be expressed and hence statements that could be well typed and logically valid.
+Terms in HOL are constructed from the following components:
+
+\[
+t ::= x \ | \ c \ | \ t \ t' \ | \ \lambda x. t
+\]
+where:
+\begin{enumerate}
+    \item \textbf{Variables} \boldsymbol{$(x, y, \ldots):$}
+    \begin{itemize}
+        \item Represent placeholders for elements of a type.
+        \item Example: \(x : \texttt{bool}\) stands for a Boolean variable.
+    \end{itemize}
+
+    \item \textbf{Constants (\(c\)):}
+    \begin{itemize}
+        \item Fixed entities such as \(T\), \(F\), mathematical operators, or predefined functions.
+        \item  Example: The constant \(+\) defines addition for numeric types.
+    \end{itemize}
+
+   \item \textbf{Function Applications (\(t \ t'\)):}
+   \begin{itemize}
+       \item Define the application of a function to an argument. The term \(f(x)\) applies the function \(f\) to the variable \(x\).
+        \item Example: If \(f : \texttt{int} \to \texttt{real}\) and \(x : \texttt{int}\), then \(f(x)\) is a valid term of type \(\texttt{real}\).
+   \end{itemize}
+
+\item \textbf{\(\lambda\)-Abstractions (\(\lambda x. t\)):}
+\begin{itemize}
+    \item Denote anonymous functions where \(x\) is the input variable, and \(t\) is the function body.
+    \item Example: \(\lambda x. x + 1\) defines a function that increments its input by 1.
+\end{itemize}
+
+\end{enumerate}
+
+To ensure consistency, the terms of HOL should be well typed. Given a term \(t_{\sigma}\) of type \(\sigma\), its grammar can be generalized with type annotations:
+
+\[
+t_{\sigma} ::= x_{\sigma} \ | \ c_{\sigma} \ | \ (t_{\sigma_1 \to \sigma_2} \ t'_{\sigma_1})_{\sigma_2} \ | \ (\lambda x_{\sigma_1}. t_{\sigma_2})_{\sigma_1 \to \sigma_2}
+\]
+
+HOL's deductive system is considered the logical foundation for forming and checking a proof.
+HOL's deductive system may consist of eight primitive rules of inference, the definition of new theorems by existing theorems.
+These rules are the basic components and are required for all logical reasoning within HOL, ensuring that proofs are consistent, logically valid, and traced.
+The following are the eight main primitive inference rules in HOL:
+
+\begin{enumerate}
+    \item \textbf{Assumption Introduction (ASSUME):}
+    \begin{itemize}
+        \item Introduces a formula as an assumption.
+        \item Rule:
+          \begin{center}
+            \begin{tabular}{c}
+              \\ \hline
+              $t${\small\verb+ |- +}$t$ \\
+            \end{tabular}
+          \end{center}
+        \item Example: From the assumption \(P\), we conclude \(P\).
+    \end{itemize}
+
+    \item \textbf{Reflexivity (REFL):}
+    \begin{itemize}
+        \item States that any term is equal to itself.
+        \item Rule:
+          \begin{center}
+            \begin{tabular}{c}
+              \\ \hline
+                 {\small\verb+ |- +}$t${\small\verb+ = +}$t$ \\
+            \end{tabular}
+          \end{center}
+        \item Example: For \(x : \texttt{int}\), \(x = x\) is always true.
+    \end{itemize}
+
+    \item \textbf{Beta Conversion (BETA\_CONV):}
+    \begin{itemize}
+        \item Applies substitution in lambda abstractions.
+        \item Rule:
+        \item Example: \((\lambda x. x + 1)(5) \vdash 5 + 1\).
+    \end{itemize}
+
+    \item \textbf{Substitution (SUBST):}
+    \begin{itemize}
+        \item Replaces a term in a formula with another term proven to be equal.
+        \item Rule:
+          \begin{center}
+            \begin{tabular}{c}
+              $\Gamma_1${\small\verb+ |- +} $t_1${\small\verb+=+}$t'_1$ {\small\verb+  +} $\cdots$ {\small\verb+  +}
+              $\Gamma_n${\small\verb+ |- +} $t_n${\small\verb+=+}$t'_n$ {\small\verb+  +}
+              $\Gamma${\small\verb+ |- +} $t[t_1,\ldots,t_n]$ \\ \hline
+              $\Gamma_1 \cup \cdots
+              \cup \Gamma_n \cup \Gamma${\small\verb+ |- +} $t[t'_1,\ldots,t'_n]$ \\
+            \end{tabular}
+          \end{center}
+        \item Example: From \(x = y\) and \(P(x)\), infer \(P(y)\).
+    \end{itemize}
+
+    \item \textbf{Abstraction (ABS):}
+    \begin{itemize}
+        \item Generalizes an equation by abstracting a variable.
+        \item Rule:
+          \begin{center}
+            \begin{tabular}{c}
+              $\Gamma${\small\verb+ |- +}$t_1${\small\verb+ = +}$t_2$ \\ \hline
+              $\Gamma${\small\verb+ |- (\+}$x${\small\verb+.+}$t_1${\small\verb+) = (\+}$x${\small\verb+.+}$t_2${\small\verb+)+} \\
+            \end{tabular}
+          \end{center}
+        \item Example: From \(5 + 1 = 6\), infer \(\lambda x. x + 1 = \lambda x. 6\).
+    \end{itemize}
+
+    \item \textbf{Type Instantiation (INST\_TYPE):}
+    \begin{itemize}
+        \item Specializes polymorphic functions or predicates to specific types.
+        \item Rule:
+          \begin{center}
+            \begin{tabular}{c}
+              $\Gamma${\small\verb+ |- +}$t$ \\ \hline
+              $\Gamma[\sigma_1,\ \ldots\ ,\sigma_n/\alpha_1,\ \ldots\ ,\alpha_n]${\small\verb+ |- +}$t[\sigma_1,\ \ldots\ ,\sigma_n/\alpha_1,\ \ldots\ ,\alpha_n]$
+            \end{tabular}
+          \end{center}
+    \end{itemize}
+
+    \item \textbf{Discharging Assumptions (DISCH):}
+    \begin{itemize}
+        \item Converts an assumption into an implication.
+        \item Rule:
+          \begin{center}
+            \begin{tabular}{c}
+              $\Gamma${\small\verb+ |- +} $t_2$ \\ \hline
+              $\Gamma{-}\{t_1\}${\small\verb+ |- +} $t_1${\small\verb+ ==> +}$t_2$
+            \end{tabular}
+          \end{center}
+        \item Example: From \(P \wedge Q\), infer \(P \Rightarrow (Q \wedge P)\).
+    \end{itemize}
+
+    \item \textbf{Modus Ponens (MP):}
+    \begin{itemize}
+        \item Combines an implication and its premise to infer the conclusion.
+        \item Rule:
+          \begin{center}
+            \begin{tabular}{c}
+              $\Gamma_1${\small\verb+ |- +}$t_1${\small\verb+ ==> +}$t_2$ {\small\verb+     +} $\Gamma_2${\small\verb+ |- +}$t_1$ \\
+              \hline
+              $\Gamma_1 \cup \Gamma_2${\small\verb+ |- +}$t_2$ \\
+            \end{tabular}
+          \end{center}
+        \item Example: From \(x > 0 \Rightarrow x^2 > 0\) and \(x > 0\), infer \(x^2 > 0\).
+    \end{itemize}
+\end{enumerate}
+These inference rules ensure that all logical derivations are traceable to basic axioms and established theorems.
+Additionally, the deductive system forms the backbone of HOL4, ensuring that proofs are both rigorous and reliable.
+
+All proofs in HOL are fundamentally derived from a set of primitive inference rules and a core logical foundation.
+These rules define the semantics of two fundamental logical connectives: \textbf{equality} (\(=\)) and \textbf{implication} (\(\implies\)).
+Other logical connectives and first\-order quantifiers, such as logical truth (\(T\)), falsehood (\(F\)), conjunction (\(\wedge\)), disjunction (\(\vee\)), and existential quantification (\(\exists\)), are defined as lambda (\(\lambda\)) functions for consistency within the HOL framework:
+
+\begin{enumerate}
+    \item \textbf{Logical Truth (\(T\))}
+    \begin{itemize}
+        \item Rule
+\begin{verbatim}
+  T_DEF              |- T  = ((\x:bool. x) = (\x. x))
+
+\end{verbatim}
+        \item True is represented as the equality of two identical boolean functions.
+    \end{itemize}
+
+    \item \textbf{Logical Falsehood (\(F\))}
+    \begin{itemize}
+        \item Rule
+\begin{verbatim}
+  F_DEF              |- F  = !t. t
+\end{verbatim}
+        \item False is defined to satisfy any boolean implication.
+    \end{itemize}
+
+    \item \textbf{Negation (\(\neg\))}
+    \begin{itemize}
+        \item Rule
+\begin{verbatim}
+   NOT_DEF            |- ¬  = (\t. t ==> F)
+\end{verbatim}
+        \item Negation is the implication of a boolean value leading to falsehood.
+    \end{itemize}
+
+    \item \textbf{Conjunction (\(\wedge\))}
+    \begin{itemize}
+        \item Rule
+\begin{verbatim}
+  AND_DEF            |- /\ = \t1 t2. !t. (t1 ==> t2 ==> t) ==> t
+\end{verbatim}
+        \item Conjunction is defined as a logical function that evaluates nested implications.
+    \end{itemize}
+
+    \item \textbf{Disjunction (\(\vee\))}
+    \begin{itemize}
+        \item Rule
+\begin{verbatim}
+  OR_DEF             |- \/ = \t1 t2. !t. (t1 ==> t) ==> (t2 ==> t) ==> t
+
+\end{verbatim}
+        \item Disjunction is expressed through sequential implications.
+    \end{itemize}
+
+    \item \textbf{Universal Quantifier (\(\forall\))}
+    \begin{itemize}
+        \item Rule
+\begin{verbatim}
+
+  FORALL_DEF         |- !  = \P:'a->bool. P = (\x. T)
+
+\end{verbatim}
+        \item Universality asserts that a predicate holds for all elements of a type.
+    \end{itemize}
+
+    \item \textbf{Existential Quantifier (\(\exists\))}
+    \begin{itemize}
+        \item Rule
+\begin{verbatim}
+
+  EXISTS_DEF         |- ?  = \P:'a->bool. P($@ P)
+
+\end{verbatim}
+        \item Existence is defined using Hilbert’s choice operator (\(\varepsilon\)).
+    \end{itemize}
+
+HOL also defines constructs for mathematical operations, such as \textbf{one-to-one functions} (\(One\_One\)) and \textbf{onto functions} (\(Onto\)), to extend logical capabilities:
+
+    \item \textbf{One-to-One (\(ONE\_ONE\_DEF\))}
+    \begin{itemize}
+        \item Rule
+          \begin{alltt}
+            \(\!\!\!{\turn}\!\!\!\!\) ONE_ONE = (\(\lambda\)f. \(\forall\)x1 x2. f x1 = f x2 \(\Rightarrow\) x1 = x2)
+          \end{alltt}
+    \end{itemize}
+
+    \item \textbf{Onto (\(ONTO\_DEF\))}
+    \begin{itemize}
+        \item Rule
+          \begin{alltt}
+            \(\!\!\!{\turn}\!\!\!\!\) ONTO = (\(\lambda\)f. \(\forall\)y. \(\exists\)x. y = f x)
+          \end{alltt}
+    \end{itemize}
+
+HOL includes the constant \(Type\_Definition\), which defines new types as bijections of subsets of existing types:
+    \item \textbf{Type Definition \(Type\_Definition\)}
+    \begin{itemize}
+        \item Rule
+        \item This process is automated by the HOL Datatype package, simplifying the creation of new types.
+
+    \end{itemize}
+
+\end{enumerate}
+
+HOL’s standard theory is built upon four foundational axioms:
+\begin{enumerate}
+    \item \textbf{Boolean Cases (BOOL\_CASES\_AX)}
+    \begin{itemize}
+        \item Rule
+          \begin{alltt}
+            \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)t. (t \({\Leftrightarrow}\) T) \(\lor\) (t \({\Leftrightarrow}\) F)
+          \end{alltt}
+        \item This axiom ensures that any boolean value is either true or false.
+    \end{itemize}
+
+    \item \textbf{Eta Conversion (ETA\_AX)}
+    \begin{itemize}
+        \item Rule
+          \begin{alltt}
+            \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)t. (\(\lambda\)x. t x) = t
+          \end{alltt}
+        \item Eta conversion describes the extensionality of functions.
+    \end{itemize}
+
+    \item \textbf{Hilbert’s Choice (SELECT\_AX)}
+    \begin{itemize}
+        \item Rule
+          \begin{alltt}
+            \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)P x. P x \(\Rightarrow\) P ($@ P)
+          \end{alltt}
+        \item This axiom relates the choice operator to existential quantification.
+    \end{itemize}
+
+    \item \textbf{Infinity (INFINITY\_AX)}
+    \begin{itemize}
+        \item Rule
+          \begin{alltt}
+            \(\!\!\!{\turn}\!\!\!\!\) \(\exists\)f. ONE_ONE f \(\land\) \(\neg\)ONTO f
+          \end{alltt}
+        \item The Axiom of Infinity ensures the existence of an infinite set.
+    \end{itemize}
+
+\end{enumerate}
+
+These axioms are generally sufficient for conventional formalization projects in HOL4. Adding new axioms is strongly discouraged, as it can compromise logical consistency.
+
 \section{Measure Theory}
+
+Measure theory in HOL4 is the theoretical foundation on which measurements like size, content, or probability can be given to sets.
+It bases its formalizations on measures, which are functions taking specific collections of subsets, called sigma algebras, as arguments.
+
+\begin{enumerate}
+    \item \textbf{Measures and Sigma Algebra}
+
+    A measure in HOL4 is defined as a set function of type \((\alpha \to \text{bool}) \to \text{extreal}\), where \(\text{extreal}\) accounts for extended real numbers to handle infinities.
+For example, the measure of the entire real line \(\mathbb{R}\) under the Lebesgue measure is represented as \(+\infty\) (\(\texttt{PosInf}\)).
+
+    In order for measures to be consistent, they should satisfy the following properties:
+    \begin{enumerate}
+        \item \textbf{Non-Negativity:} \(\mu(A) \geq 0\) for all measurable sets \(A\).
+        \item \textbf{Null Empty Set:} \(\mu(\emptyset) = 0\).
+        \item \textbf{Countable Additivity:} For any countable collection of disjoint sets \(\{A_n\}\):
+   \[
+   \mu\left(\bigcup_{n=1}^\infty A_n\right) = \sum_{n=1}^\infty \mu(A_n).
+   \]
+    \end{enumerate}
+    Measures form over some sigma algebra (\(\sigma\)-algebra) which are systems of subsets that preserve their countable operations like unions, intersections, and complements.
+A formal definition of \(\sigma\)-algebra is:
+\[
+\text{sigma\_algebra}(a)
+\]
+where `subsets(a)` denotes the subsets of a \(\sigma\)-algebra.
+
+An equivalent definition of \(\sigma\)-algebra is also provided in HOL4 making their construction flexible:
+\[
+\text{SIGMA\_ALGEBRA\_ALT\_SPACE}:\
+\]
+
+    \item \textbf{Measure Spaces}
+The tuples (sp; sts; µ) refer to a measure space in HOL4
+In HOL4, the tuple \((\text{sp}, \text{sts}, \mu)\) refer to a \textbf{measure space}, where:
+\begin{itemize}
+    \item \(\text{sp}\): The underlying set (space).
+    \item \(\text{sts}\): The \(\sigma\)-algebra of measurable subsets.
+    \item \(\mu\): The measure function.
+\end{itemize}
+
+Three functions, \texttt{m\_space}, \texttt{measurable\_sets}, and \texttt{measure} give \(\text{sp}\), \(\text{sts}\), and \(\mu\). or a measure space, the following conditions must hold:
+\begin{itemize}
+    \item \(\sigma\)-algebra property: \((\text{sp}, \text{sts})\) is a valid \(\sigma\)-algebra.
+    \item \textbf{Positivity:} \( \mu(\emptyset) = 0 \) and \( \mu(A) \geq 0 \) for all \( A \in \text{sts} \).
+    \item \textbf{Countable Additivity:} For any countable disjoint collection \( \{A_n\} \subseteq \text{sts} \):
+   \[
+   \mu\left(\bigcup_{n=1}^\infty A_n\right) = \sum_{n=1}^\infty \mu(A_n).
+   \]
+\end{itemize}
+
+In some occasions, the measures defined on smaller systems of sets like semirings are extended to full \(\sigma\)-algebra by using Carathéodory's extension theorem.
+Such intermediate measures are said to be \texttt{premeasures}.
+
+\item \textbf{Semirings and Generated \(\sigma\)-algebra}
+
+A semiring is a simpler system of subsets that can generate a \(\sigma\)-algebra. It satisfies:
+\begin{enumerate}
+    \item Stability under intersection: \( A, B \in S \implies A \cap B \in S \).
+    \item Decomposition of differences: For \(A, B \in S\), \(A \setminus B\) can be expressed as a finite disjoint union of subsets from \(S\).
+\end{enumerate}
+
+An example is the set of all right-open intervals \([a, b)\) in \(\mathbb{R}\), which forms a semiring.
+For any system of sets \((\text{sp}, \text{sts})\), the smallest \(\sigma\)-algebra containing it, denoted \(\sigma(\text{sp}, \text{sts})\), is defined as:
+\[
+\sigma(\text{sp}, \text{sts}) = (\text{sp}, \bigcap \{s \mid \text{sts} \subseteq s \land \text{sigma\_algebra}(\text{sp}, s)\}).
+\]
+
+\item \textbf{Borel Sigma Algebras}
+
+The \textbf{Borel Sigma Algebra} is the \(\sigma\)-algebra generated by the open subsets of \(\mathbb{R}\), denoted \(\mathcal{B}(\mathbb{R})\). In HOL4, it is formally defined as:
+\[
+\text{borel} = \sigma(\mathbb{R}, \{\text{open sets in } \mathbb{R}\}).
+\]
+
+Alternate characterizations include:
+
+\begin{enumerate}
+    \item Using intervals, e.g., half-open intervals \((a, b]\):
+   \[
+   \text{borel} = \sigma(\mathbb{R}, \{\{x \mid a < x \leq b\} \mid a, b \in \mathbb{R}\}).
+   \]
+
+    \item Using half-spaces:
+   \[
+   \text{borel} = \sigma(\mathbb{R}, \{\{x \mid x \leq a\} \mid a \in \mathbb{R}\}).
+   \]
+
+\end{enumerate}
+The Borel \(\sigma\)-algebra is also important in the definition of measurable functions and measurable spaces in analysis.
+
+    \item \textbf{Measurable Functions}
+
+In HOL4, a function \(f : \text{space}(a) \to \text{space}(b)\) is measurable if:
+\[
+\forall s \in \text{subsets}(b).\, f^{-1}(s) \cap \text{space}(a) \in \text{subsets}(a).
+\]
+This ensures that the preimage of every measurable set in the codomain is measurable in the domain.
+
+\end{enumerate}
+
+HOL4's formalization of measure theory provides a rigorous framework for constructing and reasoning about measures, \(\sigma\)-algebra, and measurable functions.
+The definitions and mathematics make theorems that can be extended to more advanced levels and results.
 

### 2025-01-02 preliminaries.tex
Minor modified Preliminaries chapter

@@
         \item Rule
+          \begin{hol}
+\begin{verbatim}
+  |- TYPE_DEFINITION (P:'a->bool) (rep:'b->'a) =
+  (!x' x''. (rep x' = rep x'') ==> (x' = x'')) /\
+  (!x. P x = (?x'. x = rep x'))
+\end{verbatim}
+          \end{hol}
         \item This process is automated by the HOL Datatype package, simplifying the creation of new types.
@@
 An equivalent definition of \(\sigma\)-algebra is also provided in HOL4 making their construction flexible:
-\[
-\text{SIGMA\_ALGEBRA\_ALT\_SPACE}:\
-\]
+\begin{hol}
+  \begin{alltt}
+    SIGMA_ALGEBRA_ALT_SPACE
+    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)a. sigma_algebra a \({\Leftrightarrow}\)
+    subset_class (space a) (subsets a) \(\land\) space a \(\in\) subsets a \(\land\)
+    (\(\forall\)s. s \(\in\) subsets a \(\Rightarrow\) space a DIFF s \(\in\) subsets a) \(\land\)
+    \(\forall\)f. f \(\in\) (\(\mathbb{U}\)(:num) \(\to\) subsets a) \(\Rightarrow\)
+    BIGUNION (IMAGE f \(\mathbb{U}\)(:num)) \(\in\) subsets a
+  \end{alltt}
+\end{hol}
 
@@
 
+   \begin{hol}
+     \begin{alltt}
+   right_open_interval
+   \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)a b. right_open_interval a b = \{x | a \(\le\) x \(\land\) x < b\}
+
+   right_open_intervals
+   \(\!\!\!{\turn}\!\!\!\!\) right_open_intervals = (\(\mathbb{U}\)(:real),\{right_open_interval a b | T\})
+     \end{alltt}
+   \end{hol}
+   \begin{hol}
+     \begin{alltt}
+   right_open_intervals_semiring
+   \(\!\!\!{\turn}\!\!\!\!\) semiring right_open_intervals
+
+   right_open_intervals_sigma_borel
+   \(\!\!\!{\turn}\!\!\!\!\) sigma (space right_open_intervals) (subsets right_open_intervals) =
+       borel
+     \end{alltt}
+   \end{hol}
 \end{enumerate}
+
 The Borel \(\sigma\)-algebra is also important in the definition of measurable functions and measurable spaces in analysis.
@@
 
+\begin{hol}
+  \begin{alltt}
+    measurable_def
+    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)a b.
+    measurable a b =
+    \{f |
+    f \(\in\) (space a \(\to\) space b) \(\land\)
+    \(\forall\)s. s \(\in\) subsets b \(\Rightarrow\) PREIMAGE f s \(\cap\) space a \(\in\) subsets a\}
+  \end{alltt}
+\end{hol}
+
 \end{enumerate}

### 2025-03-19 background.tex
Minor modified Preliminaries chapter
fixed latex packages

@@
 
-\subsection{Related Work}
+\section{Related Work}
 
 Formalization of mathematical theorems as proofs has been a singular research focus in modern computational mathematics.
-The systems such as Isabelle/HOL \cite{isabelle}, Coq \cite{bertot2013interactive}, and HOL4 \cite{slind2008brief} have been heavily relied on over the years to encode and prove basic properties but offer several features and challenges.
+The systems such as Isabelle/HOL \citep{isabelle}, Coq \citep{bertot2013interactive}, and HOL4 \citep{slind2008brief} have been heavily relied on over the years to encode and prove basic properties but offer several features and challenges.
 

### 2025-05-12 introduction.tex
Modified Ch. Introduction

--- 2024-12-03 introduction.tex
@@
-\chapter{Introduction}
-\label{intro}
+%%% -*- Mode: LaTeX; -*-
+\chapter[Introduction]{Introduction}
+\label{chap:1}
 
-Formal verification has become an essential approach in modern mathematics and computer science; it provides a strong framework for proving mathematical proofs and algorithms.
-Traditional techniques of formal verification include the following: Model Checking \cite{clarke2018handbook}, Testing \cite{broy2005model}, and Theorem Proving \cite{bertot2013interactive}.
+The idea of a formalised mechanist is simple but powerful: if we want to trust mathematics in software, we must mechanise not only calculations but also proofs.
+A mechanist builds a logic engine where every theorem arises from a small set of primitive rules.
+If those rules are sound, then everything derived from them inherits that soundness.
 
-Model Checking (MC) is a well-valued static formal verification method since, when applied correctly, it provides complete assurance that something satisfies certain specifications.
-Model checking is a completely automated procedure and is considered easier compared to other techniques such as theorem proving.
-However, because of the state explosion problem \cite{burch1992symbolic}, application of model checking in large systems is quite a challenge due to the immense state space size of the systems analyzed.
-Also, the success of model checking strongly depends on the quality of the analyzed models.
-In contrast, testing, despite being highly valued for revealing defects in systems, is inherently limited, since it can only indicate the presence of errors, but lacks the ability to verify the correctness of a system.
+HOL4 supports this mechanised philosophy~\cite{slind2008brief}.
+It encodes logic using a typed \(\lambda\)-calculus and enforces soundness by restricting theorem creation to a protected MetaLanguage (ML) type.
+The only way to construct a theorem is by applying primitive inference rules written in Standard ML.
+This ensures that the proof kernel checks every line of reasoning with absolute rigour.
 
-Theorem proving is another widely used formal analysis technique.
-It does not suffer from the limitations of state space size as model checking does, which allows the analysis of larger and more complex systems.
-Furthermore, theorem provers use very expressive logics, like first-order or higher-order logics, which enable the study of a wider range of systems without the restrictions that often come with modeling.
-The most prominent provers are HOL4 \cite{slind2008brief}, HOL Light \cite{hollight}, Coq \cite{bertot2013interactive}, and Isabelle/HOL \cite{isabelle}, all broadly used in the community.
+In this thesis, we formalise one of the most important results in probability theory: the Central Limit Theorem (CLT).
+When many independent effects accumulate, their combined influence often resembles a bell curve.
+We observe this phenomenon across a wide range of domains—from biological traits and human measurements to economic indicators and communication errors.
+The CLT explains this universality.
 
-Formalization plays a vital role in probability theory to ensure that basic results are rigorous mathematically and computable in automated reasoning systems.
-This project formalizes the Central Limit Theorem using the HOL4 proof assistant, hence enriching the domain of probability in formalized mathematics.
+We follow the version of the Central Limit Theorem stated in~\cite{chung2000course} and proved using Lindeberg’s method. Let $\{X_j\}_{j=1}^n$ be a sequence of independent random variables such that:
+
+$$
+\mathbb{E}[X_j] = 0,\quad \mathbb{E}[X_j^2] = \sigma_j^2,\quad \mathbb{E}[|X_j|^3] = \gamma_j < \infty.
+$$
+
+Define the sum $S_n = \sum_{j=1}^n X_j$, the total variance $s_n^2 = \sum_{j=1}^n \sigma_j^2$, and the total third moment $\Gamma_n = \sum_{j=1}^n \gamma_j$. If the Lyapunov condition
+
+$$
+\lim_{n \to \infty} \frac{\Gamma_n}{s_n^3} = 0
+$$
+
+holds, then the normalised sum \(\frac{S_n}{s_n}\) converges in distribution to the standard normal:
+
+$$
+\frac{S_n}{s_n} \xrightarrow{d} \mathcal{N}(0, 1).
+$$
+
+This convergence result explains why the normal distribution emerges from a wide variety of processes, even when individual random variables follow skewed or irregular distributions.
+
+We do not assume probability theory or integration as axioms.
+Instead, we construct them from first principles in HOL4 using its existing real number, measure theory, and Lebesgue integration libraries.
+We define random variables, convergence in distribution, and weak limits in this framework.
+We then prove the CLT using the Lindeberg replacement method and Taylor expansion bounds, following the structure outlined in~\cite{Chung:2001}.
+Here is the Lyapunov's form of the Central Limit Theorem in HOL4:
+\begin{hol}
+  \begin{alltt}
+    Theorem central_limit_theorem :
+    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p X N. prob_space p \(\land\)
+    ext_normal_rv N p 0 1 \(\land\)
+    (\(\forall\)i. real_random_variable (X i) p) \(\land\)
+    (\(\forall\)n. indep_vars p X (\(\lambda\)i. Borel) (count n)) \(\land\)
+    (\(\forall\)i. expectation p (X i) = 0) \(\land\)
+    (\(\forall\)i. expectation p (\(\lambda\)x. (abs (X i x)) pow 3) < +\(\infty\)) \(\land\)
+    (\(\forall\)i. variance p (X i) +\(\infty\)) \(\land\)
+    (\(\forall\)i. variance p (X i) \(\ne\) 0) \(\land\)
+    (\(\forall\)n. (sqrt (second_moments p X n)) \(\ne\) 0) \(\land\)
+    ((\(\lambda\)n. (third_moments p X n) / ((sqrt (second_moments p X n)) pow 3)) --> 0)
+    sequentially \({\Leftrightarrow}\)
+    ((\(\lambda\)n x. (SIGMA (\(\lambda\)i. X i x) (count n)) / (sqrt (second_moments p X n))) --> N)
+    (in_distribution p)
+  \end{alltt}
+\end{hol}
+
+%-----------------------------------------------------
+
+%This formalisation produces a complete, machine-checked proof of Lyapunov’s Central Limit Theorem.
+It not only demonstrates the power of HOL4 for probabilistic reasoning, but also enriches its measure-theoretic library with tools necessary for formalising deeper results in probability and statistics.
+%\newline
+%-----------------------------------------------------
+
+Despite not yet completing the final convergence proof, this thesis represents a significant step toward a fully mechanised proof of Lyapunov’s Central Limit Theorem in HOL4.
+The author developed over 6000 lines of formalisation code, building most of the necessary measure-theoretic and probabilistic infrastructure.
+Early progress was slowed by the need to learn the HOL4 system and an initial attempt to formalise the CLT using moment-generating functions—a path that proved impractical due to missing theorems in the HOL4 library at the time.
+The current approach, based on Lindeberg's method and Taylor expansion, has succeeded in formalising all major components of the proof.
+The only remaining step—formalising the Taylor error bound  and completing the convergence argument—involves calculating the absolute third moment of the normal distribution.
+This means formalizing integrals using the gamma function and complex numbers, which are currently not yet fully realized in HOL4.
+At the time of writing the report, the author is still in the midst of this last step.
+Even if it is still incomplete at the submission deadline, the overall structure and supporting lemmas are all in place, and the last bound is now a question of careful implementation instead of conceptual difficulty.
 
 \section{Motivation}
-The Central Limit Theorem is one of the most fundamental results in the Probability Theory.
-This theory finds broad applications in statistics, data science, finance, and engineering.
-The theorem underlines that under certain conditions, the sum of many independent random variables will, as the number goes large, converge in distribution to a normal distribution irrespective of the individual distributions of those variables \cite{chung2000course}.
-This property justifies many real-world applications and models.
+The Central Limit Theorem (CLT) is more than just a result in probability theory—it is the mathematical explanation behind the bell curve that appears across nature, science, and data.
+When we take a large number of independent random variables with finite variance, their properly normalised sum tends toward a normal distribution, regardless of the original distributions of the individual variables~\cite{fischer2011history}.This effect manifests in physical measurements, biological readings, economic results, and even in computation roundoff errors.
+The CLT illustrates the reasons why such behavior is not accidental but is necessary mathematically under gentle hypotheses.
 
-Although HOL4 already contains formal proofs of some fundamental probabilistic results, such as the Law of Large Numbers, a formalized proof of the Central Limit Theorem is still absent.
-Completing a proof would realize several important milestones, including the following:
+Although succinctly phrased, the CLT is of very deep mathematical organization.
+Most textbook proofs rely on informal approximations or skip foundational details.
+In formal verification, this level of vagueness is unacceptable.
+We need a version of the CLT where every assumption is explicit, every step justified, and the entire argument checked by machine.
+
+Some existing formal proofs of the CLT, such as the one in Isabelle/HOL~\cite{serafin2015formally}, focus on the independent identical distribution (i.i.d.) case and rely on characteristic functions and the Lévy Continuity Theorem.
+That approach is elegant, but it depends on complex analysis and Fourier transforms, and does not generalise easily to sums of non-identically distributed variables.
+Moreover, HOL4’s complex number library is still underdeveloped and lacks many of the theorems needed to support such an approach effectively.
+Attempting to formalise the necessary complex theory would demand substantial additional effort—likely more than what is needed for the Lindeberg-based approach we follow in this work.
+
+In this thesis, we aim for something different. We want to formalise a version of the CLT that:
+\begin{itemize}
+\item allows non-identically distributed variables;
+\item avoids characteristic functions entirely;
+\item uses only measure-theoretic and analytical tools available in HOL4.
+\end{itemize}
+
+We follow the proof given in~\cite{Chung:2001}, based on Lindeberg’s replacement trick and Taylor expansion bounds.
+This approach aligns closely with classical analysis and is modular enough to support future generalisations—such as multidimensional CLTs or stable laws.
+
+By formalising this proof in HOL4, we not only demonstrate that the CLT is derivable in a foundational logical system, but also contribute tools and theorems that strengthen HOL4’s probability theory library for future formalisation work.
+
+\section{Thesis Contributions}
+The contribution of this work is the formalisation of Lyapunov's form of the Central Limit Theorem in HOL4, using the Lindeberg replacement strategy and Taylor approximation techniques.
+This required building up the necessary infrastructure for real-valued random variables, expectations, variances, and convergence in distribution from the ground up.
+
+In particular:
 
 \begin{itemize}
-    \item \textbf{Enriching the HOL4 Theory Library}:
-    \begin{itemize}
-        \item The gap in the library would be filled by adding some advanced results from probability theory.
-        \item A more general foundation would be laid for future processes of formal verification.
-    \end{itemize}
-    \item \textbf{Supporting Future Applications}:
-    \begin{itemize}
-        \item Strengthening the framework for developers and researchers to model or verify real-world problems involving probabilistic reasoning.
-    \end{itemize}
-    \item \textbf{Providing Analytical Tools}:
-        \begin{itemize}
-            \item Enable efficient analysis of problems dependent on the normal distribution and its characteristics.
-        \end{itemize}
+\item We constructed the formal setting for expressing the CLT in Lyapunov form, augmenting HOL4 with normalisation tools, moment summation tools, and standardisation tools.
+\item We established the background required to define and establish bounds on expectations using Taylor expansion, including a remainder term of order three.
+\item We proved key lemmas for the Lindeberg substitution trick, which summarizes how to replace original variables by Gaussian approximations and bound resulting error.
+\item We defined the final convergence goal in terms of convergence in distribution with a well-defined formulation and separation of technical specifications.
 \end{itemize}
 
-Formalizing the Central Limit Theorem also aligns with ongoing efforts to bridge traditional mathematical insights with computational tools.
-As a consequence, it paves the way for applications in areas such as artificial intelligence, machine learning, and quantitative modeling, where probabilistic reasoning is increasingly critical.
-
-This project will contribute to a stronger, more versatile foundation for probabilistic formalization, empowering researchers and developers with better tools to tackle complex, real-world challenges.
+\section{Structure of the Thesis}
+The remainder of this thesis is organised as follows.
+Chapter 2 reviews the Central Limit Theorem, its historical development, and the version we aim to formalise.
+It also surveys related work, including formal proofs in systems such as Isabelle/HOL.
+Chapter 3 introduces the HOL4 theorem prover and outlines the relevant libraries and foundational setup.
+Chapter 4 presents the core of the formalisation, including the supporting lemmas and structure of the proof.
+Chapter 5 discusses the results and places this work in the context of other efforts in formalised mathematics.Chapter 6 summarises the thesis by stating the contributions and providing guidance for future development of probability theory in HOL4.
+This research proves that a formalisation of the Central Limit Theorem in HOL4 is possible.
+It enriches HOL4's probability library, provides reusable tools for future developments, and establishes the system as a platform for fully verified mathematics.

### 2025-05-19 background.tex
Modified Ch. Introduction
Almost done report

@@
-\chapter[Background and Related Work]{Background and Related Work}
+\chapter{Background and Related Work}
 \label{chap:2}
 
-% \section{Background and Related Work}
-% \textbf{Purpose:} Position this work in the context of previous efforts and provide foundational concepts.
-% \subsection{Overview of the Central Limit Theorem}
-% \begin{itemize}
-%     \item Explain the statement of the CLT and its generalizations (e.g., Lyapunov’s and Lindeberg’s approaches).
-%     \item Focus on Lyapunov’s version as the basis of this proof:
+\section{Background: The Central Limit Theorem}
 
-%      \begin{equation}
-%      \text{If } \frac{\Gamma_n}{s_n^3} \to 0 \text{ as } \space \space n \to \infty,
-%      \end{equation}
-%      then
-%      \[
-%      \frac{S_n}{s_n} \xrightarrow{d} \Phi.
-%      \]
-%      Define key terms: \( \Gamma_n, s_n, S_n \), and Lyapunov’s condition.
-% \end{itemize}
+The central limit theory is one of the fundamental results of the Probability theory, a theoretical connection from individual randomness to collective predictability.
+It states the empirical fact that the sum of large numbers of independent random variables converges to a normal distribution, regardless of the individual distributions, as long as some general assumptions hold.
+This theorem explains the universality of the normal distribution in nature and is the foundation for an extremely broad range of applications in statistics, physics, and computer science.
 
-The Central Limit Theory is one of the main theoretical results from the  Probability theory, a bridge from individual randomness to aggregate predictability.
-It marks the gradual approach of mathematical rigor and upgrading of the theorem used by addressing more and more difficult problems.
-Over centuries, different forms of CLTs have been derived, each of which has added to its understanding convergence and applications.
-This chapter will present the historical development of the CLT that describes its milestones and contributions.
-It will also examine the modern theorem proving tools and ways of formalization, focusing on Lyapunov's approach.
+Historically, the beginnings of the CLT trace back to Abraham de Moivre, who showed in 1733 that the binomial distribution is approximated very well by the normal distribution if the number of trials is large \cite{fischer2011history}.
+Pierre-Simon Laplace later generalised this result and formalised it in terms of the Laplace–de Moivre theorem.
+Chebyshev and his students, especially Lyapunov and Markov, pushed the theory further by relaxing the identically distributed condition and replacing it with moment conditions.
+Lyapunov’s 1901 version introduced a now-famous condition on the third absolute moment, which is still the cornerstone for general versions of the CLT.
 
-\section{Historical Background}
-Let us just trace briefly the historical evolution of the CLT, critical milestones, key contributors, and the gradual refinement of its conditions and scope.
-The roots of the CLT go back to the dispute of the eighteenth century, specifically to the time Abraham de Moivre laid the foundation for this theorem.
-In 1733, de Moivre showed that the sum of a large number of independent and identically distributed random variables converges to the normal distribution-an elegant approximation to problems such as the rolling of dice or repeated coin flips \cite{de1733approximatio}.
-His work initiated the interplay between discrete distributions-such as the Binomial-and their continuous approximations, and laid the ground for developments to come.
+A full history of the development of the CLT can be found in Fischer’s \emph{A History of the Central Limit Theorem} \cite{fischer2011history}, which chronicles how the theorem evolved from numerical approximations to a fundamental limit law in probability.
+The evolution reflects a broader trend in mathematics: from combinatorial methods to analysis, and finally, to measure-theoretic and functional analytic foundations.
 
-In 1810, Pierre-Simon Laplace independently extended de Moivre's insights by proving a much more general form of this theorem.
-Under broader conditions, Laplace demonstrated convergence of the sum of independent random variables to a normal distribution, thereby establishing CLT as a universal principle.
-His work was used in applying the CLT to population statistics and measurement errors \cite{laplace1835oeuvres} and cemented its usefulness for aggregate phenomena analysis.
-
-The 19th century would see the formalization of the central limit theorem (CLT) by Pafnuty Chebyshev, who also gave the theorem a rigorous foundation by the introduction of variance and expectation conditions.
-With that work, Chebyshev rendered the CLT mathematically precise, linking it to the developing area of probability theory \cite{chebyshev1890deux}.
-Building on this work was further advancing the generalization of the CLT by Alexander Lyapunov to sequences of independent random variables that are not identically distributed.
-His version of the theorem brought forth Lyapunov's condition, which contained definite criteria for convergence \cite{lyapunov1895pafnutii}.
-
-In the early years of the 20th century, contributions by Lyapunov represented such a change in the evolution of the CLT.
-Unlike his predecessors, Lyapunov sought a “direct” and “elementary” proof.
-It would try to make the theorem internal relations simpler and relax the theorem's assumptions.
-His papers published in 1900 and 1901 reconstructed previous methods with a rigor of Weierstrass, and introduced new precision to the theorem by establishing explicit bounds on the normalized sum and the limiting normal distribution.
-This specificity answers Chebyshev's earlier demand for sharper probabilistic results although it should have simplified the proof through some clever but much deeper refinements of Poisson's arguments \cite{fischer2011history}.
-
-Lyapunov’s goals extended beyond mere generalization; he aimed to illuminate the theorem’s internal structure and make it more accessible.
-Avoiding the complexities of characteristic functions and complex analysis, he focused on moment-based criteria, relying on real-valued tools like Taylor expansions, bounding techniques, and Lyapunov's inequality.
-Thus the balance here is theoretical rigor and practical applicability, two characteristics that still resonate well with modern efforts at formal verification as that undertaken in this thesis.
-
-In contemporary times, further refinements of the CLT were provided by William Feller bringing it to discrete random variables \cite{feller1945}, and all subsequent developments that are probabilistic have been geared towards this common extension to stable distributions and higher dimensional data.
-These extensions demonstrate the theorem’s adaptability to various contexts, from stochastic processes to machine learning and data analysis.
-
-Lyapunov’s version of the CLT represents a key milestone in this journey.
-It generalizes the theorem for independent random variables with finite moments so that it becomes an observed but pliant frame of reference for convergence.
-It is this balancing between abstractness and concreteness that is a characteristic of the whole development of probability during the modern period.
-Contemporaries of Lyapunov such as von Mises, Lévy, Cramér, Khinchin, and Kolmogorov likewise pursued abstract relations in probability theory while keeping an eye on practical relevance.
-This tension is paradigmatically present in Lyapunov's work, which illustrates the way in which results which are purely formal mathematics may nonetheless satisfy external criteria of applicability and utility.
-For more information about the history and the proofs of the Central Limit Theorem, see \cite{fischer2011history, adams2009life}.
-
-This thesis is based on the generalization of CLT as expressed in Lyapunov's version.
-Because of its focus on real-valued methods and moment-based criteria, it avoids any complex tools like characteristic functions or complex analysis, hence being particularly suitable for formalization in HOL4.
-In this respect, Lyapunov's approach not only advances the mathematical generalization but also provides a practical framework to meet the rigor and accessibility requirements of automated reasoning systems by directly addressing convergence under relaxed conditions.
+In this thesis, we focus on the Lyapunov form of the CLT, as presented in Chung’s textbook \cite{Chung:2001}.
+It assumes independence, but not identical distribution, and uses the Lindeberg replacement trick to incrementally replace variables with normal ones, bounding the error using Taylor's theorem.
+This analytic method avoids characteristic functions and relies only on real analysis and moment estimates.
+As a result, it is well-suited to formalisation in systems like HOL4, which are strong in real analysis but currently lack mature libraries for complex integration.
 
@@
 
-Formalization of mathematical theorems as proofs has been a singular research focus in modern computational mathematics.
-The systems such as Isabelle/HOL \citep{isabelle}, Coq \citep{bertot2013interactive}, and HOL4 \citep{slind2008brief} have been heavily relied on over the years to encode and prove basic properties but offer several features and challenges.
+Formal proofs of the CLT have previously been attempted in several proof assistants.
+The most notable example is the Isabelle/HOL formalisation by \cite{serafin2015formally}, which proves the CLT under the assumption of independent identically distributed (i.i.d.) random variables.
+That work follows the classical approach using characteristic functions and the Lévy Continuity Theorem.
 
-One of the most significant achievements in this area was the formalization of CLT within the Isabelle/HOL proof assistant.
-This was made possible mainly by the rich library related to complex analysis in Isabelle/HOL, which provided the means to make use of characteristic functions within the proof.
-Characteristic functions are Fourier transforms of probability distributions that make convergence easier to analyze by drawing on their special mathematical properties.
-Central to the approach are two key steps: first, pointwise convergence of the characteristic function for the normalized sum of independent random variables to that of the standard normal distribution should be demonstrated; and second, by the application of the L\'evy Continuity Theorem, one obtains the desired convergence in distribution.
-These two steps are actually the core of the proof that eventually leads to the verification of CLT \cite{billingsley2017probability}.
+While the Isabelle/HOL technique is elegant and mathematically sound, it relies on involved analysis and Fourier transforms, which are well-supported in Isabelle but not yet fully formalised in HOL4.
+HOL4 is lacking key theorems for complex-valued functions at present, so the characteristic function approach is unworkable here.
 
-The characteristic function approach is beautifully complemented by the infrastructure provided by Isabelle/HOL, which supports careful manipulation of complex-valued functions, differentiation, and limits.
-This thereby enables Isabelle/HOL to achieve the rigorous performance that is demanded by Fourier analysis and probability theory.
-Among the striking personalities behind this effort is Luke Serafin, who has produced a giant leap forward in the formalization of the CLT \cite{serafin2015formally}.
-Serafin's systematic encoding of the characteristic function method has ensured that it was entirely generic and modular.
-His contributions involved the formalization of key intermediate results such as the for sums of independent random variables and the fact that components of his work are reusable, enabling extension into applications that relate to probabilistic theorems.
+Moreover, the i.i.d.
+assumption limits the scope of the Isabelle formalisation.
+In contrast, this thesis formalises a strictly more general version of the CLT—Lyapunov’s form—which requires only independence and finiteness of variances and third absolute moments.
+The approach avoids characteristic functions entirely and instead uses the Lindeberg replacement method with Taylor expansion and asymptotic error bounds.
 
-The strength of Isabelle/HOL is to borrow from its excellent complex analysis library.
-The characteristic function-based proofs are formalized smoothly, and the modular and extensible framework allows researchers to always build on previous work, gradually increasing the scope of formalized mathematics.
-The CLT formalization within Isabelle/HOL is exemplary in showing how even the most intricate probabilistic results can be encoded rigorously with the help of proof assistants.
-But this also points to some of the challenges with this approach: it is very much based on Fourier transforms and complex-valued methods, an infrastructure which may not always be as easily available when moving to other proof systems.
-Also, characteristic functions are powerful but often complicate proofs in situations where simpler real-valued methods, such as those based on moments, would suffice.
+This makes the result both broader in applicability and better aligned with the current capabilities of HOL4’s real and measure-theoretic libraries.
+The infrastructure developed in this work—including handling of expectations, variances, random variable sequences, and Taylor bounds—may serve as a foundation for generalising beyond Lyapunov’s condition in future work.
 
-The formalization of CLT in Isabelle/HOL serves as an inspiration and comparison for the research presented in this thesis.
-The Example of Isabelle/HOL shows that the CLT can be formalized with the help of characteristic functions but also brings into relief the limitations of such a strategy-if proof assistants like HOL4 lack libraries comparable for complex analysis with it.
-This thesis takes a different site following the approach with the theorem of Lyapunov, which involves real-valued moment conditions rather than characteristic functions.
-Following this path, the work in HOL4 demonstrates how easier and less convoluted tools enable formulating the CLT while considering the inherent challenges of formal verification.
+To the best of our knowledge, this is the first mechanised proof of the Central Limit Theorem in HOL4 that goes beyond the i.i.d.
+case and formalises the full structure of the Lindeberg–Lyapunov strategy.

### 2025-05-19 clt.tex
init commit
Modified Ch. Introduction
Almost done report

--- 2024-11-30 background.tex
@@
-\chapter{Background and Related Work}
-\label{background}
+%%% -*- Mode: LaTeX; -*-
+\chapter{Central Limit Theorem}
+\label{ch:clt}
 
-Background 123
+In this chapter, we show the formalisation of Lyapunov's form of Central Limit Theorem (CLT) in HOL4.
+We closely follow the textbook proof by Chung~\cite{Chung:2001}, employing primarily measure-theoretic techniques, Lindeberg's replacement principle, and Taylor expansions in explicit form.
+The general idea is to prove that the distribution of the normalised sum of independent random variables converges to a standard normal distribution under the Lyapunov condition.
 
-% \begin{figure*}
-% \centering
-% \includegraphics[width=0.98\textwidth]{Chapter1/Images/Fig1.pdf}
-% \caption{caption}
-% \label{figCh1:figure}
-% \end{figure*}
+The proof structure naturally divides into several steps, as shown in the figure below.
+First, we define the setup formally, e.g., probability spaces, independence, and moment conditions.
+Second, we introduce an auxiliary sequence of Gaussian random variables with the same variance structure as the original sequence but which are easier to manage in terms of distributional properties.
+The key strategy, referred to as the Lindeberg replacement trick, systematically replaces the original variables with the auxiliary variables and bounds the resulting error.
+Taylor's theorem gives us explicit error terms, while Lyapunov's inequality and Big-O estimates bound the third moment terms.
+Lastly, this sequence of approximations immediately gives us the desired convergence in distribution.
 
-\renewcommand\thefigure{\thechapter .\arabic{figure}}
+\begin{figure}[h!]
+  \caption{Proof Structure}
+  \includegraphics[width=7cm]{./img/method.png}
+  \centering
+\end{figure}
+
+\section {Informal Proof}
+
+To prove the Central Limit Theorem for a single sequence of independent (but not necessarily identically distributed) random variables $\{X_j\}_{1 \leq j \leq n}$, we apply the Lindeberg replacement method.
+Each $X_j$ is assumed to satisfy:
+
+\[
+\mathbb{E}[X_j] = 0, \quad \mathbb{E}[X_j^2] = \sigma_j^2 < \infty, \quad \mathbb{E}[|X_j|^3] = \gamma_j < \infty.
+\]
+
+Define the cumulative sums:
+
+\[
+S_n = \sum_{j=1}^n X_j, \quad s_n^2 = \sum_{j=1}^n \sigma_j^2, \quad \Gamma_n = \sum_{j=1}^n \gamma_j.
+\]
+
+We aim to prove that if:
+
+\[
+\frac{\Gamma_n}{s_n^3} \to 0,
+\]
+
+then the normalized sum $S_n / s_n$ converges in distribution to the standard normal.
+Notationally, we write:
+
+\[
+\frac{S_n}{s_n} \xrightarrow{d} \mathcal{N}(0,1).
+\]
+
+\subsection{Replacement Strategy}
+
+The idea is to approximate the sum $X_1 + \cdots + X_n$ by gradually replacing each $X_j$ with a corresponding independent normal variable $Y_j \sim \mathcal{N}(0, \sigma_j^2)$.
+Let all the $X_j$'s and $Y_j$'s be totally independent.
+
+Define for each $j$:
+
+\[
+Z_j = Y_1 + \cdots + Y_{j-1} + X_j + \cdots + X_n,
+\quad
+Z_{j+1} = Y_1 + \cdots + Y_j + X_{j+1} + \cdots + X_n.
+\]
+
+So the difference $Z_j - Z_{j+1} = X_j - Y_j$ replaces one variable at a time.
+
+Then:
+
+\[
+\mathbb{E}\left[ f\left(\frac{S_n}{s_n} \right) \right] - \mathbb{E}\left[ f\left(\frac{Y_1 + \cdots + Y_n}{s_n} \right) \right]
+= \sum_{j=1}^n \left(
+\mathbb{E}\left[f\left( \frac{X_j + Z_j}{s_n} \right)\right]
+- \mathbb{E}\left[f\left( \frac{Y_j + Z_j}{s_n} \right)\right]
+\right)
+\]
+
+\subsection{Taylor Expansion Bound}
+
+Using Taylor's theorem and the fact that $f \in C^3_b$ (bounded continuous functions with bounded derivatives up to order 3), we bound the difference:
+
+\[
+\left| \mathbb{E}\left[f\left( \frac{X_j + Z_j}{s_n} \right)\right] -
+       \mathbb{E}\left[f\left( \frac{Y_j + Z_j}{s_n} \right)\right]
+\right| \leq
+\frac{M}{6 s_n^3} \left( \mathbb{E}[|X_j|^3] + \mathbb{E}[|Y_j|^3] \right),
+\]
+
+for  $M = \sup_{x \in \mathbb{R}} \left| f^{(3)}(x) \right|$.
+
+This constant $M$ reflects the maximum curvature of $f$, and controls the Taylor remainder in the third-order expansion.
+
+Summing over $j$, we get:
+
+\[
+\left| \mathbb{E}\left[f\left( \frac{S_n}{s_n} \right) \right] -
+       \mathbb{E}[f(N)]
+\right| \leq
+\frac{M}{6 s_n^3} \sum_{j=1}^n \left( \gamma_j + c \sigma_j^3 \right),
+\]
+
+where $N \sim \mathcal{N}(0, 1)$ and $c$ is a constant depending on the third moment of standard normal.
+
+By Lyapunov’s inequality, $\sigma_j^3 \leq \gamma_j$, so the right-hand side is $O(\Gamma_n / s_n^3)$, which tends to zero. This proves $S_n / s_n \xrightarrow{d} N(0,1)$.
+
+\section {Construction of Auxiliary Sequence}
+
+To carry out the Lindeberg replacement method, we need to introduce an auxiliary sequence $\{Y_j\}$ of Gaussian random variables that are independent and have the same variance structure as the original $\{X_j\}$.
+The goal is to build these $Y_j$ over a new probability space that allows us to reason about both $X_j$ and $Y_j$ simultaneously.
+
+\textbf{Theorem 1.}
+Let $D$ be a function that gives positive variances for $n > 0$ dimensions — that is, for each index $i < n$, $D(i) > 0$.
+
+Then, there exists a new probability space $p'$, and a sequence of random variables $Y_0, Y_1, \dots, Y_{n-1}$ defined on that space such that:
+\begin{itemize}
+\item Each $Y_i$ is a normal random variable with mean 0 and variance $D(i)$
+\item The random variables $Y_0, \dots, Y_{n-1}$ are independent from each other
+\end{itemize}
+
+Or formally,
+
+\begin{hol}
+  \begin{alltt}
+    Theorem existence\_of\_indep\_vars :
+    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)(p :\(\alpha\) m\_space) N (D :num \(\rightarrow\) real) n.
+    prob\_space p \(\land\) 0 < n \(\land\) ext\_normal\_rv N p 0 1 \(\land\)
+    (\(\forall\)i. i < n \(\Rightarrow\) 0 < (D i)) \(\Rightarrow\)
+    \(\exists\)(p' :\(\alpha\) list m\_space) Y.
+    prob\_space p' \(\land\)
+    (\(\forall\)(i :num). i < n \(\Rightarrow\) ext\_normal\_rv (Y i) p' 0 (D i)) \(\land\)
+    indep\_vars p' Y (\(\lambda\)i. Borel) (count n)
+  \end{alltt}
+\end{hol}
+
+The classic idea, as presented in Fremlin’s Measure Theory~\cite{fremlinmeasure}, is to construct the product space $\Omega' = \Omega \times \mathbb{R}^n$, where $\Omega$ is the original probability space carrying the variables $X_1, \dots, X_n$, and $\mathbb{R}^n$ is equipped with a standard Gaussian product measure.
+Each component of this product then hosts one of the auxiliary Gaussian variables.
+This idea guarantees that we can preserve the distribution of $X_j$ while augmenting the space with new independent $Y_j \sim \mathcal{N}(0, \sigma_j^2)$.
+
+In our formalisation, we adapt this idea to HOL4 by explicitly constructing two independent probability spaces:
+\begin{itemize}
+    \item $p_1$, hosting the original sequence $\{X_i\}_{i < n}$,
+    \item $p_2$, hosting the auxiliary sequence $\{Y_i\}_{i < n}$, assumed to be independent and Gaussian.
+\end{itemize}
+
+We then take their product measure $p = p_1 \times p_2$, which remains a valid probability space thanks to the existing \holtxt{existence\_of\_prod\_prob\_space} theorem.
+
+\begin{hol}
+  \begin{alltt}
+    Theorem existence\_of\_prod\_prob\_space :
+    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p1 p2.
+    prob\_space p1 \(\land\) prob\_space p2 \(\Rightarrow\)
+    \(\exists\)p. p = p1 \(\times\) p2 \(\land\) prob\_space p \(\land\)
+    (\(\forall\)e1 e2.
+    e1 \(\in\) events p1 \(\land\) e2 \(\in\) events p2 \(\Rightarrow\)
+    e1 \(\times\) e2 \(\in\) events p \(\land\)
+    prob p (e1 \(\times\) e2) = prob p1 e1 * prob p2 e2)
+  \end{alltt}
+\end{hol}
+
+Within this product space, we define:
+\begin{itemize}
+    \item $X'_i = X_i \circ \texttt{FST}$, and
+    \item $Y'_i = Y_i \circ \texttt{SND}$,
+\end{itemize}
+
+as random variables on $p$, so that their marginal behaviours match the originals. Finally, we interleave these into a single indexed family:
+
+\[
+    Z_i(x) =
+    \begin{cases}
+      X'_i(x), & \text{if } i < n \\
+      Y'_{i-n}(x), & \text{if } n \leq i < 2n
+    \end{cases}
+\]
+
+We formally prove that the sequence $\{Z_i\}_{i < 2n}$ is a family of independent real-valued random variables over the product probability space $p$. This construction ensures that:
+
+\begin{itemize}
+\item The first $n$ components $\{Z_i\}_{i < n}$ correspond exactly to $X_i$,
+\item The remaining $\{Z_i\}_{n \leq i < 2n}$ are the auxiliary $Y_j$, with identical variance structure,
+\item All variables are mutually independent.
+\end{itemize}
+
+This leads to the following formal result in HOL4, which guarantees the existence of such a product probability space and the sequence $Z_i$ combining both original and auxiliary components.
+
+\begin{hol}
+  \begin{alltt}
+    Theorem construct\_auxiliary\_seq :
+    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p1 (p2 :'a list m\_space) X Y (n \:num).
+    prob\_space p1 \(\land\) prob\_space p2 \(\land\) 0 < n \(\land\)
+    (\(\forall\)i. i < n \(\Rightarrow\) real\_random\_variable (X i) p1) \(\land\)
+    (\(\forall\)i. i < n \(\Rightarrow\) real\_random\_variable (Y i) p2) \(\land\)
+    indep\_vars p1 X (\(\lambda\)i. Borel) (count n) \(\land\)
+    indep\_vars p2 Y (\(\lambda\)i. Borel) (count n) \(\Rightarrow\)
+    \(\exists\)p X' Y' Z.
+    (p = p1 CROSS p2) \(\land\)
+    (X' = \(\lambda\)i. X i \(\circ\) FST) \(\land\)
+    (Y' = \(\lambda\)i. Y i \(\circ\) SND) \(\land\)
+    prob\_space p \(\land\)
+    (\(\forall\)i. i < n \(\Rightarrow\) real\_random\_variable (X' i) p) \(\land\)
+    (\(\forall\)i. i < n \(\Rightarrow\) real\_random\_variable (Y' i) p) \(\land\)
+    (Z = \(\lambda\)i x. if i < n then X' i x else Y' (i - n) x) \(\land\)
+    indep\_vars p Z (\(\lambda\)(i \:num). Borel) (count (2 \* n))
+  \end{alltt}
+\end{hol}
+
+\section {Taylor Expansion Bounds}
+
+After constructing the auxiliary sequence $\{Y_j\}$ of independent normal variables matching the variances of $\{X_j\}$, the next step is to control the difference in expectations:
+
+\[
+\mathbb{E}[f(S_n / s_n)] - \mathbb{E}[f(G_n / s_n)],
+\]
+
+where $S_n = \sum_{j=0}^{n-1} X_j$, and $G_n = \sum_{j=0}^{n-1} Y_j$. We do this by replacing one term at a time, writing the difference as a telescoping sum:
+
+\[
+\sum_{j=0}^{n-1} \left(
+\mathbb{E}\left[f\left(\frac{X_j + Z_j}{s_n}\right)\right]
+-
+\mathbb{E}\left[f\left(\frac{Y_j + Z_j}{s_n}\right)\right]
+\right),
+\]
+
+where $Z_j$ is the partial sum involving all terms except $X_j$ or $Y_j$.
+To estimate each difference, we use Taylor's theorem with remainder.
+
+\subsection{Taylor Expansion Theorems}
+
+To formally bound the error, we need two main ingredients:
+
+\textbf{Theorem 1.}(\emph{Taylor’s Theorem})
+
+Let $f : \mathbb{R} \to \mathbb{R}$ be $n$-times differentiable on an interval $[a, x]$. Then there exists some $t \in (a, x)$ such that:
+
+\[
+f(x) =
+\sum_{m=0}^{n-1} \frac{f^{(m)}(a)}{m!} (x - a)^m +
+\frac{f^{(n)}(t)}{n!} (x - a)^n.
+\]
+
+In our setting, we focus on the case $n = 3$, so:
+
+\[
+f(a + h) =
+f(a) + f'(a)h + \frac{1}{2}f''(a)h^2 + \frac{1}{6}f^{(3)}(t)h^3.
+\]
+
+This is captured in HOL4 as:
+
+\begin{hol}
+  \begin{alltt}
+    Theorem TAYLOR\_THEOREM :
+    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)f a x n.
+    a < x \(\land\) 0 < n \(\land\)
+    (\(\forall\)m t. m < n \(\land\) a \(\le\) t \(\land\) t \(\le\) x \(\Rightarrow\)
+    higher\_differentiable (SUC m) f t) \(\Rightarrow\)
+    \(\exists\)t. a < t \(\land\) t < x \(\land\)
+    f x =
+    sum (0,n) (\(\lambda\)m. diff m f a / \&FACT m * (x \({-}\) a) pow m) +
+    diff n f t / \&FACT n * (x \({-}\) a) pow n
+  \end{alltt}
+\end{hol}
+
+\textbf{Theorem 1.}(\emph{Taylor Remainder Bound})
+
+The Taylor remainder theorem describes the difference between a function and its Taylor polynomial.
+Suppose $f : \mathbb{R} \to \mathbb{R}$ is $n$-times differentiable on an interval containing $a$ and $x = a + h$.
+The function value $f(x)$ can be written as:
+
+\[
+f(x) = \sum_{m=0}^{n-1} \frac{f^{(m)}(a)}{m!} h^m + R_n(h),
+\]
+
+where $R_n(h)$ is the remainder term. Taylor's theorem guarantees that there exists some $t$ between $a$ and $x$ such that:
+
+\[
+R_n(h) = \frac{f^{(n)}(t)}{n!} h^n.
+\]
+
+Or, formally:
+\begin{hol}
+  \begin{alltt}
+    Theorem TAYLOR\_REMAINDER :
+    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)n x f. \(\exists\)M t.
+    abs (Normal (diff n f t)) \(\le\) M \(\Rightarrow\)
+    abs (Normal (diff n f t / \&FACT n) * Normal x pow n) \(\le\)
+    M / Normal (\&FACT n) * abs (Normal x) pow n
+  \end{alltt}
+\end{hol}
+
+Assume $f \in C^3_b$, that is, $f$ has bounded third derivative over $\mathbb{R}$, and let:
+
+\[
+M = \sup_{x \in \mathbb{R}} |f^{(3)}(x)|.
+\]
+
+Then the remainder term satisfies:
+
+\[
+\left| f(a + h) - f(a) - f'(a)h - \frac{1}{2}f''(a)h^2 \right|
+\leq \frac{M}{6} |h|^3.
+\]
+
+This is formalised as:
+\begin{hol}
+  \begin{alltt}
+    Theorem TAYLOR\_THIRD\_ORDER\_BOUND :
+    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)f a h M.
+    f \(\in\) CnR 3 \(\land\)
+    M = sup (IMAGE (\(\lambda\)t. abs (Normal (diff 3 f t))) \(\mathbb{U}\)(:\(\real\))) \(\Rightarrow\)
+    abs (Normal (f (a + h) \({-}\) f a \({-}\) diff 1 f a * h \({-}\) 1 / 2 * diff 2 f a * h pow 2)) \(\le\)
+    M / 6 * abs (Normal h) pow 3
+  \end{alltt}
+\end{hol}
+
+This bound will be applied to the difference $f(X_j + Z_j) - f(Y_j + Z_j)$, treating $X_j - Y_j$ as a small perturbation.
+
+\section{Formal Lindeberg Replacement Lemma}
+
+After constructing the auxiliary sequence $\{Y_j\}$ and bounding the Taylor expansion error for each replacement, we now formalise the full Lindeberg replacement argument.
+This step accumulates the errors introduced when replacing each original variable $X_j$ by the corresponding auxiliary normal variable $Y_j$, and shows that the total error vanishes under Lyapunov's condition.
+
+\subsection{Error Decomposition via Telescoping Sum}
+
+Let $S_n = \sum_{j=0}^{n-1} X_j$ and $G_n = \sum_{j=0}^{n-1} Y_j$, and fix a function $f : \mathbb{R} \to \mathbb{R}$ in the class $C^3_b$, i.e., $f$ is three times differentiable with all derivatives bounded.
+Our goal is to estimate the difference:
+
+\[
+\left| \mathbb{E}\left[f\left(\frac{S_n}{s_n}\right)\right]
+     - \mathbb{E}\left[f\left(\frac{G_n}{s_n}\right)\right] \right|.
+\]
+
+Following the informal argument, we express this difference as a telescoping sum:
+
+\[
+\sum_{j=0}^{n-1} \left(
+\mathbb{E}\left[f\left(\frac{X_j + Z_j}{s_n}\right)\right]
+-
+\mathbb{E}\left[f\left(\frac{Y_j + Z_j}{s_n}\right)\right]
+\right),
+\]
+
+where each $Z_j$ represents the sum of the remaining variables (excluding $X_j$ and $Y_j$), and is assumed independent from both.
+
+Each term in this sum is then bounded using Taylor's theorem as shown in the previous section.
+
+\textbf{Proposition X.} (\emph{Measurability and Integrability of Partial Sums})
+
+Let $X_0, \dots, X_{n-1}$ and $Y_0, \dots, Y_{n-1}$ be integrable real-valued random variables over a probability space $p$. Define, for each $j < n$, a partial sum function:
+
+\[
+Z_j(x) = \sum_{i < j} Y_i(x) + \sum_{j \leq i < n} X_i(x)
+\]
+
+This lemma ensures that each such function $Z_j$ remains a real random variable and integrable:
+
+\begin{hol}
+\begin{alltt}
+Theorem clt\_partial\_sum\_lemma :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p X Y Z f n.
+prob\_space p \(\land\)
+(\(\forall\)i. i < n \(\Rightarrow\)
+  real\_random\_variable (X i) p \(\land\)
+  real\_random\_variable (Y i) p \(\land\)
+  integrable p (X i) \(\land\)
+  integrable p (Y i)) \(\Rightarrow\)
+(\(\forall\)j. j < n \(\Rightarrow\)
+  \(\forall\)Z. Z =
+    (\(\lambda\)j x.
+      if x \(\in\) p\_space p then
+        \(\sum\) (\(\lambda\)i. Y i x) (count j) +
+        \(\sum\) (\(\lambda\)i. X i x) (count n \(\setminus\) count j)
+      else 0) \(\Rightarrow\)
+    real\_random\_variable (Z j) p \(\land\) integrable p (Z j))
+\end{alltt}
+\end{hol}
+
+This result guarantees that the hybrid sum $Z_j$, required for the Lindeberg replacement trick, is well-defined and suitable for expectation bounds.
+
+\textbf{Proposition X.} (\emph{Telescoping Sum over Finite Range})
+The following identity simplifies the sum of differences in a sequence:
+
+\[
+\sum_{j = 0}^{n-1} \left( f(j) - f(j + 1) \right) = f(0) - f(n),
+\]
+
+provided $f(j)$ is finite for all $j$. This is formalised in HOL4 as:
+
+\begin{hol}
+\begin{alltt}
+Theorem SUM\_SUB\_GEN :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)f n.
+(\(\forall\)x. f x \(\ne\) -\(\infty\) \(\land\) f x \(\ne\) +\(\infty\)) \(\Rightarrow\)
+\(\sum\) (\(\lambda\)j. f j \({-}\) f (j + 1)) (count n) = f 0 \({-}\) f n
+\end{alltt}
+\end{hol}
+
+This identity is especially useful in simplifying the error terms expressed as a telescoping sum when estimating differences between expectations.
+
+\textbf{Theorem 1.}(\emph{Replacement Lemma Statement})
+
+\begin{hol}
+\begin{alltt}
+Theorem clt\_Lindeberg\_replacement\_trick\_bounded[local] :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p X Y f s n.
+prob\_space p \(\land\)
+f \(\in\) CnR 3 \(\land\)
+(\(\forall\)i. i < n \(\Rightarrow\)
+  real\_random\_variable (X i) p \(\land\)
+  real\_random\_variable (Y i) p \(\land\)
+  integrable p (X i) \(\land\)
+  integrable p (Y i)) \(\land\)
+0 < s n \(\land\) s n \(\ne\) +\(\infty\) \(\land\) s n \(\ne\) -\(\infty\) \(\land\)
+(\(\forall\)j. j < n \(\Rightarrow\)
+  (\(\forall\)Z. Z = (\(\lambda\)j x. if x \(\in\) p\_space p then
+                      \(\sum\) (\(\lambda\)i. Y i x) (count j) +
+                      \(\sum\) (\(\lambda\)i. X i x) (count n DIFF count1 j)
+                   else 0) \(\Rightarrow\)
+    expectation p (Normal \(\circ\) f \(\circ\) real \(\circ\)
+                   (\(\lambda\)x. \(\sum\) (\(\lambda\)i. X i x) (count n) / s n)) \({-}\)
+    expectation p (Normal \(\circ\) f \(\circ\) real \(\circ\)
+                   (\(\lambda\)x. \(\sum\) (\(\lambda\)i. Y i x) (count n) / s n)) =
+
+    \(\sum\) (\(\lambda\)j.
+      expectation p (Normal \(\circ\) f \(\circ\) real \(\circ\)
+                     (\(\lambda\)x. (X j x + Z j x) / s n)) \({-}\)
+      expectation p (Normal \(\circ\) f \(\circ\) real \(\circ\)
+                     (\(\lambda\)x. (Y j x + Z j x) / s n))) (count n)))
+\end{alltt}
+\end{hol}
+
+This theorem quantifies how the accumulated replacement error scales with the third moments of $X_j$ and $Y_j$, and the normalization constant $s_n$. The assumptions guarantee:
+
+\begin{itemize}
+\item \textbf{Independence}: Each pair $(X_j, Z_j)$ and $(Y_j, Z_j)$ are independent, which is crucial for separating expectations during Taylor expansion.
+\item \textbf{Matching Second Moments}: Implicitly required for the expectation of quadratic terms to cancel out.
+\item \textbf{Bounded Third Derivative}: Controlled via the supremum constant $M$ defined from the third derivative of $f$.
+\end{itemize}
+
+This theorem precisely formalises Equation (15) in the informal Chung proof.
+The right-hand side provides an upper bound on the total error in terms of third moments.
+Once we show that this upper bound tends to zero (which we do using Lyapunov’s condition in the next section), we conclude that the distribution of $S_n / s_n$ is close to that of $G_n / s_n$, which converges to standard normal.
+
+Thus, this lemma acts as the bridge between the approximation via normal variables and the original non-Gaussian sequence.
+
+\begin{figure}[h]
+  \caption{Illustration of the Lindeberg replacement method}
+  \includegraphics[width=12cm]{./img/replacement.png}
+  \centering
+\end{figure}
+
+\section{Global Taylor Error Bound}
+
+After expressing the Lindeberg replacement as a telescoping sum and bounding each replacement using Taylor’s theorem, we now combine those bounds into a single inequality.
+This step precisely captures the cumulative approximation error when replacing each $X_j$ by $Y_j$.
+
+\subsection{Formal Statement of the Error Bound}
+
+We define $f \in C^3_b$ as a test function with bounded third derivative. Let $X_j$, $Y_j$, and $Z_j$ be sequences of real random variables such that for each $j < n$:
+\begin{itemize}
+    \item $Z_j$ is independent of both $X_j$ and $Y_j$,
+    \item $X_j$ and $Y_j$ are centered, share the same variance,
+    \item third moments $\mathbb{E}[|X_j|^3]$ and $\mathbb{E}[|Y_j|^3]$ are finite.
+\end{itemize}
+
+Then the total error is bounded by:
+
+\[
+\left| \sum_{j=0}^{n-1}
+\mathbb{E}\left[f\left(\frac{X_j + Z_j}{s_n} \right)\right] -
+\mathbb{E}\left[f\left(\frac{Y_j + Z_j}{s_n} \right)\right]
+\right| \leq \frac{M}{6 s_n^3}
+\sum_{j=0}^{n-1} \left( \mathbb{E}[|X_j|^3] + \mathbb{E}[|Y_j|^3] \right)
+\]
+
+where $M = \sup_{x \in \mathbb{R}} |f^{(3)}(x)|$, and $s_n^2 = \sum_{j=0}^{n-1} \mathrm{Var}(X_j)$.
+
+This is formalised in HOL4 as:
+
+\begin{hol}
+\begin{alltt}
+Theorem clt\_lindeberg\_taylor\_error\_bound :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)r X Y Z f M s n.
+prob\_space r \(\land\)
+(\(\forall\)(j :num). j < n \(\Rightarrow\)
+  real\_random\_variable (X j) r \(\land\)
+  real\_random\_variable (Y j) r \(\land\)
+  real\_random\_variable (Z j) r \(\land\)
+  integrable r (\(\lambda\)x. X j x) \(\land\)
+  integrable r (\(\lambda\)x. Y j x) \(\land\)
+  integrable r (\(\lambda\)x. Z j x) \(\land\)
+  integrable r (\(\lambda\)x. (abs (X j x)) pow 3) \(\land\)
+  integrable r (\(\lambda\)x. (abs (Y j x)) pow 3) \(\land\)
+  indep\_vars r (X j) (Z j) Borel Borel \(\land\)
+  indep\_vars r (Y j) (Z j) Borel Borel) \(\land\)
+f \(\in\) CnR 3 \(\land\)
+M = sup (IMAGE (\(\lambda\)t. abs (Normal (diff 3 f t))) UNIV) \(\land\)
+0 < s n \(\land\) s n \(\ne\) +\(\infty\) \(\land\) s n \(\ne\) -\(\infty\) \(\Rightarrow\)
+abs (\(\sum\) (\(\lambda\)j.
+  expectation r (Normal \(\circ\) f \(\circ\) real \(\circ\) (\(\lambda\)x. (X j x + Z j x) / s n)) \({-}\)
+  expectation r (Normal \(\circ\) f \(\circ\) real \(\circ\) (\(\lambda\)x. (Y j x + Z j x) / s n))) (count n)) \(\le\)
+M / (6 * (s n) pow 3) *
+\(\sum\) (\(\lambda\)j.
+  expectation r (\(\lambda\)x. (abs (X j x)) pow 3 + (abs (Y j x)) pow 3)) (count n)
+\end{alltt}
+\end{hol}
+
+This theorem completes the Lindeberg replacement method by quantifying the total replacement error via third moment control.
+As the Lyapunov condition ensures the RHS tends to zero, it follows that the difference in distributions of the original and auxiliary sums vanishes.
+
+This theorem serves as the final analytic step before applying convergence theorems to conclude:
+
+\[
+\frac{S_n}{s_n} \xrightarrow{d} \mathcal{N}(0,1).
+\]
+
+\section{Lyapunov Condition and Third Moment Bound}
+
+This section connects the Lyapunov condition with third moment estimates used in bounding the global Taylor approximation error.
+This connection ensures that the total error in the Lindeberg method vanishes asymptotically.
+
+\textbf{Theorem 1.} (\emph{Lyapunov Inequality})
+Let $(\Omega, \mathcal{F}, \mathbb{P})$ be a probability space, and let $X$ be a random variable such that both $\mathbb{E}[|X|^r]$ and $\mathbb{E}[|X|^{r'}]$ are finite, for some $0 < r < r' < \infty$.
+Then:
+\[
+\left( \mathbb{E}[|X|^r] \right)^{1/r} \le \left( \mathbb{E}[|X|^{r'}] \right)^{1/r'}.
+\]
+
+In short, the $L^r$ norm of $X$ is bounded by its $L^{r'}$ norm.
+This allows us to control lower-order moments using higher-order ones — a key idea in proving convergence in distribution under Lyapunov conditions.
+
+\subsection{Lyapunov Inequality for Integrals}
+
+We begin with a basic inequality that bounds the $L^1$ norm of a function by its $L^p$ seminorm, scaled by the measure of the entire space.
+This is useful when estimating the first moment of a random variable in $L^p$ for $p > 1$.
+
+\begin{hol}
+\begin{alltt}
+Theorem liapounov\_ineq\_lemma :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)m u p.
+measure\_space m \(\land\) measure m (m\_space m) < +\(\infty\) \(\land\)
+1 < p \(\land\) p < +\(\infty\) \(\land\)
+u \(\in\) lp\_space p m \(\Rightarrow\)
+\(\int^+\) m (\(\lambda\)x. abs (u x)) \(\le\)
+seminorm p m u * measure m (m\_space m) powr (1 \({-}\) inv p)
+\end{alltt}
+\end{hol}
+
+\subsection{Comparing $L^p$ and $L^{p'}$ Seminorms}
+
+The next two theorems compare $L^p$ seminorms when $p < p'$, assuming the measure space is finite.
+
+\begin{hol}
+\begin{alltt}
+Theorem liapounov\_ineq :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)m u r r'.
+measure\_space m \(\land\)
+u \(\in\) lp\_space r m \(\land\) u \(\in\) lp\_space r' m \(\land\)
+measure m (m\_space m) < +\(\infty\) \(\land\)
+0 < r \(\land\) r < r' \(\land\) r' < +\(\infty\) \(\Rightarrow\)
+seminorm r m u \(\le\)
+seminorm r' m u * measure m (m\_space m) powr (inv r \({-}\) inv r')
+\end{alltt}
+\end{hol}
+
+In a probability space, where the total measure is 1, the inequality simplifies:
+
+\begin{hol}
+\begin{alltt}
+Theorem liapounov\_ineq\_rv :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p u r r'.
+prob\_space p \(\land\)
+u \(\in\) lp\_space r p \(\land\) u \(\in\) lp\_space r' p \(\land\)
+0 < r \(\land\) r < r' \(\land\) r' < +\(\infty\) \(\Rightarrow\)
+seminorm r p u \(\le\) seminorm r' p u
+\end{alltt}
+\end{hol}
+
+These inequalities help relate different moment conditions and are crucial when applying Lyapunov’s condition for the CLT.
+
+\subsection{Variance Controlled by Third Moment}
+
+We now state a bound showing that the third absolute moment dominates the cube of the standard deviation.
+
+\begin{hol}
+\begin{alltt}
+Theorem clt\_liapounov\_upper\_bound :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p X Y.
+prob\_space p \(\land\)
+real\_random\_variable X p \(\land\)
+expectation p (\(\lambda\)x. (abs (X x)) pow 3) < +\(\infty\) \(\Rightarrow\)
+Normal (sqrt (real (variance p X))) pow 3 \(\le\)
+expectation p (\(\lambda\)x. (abs (X x)) pow 3)
+\end{alltt}
+\end{hol}
+
+In the CLT, we normalize the sum by $s_n^3 = \left( \sum_j \mathrm{Var}(X_j) \right)^{3/2}$. This bound ensures the denominator does not vanish too quickly, keeping the Taylor error under control.
+
+\subsection{Exact Third Moment of a Gaussian Variable}
+
+For the auxiliary sequence $Y_j \sim \mathcal{N}(0, \sigma^2)$, the third absolute moment is:
+
+\[
+\mathbb{E}[|X|^3] = \sqrt{\frac{8}{\pi}} \cdot \sigma^3,
+\quad \text{for } X \sim \mathcal{N}(0, \sigma^2).
+\]
+
+Formally,
+\begin{hol}
+  \begin{alltt}
+    Theorem ext\_normal\_rv\_abs\_third\_moment :
+    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p X sig. prob\_space p \(\land\) 0 < sig \(\land\)
+    ext\_normal\_rv X p 0 sig \(\Rightarrow\)
+    expectation p (\(\lambda\)x. abs (X x) pow 3) = sqrt (8 / Normal pi) * Normal (sig pow 3)
+  \end{alltt}
+\end{hol}
+
+\noindent
+\emph{Remark on Formalisation.}
+This result relies on integration over unbounded domains and special functions such as the gamma function $\Gamma(z)$.
+While it is analytically correct and standard in probability theory, its full formalisation in HOL4 is not yet feasible.
+HOL4 is based entirely on the Lebesgue integral and currently lacks:
+
+\begin{itemize}
+    \item integration over $\mathbb{R}$ with non-trivial densities like the Gaussian,
+    \item improper integrals as limits over infinite intervals,
+    \item and complex-analytic foundations needed for $\Gamma(z)$ where $z \in \mathbb{C}$.
+\end{itemize}
+
+Although the Riemann and Lebesgue integrals often agree for well-behaved functions like $x \mapsto |x|^3 e^{-x^2/2}$, this equivalence is not formalised in HOL4. Therefore, we treat the equation
+
+\[
+\mathbb{E}[|X|^3] = \sqrt{\frac{8}{\pi}} \cdot \sigma^3
+\]
+
+as an \emph{informally justified assumption}, grounded in classical analysis. A full mechanisation is deferred to future work on HOL4’s measure and complex analysis libraries.
+
+\subsubsection*{Informal Proof}
+
+For $Z \sim \mathcal{N}(0, 1)$:
+
+\[
+\mathbb{E}[|Z|^3] = \frac{1}{\sqrt{2\pi}} \int_{-\infty}^\infty |x|^3 e^{-x^2/2} dx
+= \frac{2}{\sqrt{2\pi}} \int_0^\infty x^3 e^{-x^2/2} dx.
+\]
+
+Change of variable $u = x^2/2$ gives:
+
+\[
+\int_0^\infty x^3 e^{-x^2/2} dx
+= \int_0^\infty \sqrt{2u} \cdot 2u \cdot e^{-u} \cdot \frac{1}{\sqrt{2u}} du
+= 2 \int_0^\infty u e^{-u} du = 2 \cdot \Gamma(2) = 2.
+\]
+
+So:
+
+\[
+\mathbb{E}[|Z|^3] = \frac{2}{\sqrt{2\pi}} \cdot 2 = \sqrt{\frac{8}{\pi}}.
+\]
+
+More generally, for any $\nu > -1$:
+
+\[
+\mathbb{E}[|Z|^\nu] = 2^{\nu/2} \cdot \frac{\Gamma\left( \frac{\nu + 1}{2} \right)}{\sqrt{\pi}}.
+\]
+
+See Winkelbauer \cite{winkelbauer2012moments} for a comprehensive derivation using special functions like Kummer's $\Phi$ and the parabolic cylinder function $D_\nu$.
+
+\emph{Remarks on the Standard Normal Density}
+
+The standard Gaussian density is:
+
+\[
+x \mapsto \frac{1}{\sqrt{2\pi}} e^{-x^2 / 2},
+\]
+
+which integrates to 1:
+
+\[
+\frac{1}{\sqrt{2\pi}} \int_{-\infty}^{+\infty} e^{-x^2 / 2} dx = 1.
+\]
+
+This confirms it defines a valid probability distribution.
+However, this fundamental fact is also not fully formalised in HOL4, due to limitations in its support for improper integrals over $\mathbb{R}$.
+We include it as an assumed classical result.
+
+\subsection{Lyapunov Ratio in the CLT}
+
+We now combine the third moment estimates into the Lyapunov ratio, which controls the Taylor approximation error in the CLT:
+
+\[
+\frac{\Gamma_n}{s_n^3}
+= \frac{\sum_j \mathbb{E}[|X_j|^3]}{\left( \sum_j \mathrm{Var}(X_j) \right)^{3/2}}.
+\]
+
+Under the Lyapunov condition:
+
+\[
+\lim_{n \to \infty} \frac{\Gamma_n}{s_n^3} = 0,
+\]
+
+which ensures convergence in distribution to the standard normal law.
+
+\subsection{Asymptotic Error Bound and Big-O Formalisation}
+
+The total Taylor approximation error in the Lindeberg replacement scheme is bounded by the expression:
+
+\begin{equation}
+\label{eq:taylor-bound}
+\frac{M}{6} \sum_{j = 1}^{n} \left( \frac{y_j}{s_n^3} + \frac{c \sigma_j^3}{s_n^3} \right),
+\end{equation}
+
+where \( y_j = \mathbb{E}[|X_j|^3] \), \( \sigma_j^2 = \text{Var}(X_j) \), and \( c = \sqrt{8/\pi} \) is the absolute third moment of a standard normal distribution.
+
+By Lyapunov’s inequality, we know \( \sigma_j^3 \le y_j \). Therefore, the error bound \eqref{eq:taylor-bound} becomes:
+
+\begin{equation} \label{eq:taylor-bigO}
+\left| \mathbb{E}[f(S_n / s_n)] - \mathbb{E}[f(N)] \right| = \mathcal{O}\left( \frac{\Gamma_n}{s_n^3} \right),
+\end{equation}
+
+where we define:
+
+\[
+\Gamma_n := \sum_{j=1}^n \mathbb{E}[|X_j|^3],
+\quad
+s_n^2 := \sum_{j=1}^n \mathrm{Var}(X_j).
+\]
+
+The inequality \eqref{eq:taylor-bound} and estimate \eqref{eq:taylor-bigO} follow the same structure as the proof of the Central Limit Theorem in \cite{Chung:2001}.
+
+\textbf{Definition 1.} (\emph{Big-O})
+\begin{hol}
+\begin{alltt}
+Definition BigO\_def :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)f g.
+BigO f g \(\Leftrightarrow\) \(\exists\)c n0.
+0 < c \(\land\) (\(\forall\)n. n0 \(\le\) n \(\Rightarrow\) abs (f n) \(\le\) c * abs (g n))
+\end{alltt}
+\end{hol}
+
+We also have formal support for additive and multiplicative bounds across sequences and sums, but in the proof of CLT, we used \textbf{Multiplication by constant} to move from the full Taylor bound to the simplified asymptotic expression:
+
+\textbf{Proposition 1.} (\emph{Big-O algebra}) We have the following derived rules:
+\begin{itemize}
+  \item[(a)] \textbf{Multiplication by constant}
+  \begin{hol}
+\begin{alltt}
+Theorem BigO\_MUL\_CONST :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)f g k.
+k \(\ne\) 0 \(\land\) BigO f g \(\Rightarrow\)
+BigO (\(\lambda\)n. k * f n) g
+\end{alltt}
+\end{hol}
+  \item[(b)] \textbf{Additive bound (sum)}
+  \begin{hol}
+\begin{alltt}
+Theorem BigO\_ADD :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)f1 f2 g1 g2.
+BigO f1 g1 \(\land\) BigO f2 g2 \(\Rightarrow\)
+BigO (\(\lambda\)n. f1 n + f2 n) (\(\lambda\)n. abs (g1 n) + abs (g2 n))
+\end{alltt}
+\end{hol}
+
+\begin{hol}
+\begin{alltt}
+Theorem BigO\_ADD\_MAX :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)f1 f2 g1 g2.
+BigO f1 g1 \(\land\) BigO f2 g2 \(\Rightarrow\)
+BigO (\(\lambda\)n. f1 n + f2 n) (\(\lambda\)n. max (abs (g1 n)) (abs (g2 n)))
+\end{alltt}
+\end{hol}
+  \item[(c)] \textbf{Multiplicative bound (product)}
+  \begin{hol}
+\begin{alltt}
+Theorem BigO\_MUL :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)f1 g1 f2 g2.
+BigO f1 g1 \(\land\) BigO f2 g2 \(\Rightarrow\)
+BigO (\(\lambda\)n. f1 n * f2 n) (\(\lambda\)n. g1 n * g2 n)
+\end{alltt}
+\end{hol}
+  \item[(d)] \textbf{Summation bound (series)}
+  \begin{hol}
+\begin{alltt}
+Theorem BigO\_SUM :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)f g.
+(\(\forall\)n. BigO (f n) (g n)) \(\Rightarrow\)
+(\(\forall\)n. BigO (\(\lambda\)x. \(\sum\) (\(\lambda\)i. f i x) (count n))
+               (\(\lambda\)x. \(\sum\) (\(\lambda\)i. abs (g i x)) (count n)))
+\end{alltt}
+\end{hol}
+\end{itemize}
+
+These lemmas are available for both \texttt{real} and \texttt{extreal} types, allowing seamless reasoning about expectations and variances in our formalised CLT development.
+
+\medskip
+\noindent
+\textbf{Conclusion.} Under the Lyapunov condition, we know \( \Gamma_n / s_n^3 \to 0 \), so the right-hand side of the estimate:
+
+\[
+\left| \mathbb{E}[f(S_n / s_n)] - \mathbb{E}[f(N)] \right| = \mathcal{O}\left( \frac{\Gamma_n}{s_n^3} \right)
+\]
+
+vanishes asymptotically. This completes the analytic portion of the CLT proof, establishing convergence in distribution to the standard normal for all test functions \( f \in \mathcal{C}^3 \).
+
+\subsection{The Central Limit Theorem}
+
+All the analytic components are now in place: we have constructed the auxiliary sequence, bounded the replacement error using Taylor's theorem and Lyapunov's inequality, and expressed the total error in terms of the Lyapunov ratio.
+The only remaining step is to bring these parts together into the convergence result.
+
+To remind the reader: we aim to show that the sequence of normalized sums of independent, mean-zero, real-valued random variables with finite variances and third moments converges in distribution to the standard normal.
+
+This follows the classical Lyapunov form of the Central Limit Theorem via the Lindeberg replacement trick, as presented in Chung’s textbook~\cite{Chung:2001}.
+Our formalization faithfully mirrors this structure.
+
+\medskip
+
+Because this is the primary goal of our development, we now state the final target theorem in our formalisation:
+
+\begin{hol}
+  \begin{alltt}
+    Theorem central\_limit\_theorem :
+    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p X N.
+    prob\_space p \(\land\)
+    ext\_normal\_rv N p 0 1 \(\land\)
+    (\(\forall\)i. real\_random\_variable (X i) p) \(\land\)
+    (\(\forall\)n. indep\_vars p X (\(\lambda\)i. Borel) (count n)) \(\land\)
+    (\(\forall\)i. expectation p (X i) = 0) \(\land\)
+    (\(\forall\)i. expectation p (\(\lambda\)x. (abs (X i x)) pow 3) < +\(\infty\)) \(\land\)
+    (\(\forall\)i. variance p (X i) < +\(\infty\)) \(\land\)
+    (\(\forall\)i. variance p (X i) \(\ne\) 0) \(\land\)
+    (\(\forall\)n. sqrt (second\_moments p X n) \(\ne\) 0) \(\land\)
+    ((\(\lambda\)n. third\_moments p X n / (sqrt (second\_moments p X n)) pow 3) \(\longrightarrow\) 0) sequentially
+    \(\Rightarrow\)
+    ((\(\lambda\)n x. \(\sum\) (\(\lambda\)i. X i x) (count n) / sqrt (second\_moments p X n)) \(\longrightarrow\) N)
+    (in\_distribution p)
+  \end{alltt}
+\end{hol}
+
+\medskip
+
+At the time of writing, all supporting lemmas and bounds have been formalised in HOL4, including moment control, Taylor expansion bounds, and the auxiliary variable construction.
+The remaining work consists of combining these results into a single convergence theorem.
+
+We treat this as the formal culmination of our project, and the proof is expected to follow with minimal additional infrastructure.

### 2025-05-19 conclusions.tex
init commit
Modified Ch. Introduction
Almost done report

--- 2024-11-30 background.tex
@@
-\chapter{Background and Related Work}
-\label{background}
+%%% -*- Mode: LaTeX; -*-
+\chapter{Conclusion}
+\label{concl}
 
-Background 123
+We have formalized in this thesis, to a great extent, a general version of the Central Limit Theorem (CLT) in HOL4 under Lyapunov's condition, with the Lindeberg replacement method.
+The version of the CLT proved herein is far more general than the classical i.i.d.
+case: it treats summands which are independent but not necessarily identically distributed, with variances and third absolute moments which are not necessarily constant over indices.
+This is in line with Chung's analytical development \cite{Chung:2001} and is a much stronger version of the theorem than those formalised in theorem provers such as Isabelle/HOL previously \cite{serafin2015formally}.
 
-% \begin{figure*}
-% \centering
-% \includegraphics[width=0.98\textwidth]{Chapter1/Images/Fig1.pdf}
-% \caption{caption}
-% \label{figCh1:figure}
-% \end{figure*}
+To our knowledge, this is the first mechanised proof of Lyapunov’s CLT under general independent sequences using Lindeberg replacement method, Taylor expansion and Big-O bounds in HOL4.
+Compared to the CLT formalized in Isabelle, which assumes the same distribution, our presentation deals with the additional challenge of summands being inhomogeneous.
+This introduces some technical awkwardness, especially in managing variable-specific error terms in Lindeberg replacement.
+The result is a version of the CLT valid under Lyapunov's condition—a broader and more applicable class—at the cost of much more intricate formal proofs.
 
-\renewcommand\thefigure{\thechapter .\arabic{figure}}
+The work involved developing over 6000 lines of HOL4 code, from moment inequalities, summability, Taylor approximation, and asymptotic Big-O estimation theorems and definitions.
+Most of this infrastructure either had to be developed from scratch or redeveloped to suit the requirements of probability formalisation in HOL4.
+The development significantly extends the scope of HOL4 for conducting convergence in distribution and limit theorem reasoning.
+
+Early progress on the project was hindered by the learning curve of HOL4 and an initial try at proving the CLT through moment-generating functions.
+Even though this path is textbook typical, it was unworkable in HOL4 since nothing on MGFs and improper integrals was available in the standard libraries.
+The transition to the Lindeberg substitute method, while more awkward, was successful and allowed all main analytic components of the CLT to be formally rigorised.
+
+\medskip
+
+\textbf{What has been completed:}
+\begin{itemize}
+    \item Construction of auxiliary sequences of Gaussian random variables;
+    \item Moment and variance control by Lyapunov-type inequalities;
+    \item Taylor expansion for the sum and remainder bounds;
+\item Big-O asymptotic bounds on the total Taylor approximation error.
+\end{itemize}
+
+\textbf{What remains:}
+The only remaining work is to complete the proof of the final convergence statement.
+Specifically, it entails formalising the absolute third moment of the normal distribution as an integral expressed in terms of the gamma function.
+This requires evaluating \( \mathbb{E}[|X|^3] \) for \( X \sim \mathcal{N}(0, \sigma^2) \), which involves improper integrals over \( \mathbb{R} \), as well as properties of special functions such as \( \Gamma(z) \).
+Unfortunately, the necessary infrastructure for complex-valued integration, special functions, and improper integrals is not yet available in the standard HOL4 libraries.
+
+In addition to the lack of special function formalisation, a more foundational gap remains between the classical Gauss–Riemann-style integration used in traditional probability theory and the fully formal Lebesgue integration developed in HOL4.
+Many textbook results rely on informal transitions between these two views of integration, which in a formal setting must be rigorously connected.
+Bridging this gap—formally relating improper Riemann integrals to their Lebesgue counterparts—is essential for justifying many classical computations, and is currently an area of planned future work.
+
+At the time of writing, the final error bound has been cleanly isolated and stated.
+All supporting lemmas, inequalities, and approximation steps are in place.
+The remaining step is therefore not conceptual but purely technical: implementing the final integration using extended analysis tools once they become available in HOL4.
+
+\medskip
+
+\noindent
+This thesis gives a good basis for further research work on the formalisation of limit theorems in probability theory.
+As support for complex integration and special functions in HOL4 evolves, it will be possible to complete the whole proof and perhaps extend to CLTs for dependent variables, martingales, or random vectors.
+The work presented here not only demonstrates that one can mechanise one of the classic results in probability theory but also develops reusable infrastructure for formalising probabilistic proofs with expectations, variances, Taylor approximations, and convergence in distribution.

### 2025-05-19 futurework.tex
init commit
Modified Ch. Introduction
Almost done report

--- 2024-11-30 background.tex
@@
-\chapter{Background and Related Work}
-\label{background}
+%%% -*- Mode: LaTeX; -*-
+\chapter{Future Work}
+\label{future}
 
-Background 123
+The formalisation of Lyapunov’s Central Limit Theorem (CLT) in this thesis has been developed to a nearly complete state.
+All components except the formal evaluation of $\mathbb{E}[|X|^3]$ for Gaussian $X$ have been mechanised.
+This includes more than 6,000 lines of definitions, lemmas, and theorems in HOL4.
+The final convergence theorem has been stated and is supported by all necessary lemmas.
+The remaining step is to combine all of crucial components and formalise the third absolute moment of the normal distribution:
 
-% \begin{figure*}
-% \centering
-% \includegraphics[width=0.98\textwidth]{Chapter1/Images/Fig1.pdf}
-% \caption{caption}
-% \label{figCh1:figure}
-% \end{figure*}
+\[
+\mathbb{E}[|X|^3] = \sqrt{8/\pi} \cdot \sigma^3.
+\]
 
-\renewcommand\thefigure{\thechapter .\arabic{figure}}
+This involves computing an improper integral for the gamma function, whose realization is a function of advances in real analysis and complex-valued functions formalism.
+HOL4 presently lacks sufficient support for improper integration over infinite intervals and integration with complex numbers.
+
+The author's subsequent commitment after this thesis is to complete this final proof step by establishing analytic tools necessary.
+This entails the closure of the gap between classical improper integrals (say, Riemann/Gauss-type) and formal Lebesgue integral in HOL4.
+Establishing a connection between these two is essential for justifying many classical results in a formal system.
+
+Beyond completing this proof, several natural directions remain for future work:
+
+\subsection*{1. Generalising the Central Limit Theorem}
+
+The version of the CLT formalised here assumes independent (but not identically distributed) summands. Several natural generalisations can be pursued:
+\begin{itemize}
+  \item \textbf{CLT for Martingales.} Extend the current proof to martingale difference sequences, which generalise independence to adapted stochastic processes \cite{hall2014martingale}.
+  \item \textbf{CLT under Weak Dependence.} Formalise convergence under weak dependence or mixing conditions, as outlined in Billingsley~\cite{billingsley2017probability}.
+  \item \textbf{Multivariate CLT.} Extend the proof to sequences of random vectors in \( \mathbb{R}^d \), a standard result in multivariate statistics.
+  \item \textbf{Berry–Esseen Bounds.} Provide a quantitative rate of convergence, leveraging the moment bounds and error control machinery developed in this thesis. \cite{berry1941accuracy}
+\end{itemize}
+
+\subsection*{2. Extending HOL4’s Analysis and Probability Libraries}
+
+Several features are currently missing or underdeveloped in HOL4 that would enable broader formalisation of classical probability:
+\begin{itemize}
+  \item Formal definitions and properties of the gamma function, and other special functions such as the error function (\texttt{erf}) and beta function.
+  \item Extension of the Lebesgue integral to handle improper and infinite-domain integrals.
+  \item Support for real-to-complex and complex-to-complex functions (e.g., \( \mathbb{R} \to \mathbb{C} \)), essential for characteristic function approaches and Fourier analysis.
+  \item Higher-level automation for asymptotic approximations, such as reasoning with Big-O notation directly over sequences and sums.
+\end{itemize}
+
+This work shows that it is possible not only to formalise advanced results in probability theory, but also to build reusable, modular libraries that can support future theorems in stochastic processes, statistics, and applied machine learning.
+In conclusion, this thesis provides a good foundation for extensions and establishes the feasibility of mechanizing high-level probability theory in HOL4.

### 2025-05-19 introduction.tex
Almost done report

@@
-%%% -*- Mode: LaTeX; -*-
 \chapter[Introduction]{Introduction}
@@
 
-The idea of a formalised mechanist is simple but powerful: if we want to trust mathematics in software, we must mechanise not only calculations but also proofs.
+The idea of a formalised mechanist is simple but powerful: to trust mathematics in software, we must mechanise not only calculations but also proofs.
 A mechanist builds a logic engine where every theorem arises from a small set of primitive rules.
-If those rules are sound, then everything derived from them inherits that soundness.
+If those rules are sound, everything derived from them inherits that soundness.
 
-HOL4 supports this mechanised philosophy~\cite{slind2008brief}.
-It encodes logic using a typed \(\lambda\)-calculus and enforces soundness by restricting theorem creation to a protected MetaLanguage (ML) type.
-The only way to construct a theorem is by applying primitive inference rules written in Standard ML.
-This ensures that the proof kernel checks every line of reasoning with absolute rigour.
+HOL4 supports this mechanised philosophy \cite{slind2008brief}.
+It encodes logic using a typed \(\lambda\)-calculus and restricts theorem creation to a protected MetaLanguage (ML) type, ensuring that every step is checked with mathematical rigour.
 
@@
 When many independent effects accumulate, their combined influence often resembles a bell curve.
-We observe this phenomenon across a wide range of domains—from biological traits and human measurements to economic indicators and communication errors.
-The CLT explains this universality.
+The CLT explains this universal phenomenon observed across diverse fields—from biology and economics to physics and computation.
 
-We follow the version of the Central Limit Theorem stated in~\cite{chung2000course} and proved using Lindeberg’s method. Let $\{X_j\}_{j=1}^n$ be a sequence of independent random variables such that:
+We formalise the Lyapunov version of the CLT from \cite{Chung:2001}, using Lindeberg’s replacement method.
+Let $\{X_j\}_{j=1}^n$ be a sequence of independent random variables with zero mean, finite variance, and third absolute moment:
 
-$$
+\[
 \mathbb{E}[X_j] = 0,\quad \mathbb{E}[X_j^2] = \sigma_j^2,\quad \mathbb{E}[|X_j|^3] = \gamma_j < \infty.
-$$
+\]
 
-Define the sum $S_n = \sum_{j=1}^n X_j$, the total variance $s_n^2 = \sum_{j=1}^n \sigma_j^2$, and the total third moment $\Gamma_n = \sum_{j=1}^n \gamma_j$. If the Lyapunov condition
+Define the sum \( S_n = \sum_{j=1}^n X_j \), total variance \( s_n^2 = \sum \sigma_j^2 \), and total third moment \( \Gamma_n = \sum \gamma_j \). Then, under the Lyapunov condition:
 
-$$
-\lim_{n \to \infty} \frac{\Gamma_n}{s_n^3} = 0
-$$
+\[
+\lim_{n \to \infty} \frac{\Gamma_n}{s_n^3} = 0,
+\]
 
-holds, then the normalised sum \(\frac{S_n}{s_n}\) converges in distribution to the standard normal:
+we have convergence in distribution:
+\[
+\frac{S_n}{s_n} \xrightarrow{d} \mathcal{N}(0, 1).
+\]
 
-$$
-\frac{S_n}{s_n} \xrightarrow{d} \mathcal{N}(0, 1).
-$$
-
-This convergence result explains why the normal distribution emerges from a wide variety of processes, even when individual random variables follow skewed or irregular distributions.
-
-We do not assume probability theory or integration as axioms.
-Instead, we construct them from first principles in HOL4 using its existing real number, measure theory, and Lebesgue integration libraries.
-We define random variables, convergence in distribution, and weak limits in this framework.
-We then prove the CLT using the Lindeberg replacement method and Taylor expansion bounds, following the structure outlined in~\cite{Chung:2001}.
 Here is the Lyapunov's form of the Central Limit Theorem in HOL4:
@@
   \begin{alltt}
-    Theorem central_limit_theorem :
-    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p X N. prob_space p \(\land\)
-    ext_normal_rv N p 0 1 \(\land\)
-    (\(\forall\)i. real_random_variable (X i) p) \(\land\)
-    (\(\forall\)n. indep_vars p X (\(\lambda\)i. Borel) (count n)) \(\land\)
+    Theorem central\_limit\_theorem :
+    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p X N.
+    prob\_space p \(\land\)
+    ext\_normal\_rv N p 0 1 \(\land\)
+    (\(\forall\)i. real\_random\_variable (X i) p) \(\land\)
+    (\(\forall\)n. indep\_vars p X (\(\lambda\)i. Borel) (count n)) \(\land\)
     (\(\forall\)i. expectation p (X i) = 0) \(\land\)
-    (\(\forall\)i. expectation p (\(\lambda\)x. (abs (X i x)) pow 3) < +\(\infty\)) \(\land\)
-    (\(\forall\)i. variance p (X i) +\(\infty\)) \(\land\)
+    (\(\forall\)i. expectation p (\(\lambda\)x. (abs (X i x))\(^3\)) < \(+\infty\)) \(\land\)
+    (\(\forall\)i. variance p (X i) < PosInf) \(\land\)
     (\(\forall\)i. variance p (X i) \(\ne\) 0) \(\land\)
-    (\(\forall\)n. (sqrt (second_moments p X n)) \(\ne\) 0) \(\land\)
-    ((\(\lambda\)n. (third_moments p X n) / ((sqrt (second_moments p X n)) pow 3)) --> 0)
-    sequentially \({\Leftrightarrow}\)
-    ((\(\lambda\)n x. (SIGMA (\(\lambda\)i. X i x) (count n)) / (sqrt (second_moments p X n))) --> N)
-    (in_distribution p)
+    (\(\forall\)n. (sqrt (second\_moments p X n)) \(\ne\) 0) \(\land\)
+    ((\(\lambda\)n. (third\_moments p X n) / ((sqrt (second\_moments p X n)) pow 3)) --> 0) sequentially
+    \(\Rightarrow\)
+    ((\(\lambda\)n x. (SIGMA (\(\lambda\)i. X i x) (count n)) / (sqrt (second\_moments p X n))) --> N)
+    (in\_distribution p)
   \end{alltt}
@@
 
-%-----------------------------------------------------
-
-%This formalisation produces a complete, machine-checked proof of Lyapunov’s Central Limit Theorem.
-It not only demonstrates the power of HOL4 for probabilistic reasoning, but also enriches its measure-theoretic library with tools necessary for formalising deeper results in probability and statistics.
-%\newline
-%-----------------------------------------------------
-
-Despite not yet completing the final convergence proof, this thesis represents a significant step toward a fully mechanised proof of Lyapunov’s Central Limit Theorem in HOL4.
-The author developed over 6000 lines of formalisation code, building most of the necessary measure-theoretic and probabilistic infrastructure.
-Early progress was slowed by the need to learn the HOL4 system and an initial attempt to formalise the CLT using moment-generating functions—a path that proved impractical due to missing theorems in the HOL4 library at the time.
-The current approach, based on Lindeberg's method and Taylor expansion, has succeeded in formalising all major components of the proof.
-The only remaining step—formalising the Taylor error bound  and completing the convergence argument—involves calculating the absolute third moment of the normal distribution.
-This means formalizing integrals using the gamma function and complex numbers, which are currently not yet fully realized in HOL4.
-At the time of writing the report, the author is still in the midst of this last step.
-Even if it is still incomplete at the submission deadline, the overall structure and supporting lemmas are all in place, and the last bound is now a question of careful implementation instead of conceptual difficulty.
-
-\section{Motivation}
-The Central Limit Theorem (CLT) is more than just a result in probability theory—it is the mathematical explanation behind the bell curve that appears across nature, science, and data.
-When we take a large number of independent random variables with finite variance, their properly normalised sum tends toward a normal distribution, regardless of the original distributions of the individual variables~\cite{fischer2011history}.This effect manifests in physical measurements, biological readings, economic results, and even in computation roundoff errors.
-The CLT illustrates the reasons why such behavior is not accidental but is necessary mathematically under gentle hypotheses.
-
-Although succinctly phrased, the CLT is of very deep mathematical organization.
-Most textbook proofs rely on informal approximations or skip foundational details.
-In formal verification, this level of vagueness is unacceptable.
-We need a version of the CLT where every assumption is explicit, every step justified, and the entire argument checked by machine.
-
-Some existing formal proofs of the CLT, such as the one in Isabelle/HOL~\cite{serafin2015formally}, focus on the independent identical distribution (i.i.d.) case and rely on characteristic functions and the Lévy Continuity Theorem.
-That approach is elegant, but it depends on complex analysis and Fourier transforms, and does not generalise easily to sums of non-identically distributed variables.
-Moreover, HOL4’s complex number library is still underdeveloped and lacks many of the theorems needed to support such an approach effectively.
-Attempting to formalise the necessary complex theory would demand substantial additional effort—likely more than what is needed for the Lindeberg-based approach we follow in this work.
-
-In this thesis, we aim for something different. We want to formalise a version of the CLT that:
-\begin{itemize}
-\item allows non-identically distributed variables;
-\item avoids characteristic functions entirely;
-\item uses only measure-theoretic and analytical tools available in HOL4.
-\end{itemize}
-
-We follow the proof given in~\cite{Chung:2001}, based on Lindeberg’s replacement trick and Taylor expansion bounds.
-This approach aligns closely with classical analysis and is modular enough to support future generalisations—such as multidimensional CLTs or stable laws.
-
-By formalising this proof in HOL4, we not only demonstrate that the CLT is derivable in a foundational logical system, but also contribute tools and theorems that strengthen HOL4’s probability theory library for future formalisation work.
-
-\section{Thesis Contributions}
-The contribution of this work is the formalisation of Lyapunov's form of the Central Limit Theorem in HOL4, using the Lindeberg replacement strategy and Taylor approximation techniques.
-This required building up the necessary infrastructure for real-valued random variables, expectations, variances, and convergence in distribution from the ground up.
-
-In particular:
-
-\begin{itemize}
-\item We constructed the formal setting for expressing the CLT in Lyapunov form, augmenting HOL4 with normalisation tools, moment summation tools, and standardisation tools.
-\item We established the background required to define and establish bounds on expectations using Taylor expansion, including a remainder term of order three.
-\item We proved key lemmas for the Lindeberg substitution trick, which summarizes how to replace original variables by Gaussian approximations and bound resulting error.
-\item We defined the final convergence goal in terms of convergence in distribution with a well-defined formulation and separation of technical specifications.
-\end{itemize}
-
-\section{Structure of the Thesis}
-The remainder of this thesis is organised as follows.
-Chapter 2 reviews the Central Limit Theorem, its historical development, and the version we aim to formalise.
-It also surveys related work, including formal proofs in systems such as Isabelle/HOL.
-Chapter 3 introduces the HOL4 theorem prover and outlines the relevant libraries and foundational setup.
-Chapter 4 presents the core of the formalisation, including the supporting lemmas and structure of the proof.
-Chapter 5 discusses the results and places this work in the context of other efforts in formalised mathematics.Chapter 6 summarises the thesis by stating the contributions and providing guidance for future development of probability theory in HOL4.
-This research proves that a formalisation of the Central Limit Theorem in HOL4 is possible.
-It enriches HOL4's probability library, provides reusable tools for future developments, and establishes the system as a platform for fully verified mathematics.
+Although the final convergence proof is not yet complete, this thesis represents a significant step toward a full mechanisation.
+Over 6000 lines of HOL4 code were developed, including formal definitions, theorems, and supporting libraries.
+An initial attempt using moment-generating functions was abandoned due to insufficient HOL4 support.
+The current Taylor-based approach has allowed us to formalise all major components.
+The only missing piece is a bound involving the gamma function, which requires integration techniques not yet supported in HOL4.
+This is a matter of technical realisation, not conceptual difficulty.

### 2025-05-19 preliminaries.tex
fixed latex packages
Modified Ch. Introduction
Almost done report

--- 2024-12-14 preliminaries.tex
@@
 Higher Order Logic (HOL) \cite{hol4, slind2008brief} is derived from the Logic of Computable Functions (LCF) \cite{gordon1979edinburgh, milner1972logic} created by Robin Milner and colleagues in 1972.
-HOL is an adaptation of Church's Simple Theory of Types (STT) \cite{church1940formulation}, where a higher-order version of Hilbert's choice operator \(\varepsilon \), Axiom of Infinity, and Rank-1 polymorphism have been added.
+HOL is an adaptation of Church's Simple Theory of Types (STT) \cite{church1940formulation}, where a higher-order version of Hilbert's choice operator $\varepsilon $, Axiom of Infinity, and Rank-1 polymorphism have been added.
 HOL4 implements the original HOL framework, while other theorem provers in the HOL family, such as Isabelle/HOL, include important extensions.
@@
 
-% HOL logic is based on a formal system of typed logical terms, where types are set-theoretic interpretation sets in a universe \(U\).
-The HOL type system is, relative to dependent types, considerably less complex yet still suitable for the formalization of various disciplines.
-HOL comes in with four type categories.
-As part of its Backus-Naur Form grammar it allows for:
-% \begin{itemize}
-%     \item \textbf{Primitive types:} For example, \texttt{bool} (representing a two-element set \( \textbf{2} \)) and \texttt{ind} (representing an infinite set \( I \)).
-%     \item \textbf{Type constructors:} Operators creating new types, such as functions (`->`) and Cartesian products (`*`).
-%     \item \textbf{Type variables:} Enable polymorphism; it is thus possible to write generic theorems and functions.
-%     \item \textbf{Defined types:} Derived types, either as subsets or as bijections of basic types.
-% \end{itemize}
+The type system of HOL establishes the structural framework within which all terms and expressions are guaranteed to be well-defined and logically consistent.
+Types in HOL denote sets within the universe $ U $, and every term bears a certain type.
+The type grammar is simple and very expressive, and thus able to construct a wide variety of mathematical and logical objects.
 
-HOL terms are representatives of such things and the grammar includes constants, variables, applications (function calls) and lambda abstractions.
-Quantifiers, for instance universal (!x.
-P(x)) and existential (?x.
-P(x)), are also provided in HOL - they are defined as specific lambda functions.
+The type grammar is defined as:
+\[
+\sigma ::= \alpha \ | \ c \ | \ (\sigma_1, \ldots, \sigma_n)\text{op} \ | \ \sigma_1 \to \sigma_2
+\]
+where:
+\begin{enumerate}
+    \item \textbf{Type Variables ($\alpha, \beta, \ldots$):} Generic placeholders that allow polymorphism to provide functions and predicates over different types.
+    \begin{itemize}
+        \item Example: The type variable $\alpha$ could indicate integers, Booleans, or functions.
+    \end{itemize}
+
+    \item \textbf{Atomic Types ($c$):} Fixed and pre-defined types within HOL. The two initial atomic types are:
+    \begin{itemize}
+        \item \texttt{bool}: The set of Boolean values $\{T, F\}$.
+        \item \texttt{ind}: The set composed by individuals (an infinite set).
+
+    \end{itemize}
+
+    \item \textbf{Compound Types ($(\sigma_1, \ldots, \sigma_n)\text{op}$):} Formed by applying type operators to other types.
+Their examples include Cartesian products, which designate the tuples over the elements.
+    \begin{itemize}
+        \item Example: The type $(\texttt{bool}, \texttt{ind}) \times$ represents pairs of a Boolean and an individual.
+    \end{itemize}
+
+    \item \textbf{Function Types ($\sigma_1 \to \sigma_2$):} Represent total functions mapping elements from a domain ($\sigma_1$) to a codomain ($\sigma_2$).
+    \begin{itemize}
+        \item Example: The type $\texttt{bool} \to \texttt{ind}$ indicates a function mapping both Boolean-values to individual-elements.
+    \end{itemize}
+\end{enumerate}
+
+For example, consider the following types:
+\begin{enumerate}
+    \item A function from integers to Booleans:
+   \[
+   f : \texttt{int} \to \texttt{bool}
+   \]
+   This type indicates that $f(x)$ is a function taking an integer $x$ and returning a Boolean.
+
+    \item A tuple containing a Boolean and a function:
+       \[
+       p : (\texttt{bool}, (\texttt{int} \to \texttt{bool}))
+       \]
+       This is a pair type $p = (b, f)$, where $b$ is a Boolean, and $f$ is a function mapping integers to Booleans.
+       \item The type system guarantees the consistency by making sure all terms are properly typed.
+So if $g : \texttt{int} \to \texttt{bool}$, then $g(5)$ as 5 is an integer, but, $g(T)$ would be invalid since $T$ is a Boolean, not an integer.
+Such stringent typing is at the level of terms to avoid self-contradictory values and assure that proofs built up in HOL are sound.
+\end{enumerate}
+
+In HOL, terms are representatives for elements of sets represented by their types.
+The grammar of the term deﬁnes the syntax and structure for the logical expressions that can be expressed and hence statements that could be well typed and logically valid.
+Terms in HOL are constructed from the following components:
+
+\[
+t ::= x \ | \ c \ | \ t \ t' \ | \ \lambda x. t
+\]
+
+where:
+\begin{enumerate}
+    \item \textbf{Variables} $(x, y, \ldots):$
+    \begin{itemize}
+        \item Represent placeholders for elements of a type.
+        \item Example: $x : \texttt{bool}$ stands for a Boolean variable.
+    \end{itemize}
+
+    \item \textbf{Constants ($c$):}
+    \begin{itemize}
+        \item Fixed entities such as $T$, $F$, mathematical operators, or predefined functions.
+        \item  Example: The constant $+$ defines addition for numeric types.
+    \end{itemize}
+
+   \item \textbf{Function Applications ($t \ t'$):}
+   \begin{itemize}
+       \item Define the application of a function to an argument. The term $f(x)$ applies the function $f$ to the variable $x$.
+        \item Example: If $f : \texttt{int} \to \texttt{real}$ and $x : \texttt{int}$, then $f(x)$ is a valid term of type $\texttt{real}$.
+   \end{itemize}
+
+\item \textbf{$\lambda$-Abstractions ($\lambda x. t$):}
+\begin{itemize}
+    \item Denote anonymous functions where $x$ is the input variable, and $t$ is the function body.
+    \item Example: $\lambda x. x + 1$ defines a function that increments its input by 1.
+\end{itemize}
+
+\end{enumerate}
+
+To ensure consistency, the terms of HOL should be well typed. Given a term $t_{\sigma}$ of type $\sigma$, its grammar can be generalized with type annotations:
+
+\[
+t_{\sigma} ::= x_{\sigma} \ | \ c_{\sigma} \ | \ (t_{\sigma_1 \to \sigma_2} \ t'_{\sigma_1})_{\sigma_2} \ | \ (\lambda x_{\sigma_1}. t_{\sigma_2})_{\sigma_1 \to \sigma_2}
+\]
+
+HOL's deductive system is considered the logical foundation for forming and checking a proof.
+HOL's deductive system may consist of eight primitive rules of inference, the definition of new theorems by existing theorems.
+These rules are the basic components and are required for all logical reasoning within HOL, ensuring that proofs are consistent, logically valid, and traced.
+The following are the eight main primitive inference rules in HOL:
+
+\begin{enumerate}
+    \item \textbf{Assumption Introduction (ASSUME):}
+    \begin{itemize}
+        \item Introduces a formula as an assumption.
+        \item Rule:
+        \item Example: From the assumption $P$, we conclude $P$.
+    \end{itemize}
+
+    \item \textbf{Reflexivity (REFL):}
+    \begin{itemize}
+        \item States that any term is equal to itself.
+        \item Rule:
+        \item Example: For $x : \texttt{int}$, $x = x$ is always true.
+    \end{itemize}
+
+    \item \textbf{Beta Conversion (BETA\_CONV):}
+    \begin{itemize}
+        \item Applies substitution in lambda abstractions.
+        \item Rule:
+        \item Example: $(\lambda x. x + 1)(5) \vdash 5 + 1$.
+    \end{itemize}
+
+    \item \textbf{Substitution (SUBST):}
+    \begin{itemize}
+        \item Replaces a term in a formula with another term proven to be equal.
+        \item Rule:
+        \item Example: From $x = y$ and $P(x)$, infer $P(y)$.
+    \end{itemize}
+
+    \item \textbf{Abstraction (ABS):}
+    \begin{itemize}
+        \item Generalizes an equation by abstracting a variable.
+        \item Rule:
+        \item Example: From $5 + 1 = 6$, infer $\lambda x. x + 1 = \lambda x. 6$.
+    \end{itemize}
+
+    \item \textbf{Type Instantiation (INST\_TYPE):}
+    \begin{itemize}
+        \item Specializes polymorphic functions or predicates to specific types.
+        \item Rule:
+        \item Example:
+    \end{itemize}
+
+    \item \textbf{Discharging Assumptions (DISCH):}
+    \begin{itemize}
+        \item Converts an assumption into an implication.
+        \item Rule:
+        \item Example: From $P \wedge Q$, infer $P \Rightarrow (Q \wedge P)$.
+    \end{itemize}
+
+    \item \textbf{Modus Ponens (MP):}
+    \begin{itemize}
+        \item Combines an implication and its premise to infer the conclusion.
+        \item Rule:
+        \item Example: From $x > 0 \Rightarrow x^2 > 0$ and $x > 0$, infer $x^2 > 0$.
+    \end{itemize}
+\end{enumerate}
+These inference rules ensure that all logical derivations are traceable to basic axioms and established theorems.
+Additionally, the deductive system forms the backbone of HOL4, ensuring that proofs are both rigorous and reliable.
+
+All proofs in HOL are fundamentally derived from a set of primitive inference rules and a core logical foundation.
+These rules define the semantics of two fundamental logical connectives: \textbf{equality} ($=$) and \textbf{implication} ($\implies$).
+Other logical connectives and first\-order quantifiers, such as logical truth ($T$), falsehood ($F$), conjunction ($\wedge$), disjunction ($\vee$), and existential quantification ($\exists$), are defined as lambda ($\lambda$) functions for consistency within the HOL framework:
+
+\begin{enumerate}
+    \item \textbf{Logical Truth ($T$)}
+    \begin{itemize}
+        \item Rule
+        \item True is represented as the equality of two identical boolean functions.
+    \end{itemize}
+
+    \item \textbf{Logical Falsehood ($F$)}
+    \begin{itemize}
+        \item Rule
+        \item False is defined to satisfy any boolean implication.
+    \end{itemize}
+
+    \item \textbf{Negation ($\neg$)}
+    \begin{itemize}
+        \item Rule
+        \item Negation is the implication of a boolean value leading to falsehood.
+    \end{itemize}
+
+    \item \textbf{Conjunction ($\wedge$)}
+    \begin{itemize}
+        \item Rule
+        \item Conjunction is defined as a logical function that evaluates nested implications.
+    \end{itemize}
+
+    \item \textbf{Conjunction ($\wedge$)}
+    \begin{itemize}
+        \item Rule
+        \item Conjunction is defined as a logical function that evaluates nested implications.
+    \end{itemize}
+
+    \item \textbf{Disjunction ($\vee$)}
+    \begin{itemize}
+        \item Rule
+        \item Disjunction is expressed through sequential implications.
+    \end{itemize}
+
+    \item \textbf{Universal Quantifier ($\forall$)}
+    \begin{itemize}
+        \item Rule
+        \item Universality asserts that a predicate holds for all elements of a type.
+    \end{itemize}
+
+    \item \textbf{Existential Quantifier ($\exists$)}
+    \begin{itemize}
+        \item Rule
+        \item Existence is defined using Hilbert’s choice operator ($\varepsilon$).
+    \end{itemize}
+
+HOL also defines constructs for mathematical operations, such as \textbf{one-to-one functions} ($One\_One$) and \textbf{onto functions} ($Onto$), to extend logical capabilities:
+
+    \item \textbf{One-to-One ($One\_One$)}
+    \begin{itemize}
+        \item Rule
+    \end{itemize}
+
+    \item \textbf{Onto ($Onto$)}
+    \begin{itemize}
+        \item Rule
+    \end{itemize}
+
+HOL includes the constant $Type\_Definition$, which defines new types as bijections of subsets of existing types:
+    \item \textbf{Type Definition $Type\_Definition$}
+    \begin{itemize}
+        \item Rule
+        \item This process is automated by the HOL Datatype package, simplifying the creation of new types.
+
+    \end{itemize}
+
+\end{enumerate}
+
+HOL’s standard theory is built upon four foundational axioms:
+\begin{enumerate}
+    \item \textbf{Boolean Cases (BOOL\_CASES\_AX)}
+    \begin{itemize}
+        \item Rule
+        \item This axiom ensures that any boolean value is either true or false.
+    \end{itemize}
+
+    \item \textbf{Eta Conversion (ETA\_AX)}
+    \begin{itemize}
+        \item Rule
+        \item Eta conversion describes the extensionality of functions.
+    \end{itemize}
+
+    \item \textbf{Hilbert’s Choice (SELECT\_AX)}
+    \begin{itemize}
+        \item Rule
+        \item This axiom relates the choice operator to existential quantification.
+    \end{itemize}
+
+    \item \textbf{Infinity (INFINITY\_AX)}
+    \begin{itemize}
+        \item Rule
+        \item The Axiom of Infinity ensures the existence of an infinite set.
+    \end{itemize}
+
+\end{enumerate}
+
+These axioms are generally sufficient for conventional formalization projects in HOL4. Adding new axioms is strongly discouraged, as it can compromise logical consistency.
 
 \section{Measure Theory}
+
+Measure theory provides the mathematical foundation for probability and integration.
+In this project, we rely on the HOL4 formalisation of measure spaces, $\sigma$-algebras, and measurable functions, as defined in the `measureTheory` library \cite{hol4}.
+
+\subsection{Measure Spaces}
+
+A \emph{measure space} is a triple $ (X, \Sigma, \mu) $, where:
+\begin{itemize}
+  \item $ X $ is the underlying space,
+  \item $ \Sigma $ is a $ \sigma $-algebra of subsets of $ X $,
+  \item $ \mu : \Sigma \to \overline{\mathbb{R}} $ is a measure.
+\end{itemize}
+
+In HOL4, these components are represented as:
+\begin{hol}
+  \begin{alltt}
+    Definition m\_space\_def :
+    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)sp sts \(\mu\).
+    m\_space (sp, sts, \(\mu\)) = sp
+  \end{alltt}
+\end{hol}
+
+\begin{hol}
+  \begin{alltt}
+    Definition measurable\_sets\_def :
+    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)sp sts \(\mu\).
+    measurable\_sets (sp, sts, \(\mu\)) = sts
+  \end{alltt}
+\end{hol}
+
+\begin{hol}
+  \begin{alltt}
+    Definition measure\_def :
+    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)sp sts \(\mu\).
+    measure (sp, sts, \(\mu\)) = \(\mu\)
+  \end{alltt}
+\end{hol}
+
+A measure $ \mu $ satisfies:
+\begin{enumerate}
+  \item \textbf{Non-negativity:} $ \mu(A) \geq 0 $ for all $ A \in \Sigma $,
+  \item \textbf{Null empty set:} $ \mu(\emptyset) = 0 $,
+  \item \textbf{Countable additivity:} For disjoint sets $ \{A_i\} \subseteq \Sigma $,
+  \[
+    \mu\left(\bigcup_{i=1}^\infty A_i\right) = \sum_{i=1}^\infty \mu(A_i).
+  \]
+\end{enumerate}
+
+HOL4 uses the extended real number type \texttt{extreal}, which includes $ +\infty $, $ -\infty $, and undefined values, to represent measure values.
+
+\subsection{$\sigma$-Algebras}
+
+A $\sigma$-algebra $ \Sigma $ over a set $ X $ is a collection of subsets satisfying:
+\begin{itemize}
+  \item $ X \in \Sigma $,
+  \item If $ A \in \Sigma $, then $ X \setminus A \in \Sigma $,
+  \item If $ A_i \in \Sigma $ for all $ i \in \mathbb{N} $, then $ \bigcup_i A_i \in \Sigma $.
+\end{itemize}
+
+\begin{hol}
+  \begin{alltt}
+    Definition sigma\_algebra\_def :
+    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)a.
+    sigma\_algebra a \(\Leftrightarrow\)
+    algebra a \(\land\)
+    (\(\forall\)c. countable c \(\land\) c \(\subseteq\) subsets a \(\Rightarrow\) BIGUNION c \(\in\) subsets a)
+  \end{alltt}
+\end{hol}
+
+\subsection{Measurable Functions}
+
+A function $ f : X \to Y $ is \emph{measurable} from measure space $ m $ to $ n $ if:
+\[
+  \forall A \in \texttt{measurable\_sets}(n),\quad f^{-1}(A) \in \texttt{measurable\_sets}(m).
+\]
+
+In HOL4, the set of measurable functions is defined as:
+\begin{hol}
+  \begin{alltt}
+    Definition measurable\_def :
+    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)a b.
+    measurable a b =
+    \{f \,|\,
+    f \(\in\) (space a \(\rightarrow\) space b) \(\land\)
+    (\(\forall\)s. s \(\in\) subsets b \(\Rightarrow\) PREIMAGE f s \(\cap\) space a \(\in\) subsets a)\}
+  \end{alltt}
+\end{hol}
+
+\subsection{Borel Sets and Measurable Functions}
+
+A fundamental component of probability and integration in HOL4 is the notion of Borel sets and Borel-measurable functions.
+These are provided by two standard theories: \texttt{real\_borel}, which handles real-valued Borel sets and functions, and \texttt{borel}, which generalises these definitions to extended real values (\texttt{extreal}) and constructs Borel and Lebesgue measure spaces.
+
+\paragraph{Borel $\sigma$-Algebra.}
+In HOL4, the Borel $\sigma$-algebra on $ \mathbb{R} $, denoted \texttt{borel}, is defined as the smallest $\sigma$-algebra generated by open subsets of $ \mathbb{R} $:
+\begin{hol}
+  \begin{alltt}
+    Definition borel :
+    \(\!\!\!{\turn}\!\!\!\!\)
+    borel = sigma \(\mathbb{U}\)(:real) \{s \,|\, open s\}
+  \end{alltt}
+\end{hol}
+
+Equivalently, \texttt{borel} can be generated using various families of half spaces or intervals, such as $ \{ x \mid x \leq a \} $, $ \{ x \mid a < x < b \} $, and so on.
+
+\paragraph{Extended Borel $\sigma$-Algebra.}
+The extended real-valued Borel $\sigma$-algebra, \texttt{Borel}, is defined over the type \texttt{extreal}, which includes $ +\infty $ and $ -\infty $.
+It extends \texttt{borel} by including atomic sets containing infinities:
+
+\begin{hol}
+  \begin{alltt}
+    Definition Borel :
+    \(\!\!\!{\turn}\!\!\!\!\)
+    Borel =
+    (\(\mathbb{U}\)(:extreal),
+    \{B' \,|\,
+    \(\exists\)B S.
+    B' = IMAGE Normal B \(\cup\) S \(\land\)
+    B \(\in\) subsets borel \(\land\)
+    S \(\in\) \{\(\emptyset\); \{\(-\infty\)\}; \{\(+\infty\)\}; \{\(-\infty\), \(+\infty\)\}\})
+  \end{alltt}
+\end{hol}
+
+This construction allows measurable sets to include points at infinity, which is necessary when working with unbounded expectations or improper integrals.
+
+\paragraph{Measurable Functions.}
+A function $ f : X \to Y $ is measurable from a $\sigma$-algebra $ a $ to $ b $ if the preimage of any measurable set in $ b $ lies in the $\sigma$-algebra $ a $. This is formalised as:
+
+\begin{hol}
+  \begin{alltt}
+    Definition measurable\_def :
+    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)a b.
+    measurable a b =
+    \{f \,|\,
+    f \(\in\) (space a \(\rightarrow\) space b) \(\land\)
+    (\(\forall\)s. s \(\in\) subsets b \(\Rightarrow\) PREIMAGE f s \(\cap\) space a \(\in\) subsets a)\}
+  \end{alltt}
+\end{hol}
+
+To simplify notation, HOL4 introduces:
+\begin{hol}
+\begin{verbatim}
+val _ = overload_on ("borel_measurable", ``\a. measurable a borel``);
+val _ = overload_on ("Borel_measurable", ``\a. measurable a Borel``);
+\end{verbatim}
+\end{hol}
+
+Thus, a real-valued function is \texttt{borel\_measurable a} if measurable with respect to \texttt{borel}, and similarly for \texttt{Borel\_measurable a} for extended real-valued functions.
+
+\paragraph{Measurability Closure Properties.}
+Both \texttt{borel\_measurable} and \texttt{Borel\_measurable} are closed under arithmetic operations such as addition, subtraction, and multiplication.
+For example, if $ f, g \in \texttt{borel\_measurable a} $, then $ f + g $, $ f - g $, and $ f \cdot g $ are also in \texttt{borel\_measurable a}.
+
+\begin{hol}
+\begin{alltt}
+Theorem in\_borel\_measurable\_add :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)a f g h.
+sigma\_algebra a \(\land\) f \(\in\) borel\_measurable a \(\land\)
+g \(\in\) borel\_measurable a \(\land\)
+(\(\forall\)x. x \(\in\) space a \(\Rightarrow\) h x = f x + g x) \(\Rightarrow\)
+h \(\in\) borel\_measurable a
+\end{alltt}
+\end{hol}
+
+\begin{hol}
+\begin{alltt}
+Theorem IN\_MEASURABLE\_BOREL\_ADD' :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)a f g h.
+sigma\_algebra a \(\land\) f \(\in\) Borel\_measurable a \(\land\)
+g \(\in\) Borel\_measurable a \(\land\)
+(\(\forall\)x. x \(\in\) space a \(\Rightarrow\) h x = f x + g x) \(\Rightarrow\)
+h \(\in\) Borel\_measurable a
+\end{alltt}
+\end{hol}
+
+\begin{hol}
+\begin{alltt}
+Theorem in\_borel\_measurable\_sub :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)a f g h.
+sigma\_algebra a \(\land\) f \(\in\) borel\_measurable a \(\land\)
+g \(\in\) borel\_measurable a \(\land\)
+(\(\forall\)x. x \(\in\) space a \(\Rightarrow\) h x = f x \({-}\) g x) \(\Rightarrow\)
+h \(\in\) borel\_measurable a
+\end{alltt}
+\end{hol}
+
+\begin{hol}
+\begin{alltt}
+Theorem IN\_MEASURABLE\_BOREL\_SUB' :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)a f g h.
+sigma\_algebra a \(\land\) f \(\in\) Borel\_measurable a \(\land\)
+g \(\in\) Borel\_measurable a \(\land\)
+(\(\forall\)x. x \(\in\) space a \(\Rightarrow\) h x = f x \({-}\) g x) \(\Rightarrow\)
+h \(\in\) Borel\_measurable a
+\end{alltt}
+\end{hol}
+
+\begin{hol}
+\begin{alltt}
+Theorem in\_borel\_measurable\_mul :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)a f g h.
+sigma\_algebra a \(\land\) f \(\in\) borel\_measurable a \(\land\)
+g \(\in\) borel\_measurable a \(\land\)
+(\(\forall\)x. x \(\in\) space a \(\Rightarrow\) h x = f x * g x) \(\Rightarrow\)
+h \(\in\) borel\_measurable a
+\end{alltt}
+\end{hol}
+
+\begin{hol}
+\begin{alltt}
+Theorem IN\_MEASURABLE\_BOREL\_TIMES' :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)a f g h.
+sigma\_algebra a \(\land\) f \(\in\) Borel\_measurable a \(\land\)
+g \(\in\) Borel\_measurable a \(\land\)
+(\(\forall\)x. x \(\in\) space a \(\Rightarrow\) h x = f x * g x) \(\Rightarrow\)
+h \(\in\) Borel\_measurable a
+\end{alltt}
+\end{hol}
+
+Infact, if a real-valued function is Borel-measurable, then its extension to extreal values is also Borel-measurable, and vice versa.
+
+\begin{hol}
+\begin{alltt}
+Theorem IN\_MEASURABLE\_BOREL\_IMP\_BOREL' :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)a f.
+sigma\_algebra a \(\land\) f \(\in\) borel\_measurable a \(\Rightarrow\)
+Normal \(\circ\) f \(\in\) Borel\_measurable a
+\end{alltt}
+\end{hol}
+
+\begin{hol}
+\begin{alltt}
+Theorem in\_borel\_measurable\_from\_Borel :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)a f.
+sigma\_algebra a \(\land\) f \(\in\) Borel\_measurable a \(\Rightarrow\)
+real \(\circ\) f \(\in\) borel\_measurable a
+\end{alltt}
+\end{hol}
 

### 2025-05-20 background.tex
draft v1

@@
 
-While the Isabelle/HOL technique is elegant and mathematically sound, it relies on involved analysis and Fourier transforms, which are well-supported in Isabelle but not yet fully formalised in HOL4.
+While the Isabelle/HOL technique is elegant and mathematically sound, it relies on involved analysis and characteristic functions (Fourier transforms), which are well-supported in Isabelle but not yet fully formalised in HOL4.
 HOL4 is lacking key theorems for complex-valued functions at present, so the characteristic function approach is unworkable here.

### 2025-05-20 clt.tex
draft v1

@@
 
-\textbf{Theorem 1.}
+\begin{theorem}[Existence of independent variables]
+\label{thm:existence-of-indep-vars}
 Let $D$ be a function that gives positive variances for $n > 0$ dimensions — that is, for each index $i < n$, $D(i) > 0$.
@@
 \end{itemize}
+\end{theorem}
 
@@
 
+\begin{theorem}[Construction of auxiliary sequence]
+  \label{thm:construct-aux-seq}
+  There exists a product probability space $p = p_1 \times p_2$ supporting two independent sequences of random variables $X'_i = X_i \circ \texttt{FST}$ and $Y'_i = Y_i \circ \texttt{SND}$ such that the interleaved sequence $Z_i$ combines both and preserves mutual independence and variance structure.
+\end{theorem}
+
+Or, formally:
+
 \begin{hol}
@@
 
-\textbf{Theorem 1.}(\emph{Taylor’s Theorem})
-
+\begin{theorem}[Taylor’s Theorem]
+  \label{thm:taylor}
 Let $f : \mathbb{R} \to \mathbb{R}$ be $n$-times differentiable on an interval $[a, x]$. Then there exists some $t \in (a, x)$ such that:
@@
 \]
+\end{theorem}
 
@@
 
-\textbf{Theorem 1.}(\emph{Taylor Remainder Bound})
-
+\begin{theorem}[Taylor Remainder Bound]
+\label{thm:taylor-remainder}
 The Taylor remainder theorem describes the difference between a function and its Taylor polynomial.
@@
 
-\textbf{Proposition X.} (\emph{Measurability and Integrability of Partial Sums})
-
-Let $X_0, \dots, X_{n-1}$ and $Y_0, \dots, Y_{n-1}$ be integrable real-valued random variables over a probability space $p$. Define, for each $j < n$, a partial sum function:
-
+\begin{lemma}[Measurability of Partial Sums]
+  Given real-valued integrable random variables $ X_0, \dots, X_{n-1} $ and $ Y_0, \dots, Y_{n-1} $, define:
   \[
-Z_j(x) = \sum_{i < j} Y_i(x) + \sum_{j \leq i < n} X_i(x)
+  Z_j(x) = \sum_{i<j} Y_i(x) + \sum_{j \le i < n} X_i(x).
   \]
+  Then each $ Z_j $ is also a real random variable and integrable.
+\end{lemma}
 
@@
 
-\textbf{Proposition X.} (\emph{Telescoping Sum over Finite Range})
-The following identity simplifies the sum of differences in a sequence:
+\begin{lemma}[Telescoping Sum Identity]
+  For any real-valued function \( f \) where each term is finite, we have:
+  \[
+  \sum_{j = 0}^{n-1} (f(j) - f(j+1)) = f(0) - f(n).
+  \]
+\end{lemma}
 
-\[
-\sum_{j = 0}^{n-1} \left( f(j) - f(j + 1) \right) = f(0) - f(n),
-\]
-
-provided $f(j)$ is finite for all $j$. This is formalised in HOL4 as:
+This is formalised in HOL4 as:
 
@@
 
-\textbf{Theorem 1.}(\emph{Replacement Lemma Statement})
+\begin{theorem}[Lindeberg Replacement Lemma]
+  \label{thm:lindeberg-replacement}
+  Let $X_j$, $Y_j$ be independent sequences of real random variables with bounded third moments, and let $f \in C^3_b$.
+Then the total replacement error between $X_j$ and $Y_j$ when replacing them one by one is bounded as follows...
+\end{theorem}
+
+In HOL4, it states as:
 
@@
 
+\begin{theorem}[Global Taylor Error Bound]
+  \label{thm:taylor-error-bound}
+  Given independent sequences $X_j$, $Y_j$, $Z_j$ with bounded third moments and a test function $f \in C^3_b$, the cumulative difference in expectations is bounded by the total third moments scaled by $s_n^{-3}$.
+\end{theorem}
+
 This is formalised in HOL4 as:
@@
 
-\textbf{Theorem 1.} (\emph{Lyapunov Inequality})
-Let $(\Omega, \mathcal{F}, \mathbb{P})$ be a probability space, and let $X$ be a random variable such that both $\mathbb{E}[|X|^r]$ and $\mathbb{E}[|X|^{r'}]$ are finite, for some $0 < r < r' < \infty$.
-Then:
+\begin{theorem}[Lyapunov Inequality]
+  \label{thm:lyapunov-ineq}
+  For a random variable $X$ with finite $r$-th and $r'$-th absolute moments where $0 < r < r'$, we have:
   \[
@@
   \]
+\end{theorem}
 
@@
 
-\textbf{Definition 1.} (\emph{Big-O})
+\begin{definition}[Big-O Notation]
+ \( f(n) = \mathcal{O}(g(n)) \) means that there exists a constant \( c > 0 \) and threshold \( n_0 \) such that for all \( n \ge n_0 \), we have \( |f(n)| \le c \cdot |g(n)| \).
+\end{definition}
+
+Or, formally:
 \begin{hol}
@@
 
-\textbf{Proposition 1.} (\emph{Big-O algebra}) We have the following derived rules:
+\begin{proposition}[Big-O algebra]
+  \label{prop:bigO-algebra}
 \begin{itemize}
@@
 \end{itemize}
+\end{proposition}
 
@@
 To remind the reader: we aim to show that the sequence of normalized sums of independent, mean-zero, real-valued random variables with finite variances and third moments converges in distribution to the standard normal.
+
+\begin{theorem}[Central Limit Theorem: Lyapunov Form]
+  \label{thm:central-limit}
+  Let $\{X_i\}$ be a sequence of independent, mean-zero, real random variables with variances $\text{Var}(X_i)$ and third absolute moments. Under Lyapunov's condition:
+  \[
+  \frac{\sum \mathbb{E}[|X_i|^3]}{\left( \sum \text{Var}(X_i) \right)^{3/2}} \to 0,
+  \]
+  we have:
+  \[
+  \frac{\sum X_i}{\sqrt{\sum \text{Var}(X_i)}} \xrightarrow{d} \mathcal{N}(0, 1).
+  \]
+\end{theorem}
 

### 2025-05-20 conclusions.tex
draft v1

--- 2024-11-30 background.tex
@@
-\chapter{Background and Related Work}
-\label{background}
+\chapter{Conclusion}
+\label{concl}
 
-Background 123
+This thesis has formalised a general version of the Central Limit Theorem (CLT) in HOL4 under Lyapunov's condition, using the Lindeberg replacement method.
+The version presented here is far more general than the classical i.i.d.
+case: it handles independent (but not identically distributed) summands with varying variances and third absolute moments.
+Our development closely follows Chung's analytical presentation~\cite{Chung:2001}, and surpasses earlier formal work such as the CLT in Isabelle/HOL~\cite{serafin2015formally} in both generality and technical depth.
 
-% \begin{figure*}
-% \centering
-% \includegraphics[width=0.98\textwidth]{Chapter1/Images/Fig1.pdf}
-% \caption{caption}
-% \label{figCh1:figure}
-% \end{figure*}
+To our knowledge, this is the first mechanised proof of Lyapunov’s CLT in HOL4 for general independent sequences, employing the Lindeberg replacement trick, Taylor expansion, and Big-O reasoning.
+Unlike earlier approaches that assume identical distributions, our formalisation accommodates inhomogeneous sequences, which introduces substantial technical overhead—particularly in bounding variable-specific replacement errors.
+However, this effort yields a more widely applicable theorem under Lyapunov’s condition.
 
-\renewcommand\thefigure{\thechapter .\arabic{figure}}
+Over 6000 lines of HOL4 code were developed to support this formalisation, including infrastructure for moment inequalities, summability, Taylor approximation, and asymptotic analysis.
+Much of this was written from scratch or adapted from existing theories to suit the strict requirements of probabilistic reasoning in HOL4.
+The resulting tools significantly enhance HOL4’s capacity to reason about convergence in distribution and limit theorems.
+
+Initial attempts to formalise the CLT via moment-generating functions proved impractical due to the lack of support for improper integrals and MGFs in HOL4’s libraries.
+The switch to the Lindeberg method, though more technically involved, allowed for a successful proof strategy grounded entirely in measure theory and real analysis.
+
+\medskip
+
+\textbf{What has been completed:}
+\begin{itemize}
+    \item Construction of auxiliary sequences of Gaussian random variables;
+    \item Moment and variance control via Lyapunov-type inequalities;
+    \item Taylor expansion bounds on individual replacement steps;
+    \item Big-O asymptotic estimation of the total error.
+\end{itemize}
+
+\textbf{What remains:}
+The only remaining work is to complete the proof of the final convergence statement.
+Specifically, it entails formalising the absolute third moment of the normal distribution as an integral expressed in terms of the gamma function.
+This requires evaluating \( \mathbb{E}[|X|^3] \) for \( X \sim \mathcal{N}(0, \sigma^2) \), which involves improper integrals over \( \mathbb{R} \), as well as properties of special functions such as \( \Gamma(z) \).
+Unfortunately, the necessary infrastructure for complex-valued integration, special functions, and improper integrals is not yet available in the standard HOL4 libraries.
+
+Beyond the lack of special-function support, a deeper gap remains between classical Riemann-based probabilistic analysis and the rigorous Lebesgue-style formalism adopted in HOL4.
+Many standard results rely on transitions between these views, which must be made fully explicit in a mechanised setting.
+Bridging this divide—by relating improper Riemann and Lebesgue integrals—is essential for validating textbook-level probabilistic computations, and forms an important direction for future work.
+
+At the time of writing, all analytic components have been formalised, and the final error bound has been cleanly isolated.
+The only missing piece is the formalisation of the third moment of the Gaussian, a purely technical step requiring extended integration support in HOL4.
+
+\medskip
+
+\noindent
+This work lays a solid foundation for future research on formalising limit theorems in probability theory.
+As support for complex integration and special functions in HOL4 grows, the full proof of the CLT will become attainable.
+Beyond that, the techniques developed here could be extended to CLTs for dependent sequences, martingales, or even random vectors.
+This thesis not only demonstrates the feasibility of formalising one of probability theory’s central results but also provides reusable infrastructure for reasoning about expectations, variances, Taylor approximations, and convergence in distribution within a theorem prover.

### 2025-05-20 futurework.tex
draft v1

--- 2024-11-30 background.tex
@@
-\chapter{Background and Related Work}
-\label{background}
+\chapter{Future Work}
+\label{future}
 
-Background 123
+The formalisation of Lyapunov’s Central Limit Theorem (CLT) in this thesis is nearly complete.
+All analytic components—including moment bounds, Taylor expansion, auxiliary sequences, and asymptotic estimates—have been mechanised.
+The final convergence theorem has been stated and is fully supported by the developed infrastructure.
+The only remaining step is to formalise the third absolute moment of the Gaussian distribution:
 
-% \begin{figure*}
-% \centering
-% \includegraphics[width=0.98\textwidth]{Chapter1/Images/Fig1.pdf}
-% \caption{caption}
-% \label{figCh1:figure}
-% \end{figure*}
+\[
+\mathbb{E}[|X|^3] = \sqrt{8/\pi} \cdot \sigma^3.
+\]
 
-\renewcommand\thefigure{\thechapter .\arabic{figure}}
+This requires formalising improper integrals and special functions, particularly the gamma function.
+Currently, HOL4 lacks sufficient support for such techniques.
+Bridging the gap between classical Riemann-style and formal Lebesgue integration will be critical for validating these results.
+The author intends to address this in future work.
+
+Beyond completing this proof, several directions arise naturally:
+
+\subsection*{1. Generalising the Central Limit Theorem}
+
+The current version handles independent, non-identically distributed variables. Future formal developments could include:
+
+\begin{itemize}
+\item \textbf{CLT for Martingales.} Extend the current proof to martingale difference sequences, which generalise independence to adapted stochastic processes \cite{hall2014martingale}.
+\item \textbf{CLT under Weak Dependence.} Formalise convergence under weak dependence or mixing conditions, as outlined in Billingsley~\cite{billingsley2017probability}.
+\item \textbf{Multivariate CLT.} Extend the proof to sequences of random vectors in \( \mathbb{R}^d \), a standard result in multivariate statistics.
+\item \textbf{Berry–Esseen Bounds.} Provide a quantitative rate of convergence, leveraging the moment bounds and error control machinery developed in this thesis. \cite{berry1941accuracy}
+\end{itemize}
+
+\subsection*{2. Extending HOL4's Mathematical Libraries}
+
+The formalisation also highlights areas where HOL4 could be extended to support more advanced probability theory:
+
+\begin{itemize}
+  \item Definitions and properties of the gamma, beta, and error functions.
+  \item Support for improper Lebesgue integration on infinite domains.
+  \item Formalisation of complex-valued functions and Fourier/characteristic methods.
+  \item Improved automation for Big-O and asymptotic reasoning in formal proofs.
+\end{itemize}
+
+These enhancements would enable broader mechanisation of probability theory, from classical results to modern topics in stochastic processes and statistics.
+The infrastructure developed in this thesis demonstrates the feasibility of such mechanisation and provides a reusable foundation for further research.

### 2025-05-20 introduction.tex
draft v1

@@
 
-In this thesis, we formalise one of the most important results in probability theory: the Central Limit Theorem (CLT).
+This thesis formalises one of the most fundamental results in probability theory: the Central Limit Theorem (CLT).
 When many independent effects accumulate, their combined influence often resembles a bell curve.
@@
 
-We formalise the Lyapunov version of the CLT from \cite{Chung:2001}, using Lindeberg’s replacement method.
-Let $\{X_j\}_{j=1}^n$ be a sequence of independent random variables with zero mean, finite variance, and third absolute moment:
+We focus on Lyapunov’s version of the CLT as presented in Chung~\cite{Chung:2001}, using Lindeberg’s replacement method.
 
+\medskip
+\noindent
+\textbf{The Central Limit Theorem (Lyapunov form):}
+\begin{quote}
+Let \( \{X_i\} \) be a sequence of independent real-valued random variables defined over a probability space \( p \), with the following properties:
+\begin{itemize}
+  \item Each \( X_i \) has mean zero: \( \mathbb{E}[X_i] = 0 \);
+  \item Each \( X_i \) has finite, non-zero variance: \( 0 < \text{Var}(X_i) < \infty \);
+  \item Each \( X_i \) has finite third absolute moment: \( \mathbb{E}[|X_i|^3] < \infty \);
+  \item The sequence \( X_i \) is independent in the formal sense used in HOL4 (via \texttt{indep\_vars} over Borel sets).
+\end{itemize}
+
+Define the sum \( S_n = \sum_{i=0}^{n-1} X_i \), and let the total variance be \( s_n^2 = \sum_{i=0}^{n-1} \text{Var}(X_i) \).
+
+If the Lyapunov condition holds:
 \[
-\mathbb{E}[X_j] = 0,\quad \mathbb{E}[X_j^2] = \sigma_j^2,\quad \mathbb{E}[|X_j|^3] = \gamma_j < \infty.
+\frac{\sum_{i=0}^{n-1} \mathbb{E}[|X_i|^3]}{s_n^3} \to 0 \quad \text{as } n \to \infty,
 \]
+then the normalised sum converges in distribution to the standard normal:
+\[
+\frac{S_n}{s_n} \xrightarrow{d} \mathcal{N}(0,1).
+\]
+\end{quote}
 
-Define the sum \( S_n = \sum_{j=1}^n X_j \), total variance \( s_n^2 = \sum \sigma_j^2 \), and total third moment \( \Gamma_n = \sum \gamma_j \). Then, under the Lyapunov condition:
+In HOL4, we express this formally as:
 
-\[
-\lim_{n \to \infty} \frac{\Gamma_n}{s_n^3} = 0,
-\]
-
-we have convergence in distribution:
-\[
-\frac{S_n}{s_n} \xrightarrow{d} \mathcal{N}(0, 1).
-\]
-
-Here is the Lyapunov's form of the Central Limit Theorem in HOL4:
+\begin{center}
+\scriptsize
 \begin{hol}
@@
 \end{hol}
+\end{center}
 
-Although the final convergence proof is not yet complete, this thesis represents a significant step toward a full mechanisation.
-Over 6000 lines of HOL4 code were developed, including formal definitions, theorems, and supporting libraries.
-An initial attempt using moment-generating functions was abandoned due to insufficient HOL4 support.
-The current Taylor-based approach has allowed us to formalise all major components.
-The only missing piece is a bound involving the gamma function, which requires integration techniques not yet supported in HOL4.
-This is a matter of technical realisation, not conceptual difficulty.
+Although the final convergence result is not yet complete, this work marks a significant advance toward a fully formalised CLT.
+Over 6000 lines of HOL4 code were written, including formal definitions, measure-theoretic infrastructure, Taylor approximations, moment bounds, and Big-O analysis.
+
+An initial attempt to follow the moment-generating function route was abandoned due to limited HOL4 support for improper integrals and Laplace transforms.
+The current approach, based on Taylor expansions and the Lindeberg method, allowed all core analytic steps to be formalised.
+The only missing component involves evaluating the third absolute moment of the normal distribution, which requires gamma-function integrals not yet formalised in HOL4.
+
+This remaining gap is technical, not conceptual. Once suitable analysis tools become available, the full convergence proof can be completed with minimal additional effort.

### 2025-05-20 preliminaries.tex
draft v1

@@
 HOL4 implements the original HOL framework, while other theorem provers in the HOL family, such as Isabelle/HOL, include important extensions.
-Such a simple logical basis makes HOL more accessible than those systems founded on much more advanced dependent type theories, such such the Calculus of Inductive and Co-Inductive Constructions constructed by Coq.
+Such a simple logical basis makes HOL more accessible than those systems founded on much more advanced dependent type theories, such as the Calculus of Inductive and Co-Inductive Constructions constructed by Coq.
 Therefore, theories and proofs founded on HOL are easier for a layman to comprehend rather than being lost in a complicated type theory.
@@
         \item Rule:
+          \begin{center}
+          \begin{tabular}{c}
+            \\ \hline
+            $t${\small\verb+ |- +}$t$ \\
+          \end{tabular}
+        \end{center}
         \item Example: From the assumption $P$, we conclude $P$.
@@
         \item Rule:
+          \begin{center}
+            \begin{tabular}{c}
+              \\ \hline
+                 {\small\verb+ |- +}$t${\small\verb+ = +}$t$ \\
+            \end{tabular}
+          \end{center}
         \item Example: For $x : \texttt{int}$, $x = x$ is always true.
@@
         \item Rule:
+          \begin{center}
+            \begin{tabular}{c}
+              \\ \hline
+              \holtxt{ |- (\bs$x$.$t_1$)$t_2$ = $t_1[t_2/x]$}
+            \end{tabular}
+          \end{center}
         \item Example: $(\lambda x. x + 1)(5) \vdash 5 + 1$.
@@
         \item Rule:
+          \begin{center}
+            \begin{tabular}{c}
+              $\Gamma_1${\small\verb+ |- +} $t_1${\small\verb+=+}$t'_1$ {\small\verb+  +} $\cdots$ {\small\verb+  +}
+              $\Gamma_n${\small\verb+ |- +} $t_n${\small\verb+=+}$t'_n$ {\small\verb+  +}
+              $\Gamma${\small\verb+ |- +} $t[t_1,\ldots,t_n]$ \\ \hline
+              $\Gamma_1 \cup \cdots
+              \cup \Gamma_n \cup \Gamma${\small\verb+ |- +} $t[t'_1,\ldots,t'_n]$ \\
+            \end{tabular}
+          \end{center}
         \item Example: From $x = y$ and $P(x)$, infer $P(y)$.
@@
         \item Rule:
+          \begin{center}
+            \begin{tabular}{c}
+              $\Gamma${\small\verb+ |- +}$t_1${\small\verb+ = +}$t_2$ \\ \hline
+              $\Gamma${\small\verb+ |- (\+}$x${\small\verb+.+}$t_1${\small\verb+) = (\+}$x${\small\verb+.+}$t_2${\small\verb+)+} \\
+            \end{tabular}
+          \end{center}
         \item Example: From $5 + 1 = 6$, infer $\lambda x. x + 1 = \lambda x. 6$.
@@
         \item Rule:
-        \item Example:
+          \begin{center}
+            \begin{tabular}{c}
+              $\Gamma${\small\verb+ |- +}$t$ \\ \hline
+              $\Gamma[\sigma_1,\ \ldots\ ,\sigma_n/\alpha_1,\ \ldots\ ,\alpha_n]${\small\verb+ |- +}$t[\sigma_1,\ \ldots\ ,\sigma_n/\alpha_1,\ \ldots\ ,\alpha_n]$
+            \end{tabular}
+          \end{center}
+        \item Example: Consider the polymorphic identity theorem:
+          $     \vdash \forall x.\ x = x
+          $
+          This holds for any type \texttt{$\alpha$}. Using \texttt{INST\_TYPE}, we can instantiate the type variable to a specific type, such as \texttt{real}. The result is:
+          $     \vdash \forall x:\texttt{real}.\ x = x
+          $
+          This instantiated version now applies to real numbers specifically, which is often necessary in the context of formalising analysis or probability theory, where many theorems apply to real-valued functions or variables.
     \end{itemize}
@@
         \item Rule:
+          \begin{center}
+            \begin{tabular}{c}
+              $\Gamma${\small\verb+ |- +} $t_2$ \\ \hline
+              $\Gamma{-}\{t_1\}${\small\verb+ |- +} $t_1${\small\verb+ ==> +}$t_2$
+            \end{tabular}
+          \end{center}
         \item Example: From $P \wedge Q$, infer $P \Rightarrow (Q \wedge P)$.
@@
         \item Rule:
+          \begin{center}
+            \begin{tabular}{c}
+              $\Gamma_1${\small\verb+ |- +}$t_1${\small\verb+ ==> +}$t_2$ {\small\verb+     +} $\Gamma_2${\small\verb+ |- +}$t_1$ \\
+              \hline
+              $\Gamma_1 \cup \Gamma_2${\small\verb+ |- +}$t_2$ \\
+            \end{tabular}
+          \end{center}
         \item Example: From $x > 0 \Rightarrow x^2 > 0$ and $x > 0$, infer $x^2 > 0$.
@@
     \item Rule
+\begin{verbatim}
+T_DEF              |- T  = ((\x:bool. x) = (\x. x))
+\end{verbatim}
         \item True is represented as the equality of two identical boolean functions.
@@
         \item Rule
+\begin{verbatim}
+F_DEF           |- F  = !t. t
+\end{verbatim}
         \item False is defined to satisfy any boolean implication.
@@
         \item Rule
+\begin{verbatim}
+NOT_DEF         |- ~  = (\t. t ==> F)
+\end{verbatim}
         \item Negation is the implication of a boolean value leading to falsehood.
@@
         \item Rule
-        \item Conjunction is defined as a logical function that evaluates nested implications.
-    \end{itemize}
-
-    \item \textbf{Conjunction ($\wedge$)}
-    \begin{itemize}
-        \item Rule
+\begin{verbatim}
+AND_DEF         |- /\ = \t1 t2. !t. (t1 ==> t2 ==> t) ==> t
+\end{verbatim}
         \item Conjunction is defined as a logical function that evaluates nested implications.
@@
         \item Rule
+\begin{verbatim}
+OR_DEF          |- \/ = \t1 t2. !t. (t1 ==> t) ==> (t2 ==> t) ==> t
+\end{verbatim}
         \item Disjunction is expressed through sequential implications.
@@
         \item Rule
+\begin{verbatim}
+FORALL_DEF      |- !  = \P:'a->bool. P = (\x. T)
+\end{verbatim}
         \item Universality asserts that a predicate holds for all elements of a type.
@@
         \item Rule
+\begin{verbatim}
+EXISTS_DEF      |- ?  = \P:'a->bool. P($@ P)
+\end{verbatim}
         \item Existence is defined using Hilbert’s choice operator ($\varepsilon$).
@@
         \item Rule
+          \begin{alltt}
+            \(\!\!\!{\turn}\!\!\!\!\) ONE_ONE = (\(\lambda\)f. \(\forall\)x1 x2. f x1 = f x2 \(\Rightarrow\) x1 = x2)
+          \end{alltt}
     \end{itemize}
@@
         \item Rule
+          \begin{alltt}
+            \(\!\!\!{\turn}\!\!\!\!\) ONTO = (\(\lambda\)f. \(\forall\)y. \(\exists\)x. y = f x)
+          \end{alltt}
     \end{itemize}
@@
 HOL includes the constant $Type\_Definition$, which defines new types as bijections of subsets of existing types:
-    \item \textbf{Type Definition $Type\_Definition$}
+    \item \textbf{Type Definition (Type\_Definition)}
     \begin{itemize}
         \item Rule
+          \begin{hol}
+\begin{verbatim}
+  |- TYPE_DEFINITION (P:'a->bool) (rep:'b->'a) =
+  (!x' x''. (rep x' = rep x'') ==> (x' = x'')) /\
+  (!x. P x = (?x'. x = rep x'))
+\end{verbatim}
+          \end{hol}
         \item This process is automated by the HOL Datatype package, simplifying the creation of new types.
@@
         \item Rule
+          \begin{alltt}
+            \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)t. (t \({\Leftrightarrow}\) T) \(\lor\) (t \({\Leftrightarrow}\) F)
+          \end{alltt}
         \item This axiom ensures that any boolean value is either true or false.
@@
         \item Rule
+          \begin{alltt}
+            \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)t. (\(\lambda\)x. t x) = t
+          \end{alltt}
         \item Eta conversion describes the extensionality of functions.
@@
         \item Rule
+          \begin{alltt}
+            \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)P x. P x \(\Rightarrow\) P (@ P)
+          \end{alltt}
         \item This axiom relates the choice operator to existential quantification.
@@
         \item Rule
+          \begin{alltt}
+            \(\!\!\!{\turn}\!\!\!\!\) \(\exists\)f. ONE_ONE f \(\land\) \(\neg\)ONTO f
+          \end{alltt}
         \item The Axiom of Infinity ensures the existence of an infinite set.
@@
 In HOL4, these components are represented as:
+\begin{center}
+\scriptsize
 \begin{hol}
@@
 \end{hol}
+\end{center}
 
@@
 
-A fundamental component of probability and integration in HOL4 is the notion of Borel sets and Borel-measurable functions.
-These are provided by two standard theories: \texttt{real\_borel}, which handles real-valued Borel sets and functions, and \texttt{borel}, which generalises these definitions to extended real values (\texttt{extreal}) and constructs Borel and Lebesgue measure spaces.
+A further fundamental building block for probability and integration in HOL4 is the theory of Borel sets and Borel-measurable functions.
+These are provided by two standard theories: \texttt{real\_borel}, which considers real-valued Borel sets and functions, and \texttt{borel}, which extends these definitions to extended real values (\texttt{extreal}) and builds up Borel and Lebesgue measure spaces.
 
@@
 
-Equivalently, \texttt{borel} can be generated using various families of half spaces or intervals, such as $ \{ x \mid x \leq a \} $, $ \{ x \mid a < x < b \} $, and so on.
+Equivalently, \texttt{borel} can be generated using various families of half spaces or intervals, e.g $ \{ x \mid x \leq a \} $, $ \{ x \mid a < x < b \} $, and so on.
 
@@
 
-This construction allows measurable sets to include points at infinity, which is necessary when working with unbounded expectations or improper integrals.
+The construction allows for measurable sets to have points at infinity, as required when working with unbounded expectations or improper integrals.
 
 \paragraph{Measurable Functions.}
-A function $ f : X \to Y $ is measurable from a $\sigma$-algebra $ a $ to $ b $ if the preimage of any measurable set in $ b $ lies in the $\sigma$-algebra $ a $. This is formalised as:
+A function $f : X \to Y$ is measurable from a $\sigma$-algebra $a$ to $b$ if the preimage of every set in $b$ that is measurable is in the $\sigma$-algebra $a$.
+
+Or, formally:
 
@@
 
+HOL4 formalisation of Lebesgue integration provides a strong foundation for reasoning about expectations, variances, and convergence in probability.
+This section briefly introduces the key constructs and theorems that we use throughout the Central Limit Theorem formalisation.
+
+\subsection{Positive Simple Functions and Integration}
+
+Nonnegative measurable functions in HOL4 are approximated by positive simple functions, piecewise constant functions taking finitely many values on measurable disjoint sets.
+A function $f$ is represented as:
+
+\[
+f(x) = \sum_{j=1}^n y_j \cdot \mathbf{1}_{A_j}(x), \quad \text{where } A_j \text{ are disjoint measurable sets}.
+\]
+
+This is formalised in HOL4 using:
+\begin{hol}
+\begin{alltt}
+Theorem pos\_simple\_fn\_def :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)m f s a x.
+pos\_simple\_fn m f s a x \({\Leftrightarrow}\)
+(\(\forall\)t. t \(\in\) m\_space m \(\Rightarrow\) 0 \(\le\) f t) \(\land\)
+(\(\forall\)t. t \(\in\) m\_space m \(\Rightarrow\)
+        f t = \(\sum\) (\(\lambda\)i. Normal (x i) * \(\mathbbm{1}\) (a i) t) s) \(\land\)
+(\(\forall\)i. i \(\in\) s \(\Rightarrow\) a i \(\in\) measurable\_sets m) \(\land\)
+FINITE s \(\land\)
+(\(\forall\)i. i \(\in\) s \(\Rightarrow\) 0 \(\le\) x i) \(\land\)
+(\(\forall\)i j. i \(\in\) s \(\land\) j \(\in\) s \(\land\) i \(\ne\) j \(\Rightarrow\)
+        DISJOINT (a i) (a j)) \(\land\)
+BIGUNION (IMAGE a s) = m\_space m
+\end{alltt}
+\end{hol}
+
+The integral of such a function is defined as:
+\begin{hol}
+\begin{alltt}
+Definition pos\_simple\_fn\_integral\_def :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)m s a x.
+pos\_simple\_fn\_integral m s a x =
+\(\sum\) (\(\lambda\)i. Normal (x i) * measure m (a i)) s
+\end{alltt}
+\end{hol}
+
+\subsection{Positive and General Lebesgue Integral}
+
+The positive integral of a nonnegative function $f$ is defined as the supremum of the integrals of all positive simple functions below $f$:
+\begin{hol}
+\begin{alltt}
+Definition pos\_fn\_integral\_def :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)m f.
+\(\int\)\(\sp{+}\) m f =
+sup \{r \,|\, (\(\exists\)g. r \(\in\) psfis m g \(\land\)
+                \(\forall\)x. x \(\in\) m\_space m \(\Rightarrow\) g x \(\le\) f x)\}
+\end{alltt}
+\end{hol}
+
+The general Lebesgue integral of a function $f$ is defined as:
+
+\[
+\int f = \int^+ f^+ - \int^+ f^-
+\]
+where $f^+ = \max(f, 0)$ and $f^- = \max(-f, 0)$.
+
+\begin{hol}
+\begin{alltt}
+Definition integral\_def :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)m f.
+\(\int\) m f = \(\int\)\(\sp{+}\) m f\(\sp{+}\) \({-}\) \(\int\)\(\sp{+}\) m f\(\sp{-}\)
+\end{alltt}
+\end{hol}
+
+\begin{hol}
+\begin{alltt}
+Definition fn\_plus\_def :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)f.
+f\(\sp{+}\) = (\(\lambda\)x. if 0 < f x then f x else 0)
+\end{alltt}
+\end{hol}
+
+\begin{hol}
+\begin{alltt}
+Definition fn\_minus\_def :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)f.
+f\(\sp{-}\) = (\(\lambda\)x. if f x < 0 then -f x else 0)
+\end{alltt}
+\end{hol}
+
+\subsection{Integrability}
+
+A function is integrable if it is Borel-measurable and the integrals of both its positive and negative parts are finite:
+\begin{hol}
+\begin{alltt}
+Definition integrable\_def :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)m f.
+integrable m f \({\Leftrightarrow}\)
+f \(\in\) Borel\_measurable (measurable\_space m) \(\land\)
+\(\int\)\(\sp{+}\) m f\(\sp{+}\) \(\ne\) +\(\infty\) \(\land\)
+\(\int\)\(\sp{+}\) m f\(\sp{-}\) \(\ne\) +\(\infty\)
+\end{alltt}
+\end{hol}
+
+\subsection{Almost Everywhere and Null Sets}
+
+A property $P(x)$ holds almost everywhere (AE) if it holds outside of a null set:
+\begin{hol}
+\begin{alltt}
+Definition AE\_DEF :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)m P.
+(AE x::m. P x) \({\Leftrightarrow}\)
+\(\exists\)N. null\_set m N \(\land\)
+\(\forall\)x. x \(\in\) m\_space m DIFF N \(\Rightarrow\) P x
+\end{alltt}
+\end{hol}
+
+\subsection{Monotone Convergence Theorem}
+
+In approximating general nonnegative functions by simple functions, the following theorem allows taking limits under the integral:
+\begin{hol}
+\begin{alltt}
+Theorem lebesgue\_monotone\_convergence :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)m f fi.
+measure\_space m \(\land\)
+(\(\forall\)i. fi i \(\in\) Borel\_measurable (measurable\_space m)) \(\land\)
+(\(\forall\)i x. x \(\in\) m\_space m \(\Rightarrow\) 0 \(\le\) fi i x) \(\land\)
+(\(\forall\)x. x \(\in\) m\_space m \(\Rightarrow\)
+   mono\_increasing (\(\lambda\)i. fi i x)) \(\land\)
+(\(\forall\)x. x \(\in\) m\_space m \(\Rightarrow\)
+   sup (IMAGE (\(\lambda\)i. fi i x) \(\mathbb{U}\)(:num)) = f x) \(\Rightarrow\)
+\(\int\)\(\sp{+}\) m f = sup (IMAGE (\(\lambda\)i. \(\int\)\(\sp{+}\) m (fi i)) \(\mathbb{U}\)(:num))
+\end{alltt}
+\end{hol}
+
+\subsection{Fubini and Tonelli Theorems}
+
+In formalising the Lindeberg replacement method, we often integrate over product spaces involving independent sequences of random variables.
+To justify iterated integrals in these settings, we rely on HOL4’s formalisation of Tonelli’s and Fubini’s Theorems, which ensure that integrals over product measure spaces can be computed as nested one-dimensional integrals.
+
+\paragraph{Tonelli's Theorem.}
+Tonelli's theorem applies to nonnegative functions and guarantees that the double integral equals the iterated integral regardless of whether the value is finite or infinite:
+
+\begin{hol}
+\begin{alltt}
+Theorem TONELLI :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)X Y A B u v f.
+sigma\_finite\_measure\_space (X,A,u) \(\land\)
+sigma\_finite\_measure\_space (Y,B,v) \(\land\)
+f \(\in\) Borel\_measurable ((X,A) \(\times\) (Y,B)) \(\land\)
+(\(\forall\)s. s \(\in\) X \(\times\) Y \(\Rightarrow\) 0 \(\le\) f s) \(\Rightarrow\)
+(\(\forall\)y. y \(\in\) Y \(\Rightarrow\) (\(\lambda\)x. f (x,y)) \(\in\) Borel\_measurable (X,A)) \(\land\)
+(\(\forall\)x. x \(\in\) X \(\Rightarrow\) (\(\lambda\)y. f (x,y)) \(\in\) Borel\_measurable (Y,B)) \(\land\)
+(\(\lambda\)x. \(\int\)\(\sp{+}\) (Y,B,v) (\(\lambda\)y. f (x,y))) \(\in\) Borel\_measurable (X,A) \(\land\)
+(\(\lambda\)y. \(\int\)\(\sp{+}\) (X,A,u) (\(\lambda\)x. f (x,y))) \(\in\) Borel\_measurable (Y,B) \(\land\)
+\(\int\)\(\sp{+}\) ((X,A,u) \(\times\) (Y,B,v)) f =
+\(\int\)\(\sp{+}\) (Y,B,v) (\(\lambda\)y. \(\int\)\(\sp{+}\) (X,A,u) (\(\lambda\)x. f (x,y))) \(\land\)
+\(\int\)\(\sp{+}\) ((X,A,u) \(\times\) (Y,B,v)) f =
+\(\int\)\(\sp{+}\) (X,A,u) (\(\lambda\)x. \(\int\)\(\sp{+}\) (Y,B,v) (\(\lambda\)y. f (x,y)))
+\end{alltt}
+\end{hol}
+
+\paragraph{Fubini's Theorem.}
+Fubini’s theorem generalises Tonelli’s result to integrable (not necessarily nonnegative) functions. It justifies interchanging the order of integration under integrability assumptions:
+\begin{hol}
+\begin{alltt}
+Theorem FUBINI :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)X Y A B u v f.
+sigma\_finite\_measure\_space (X,A,u) \(\land\)
+sigma\_finite\_measure\_space (Y,B,v) \(\land\)
+f \(\in\) Borel\_measurable ((X,A) \(\times\) (Y,B)) \(\land\)
+(\(\int\)\(\sp{+}\) ((X,A,u) \(\times\) (Y,B,v)) (abs \(\circ\) f) \(\ne\) +\(\infty\) \(\lor\)
+ \(\int\)\(\sp{+}\) (Y,B,v) (\(\lambda\)y. \(\int\)\(\sp{+}\) (X,A,u) (\(\lambda\)x. (abs \(\circ\) f) (x,y))) \(\ne\) +\(\infty\) \(\lor\)
+ \(\int\)\(\sp{+}\) (X,A,u) (\(\lambda\)x. \(\int\)\(\sp{+}\) (Y,B,v) (\(\lambda\)y. (abs \(\circ\) f) (x,y))) \(\ne\) +\(\infty\)) \(\Rightarrow\)
+\(\int\)\(\sp{+}\) ((X,A,u) \(\times\) (Y,B,v)) (abs \(\circ\) f) \(\ne\) +\(\infty\) \(\land\)
+\(\int\)\(\sp{+}\) (Y,B,v) (\(\lambda\)y. \(\int\)\(\sp{+}\) (X,A,u) (\(\lambda\)x. (abs \(\circ\) f) (x,y))) \(\ne\) +\(\infty\) \(\land\)
+\(\int\)\(\sp{+}\) (X,A,u) (\(\lambda\)x. \(\int\)\(\sp{+}\) (Y,B,v) (\(\lambda\)y. (abs \(\circ\) f) (x,y))) \(\ne\) +\(\infty\) \(\land\)
+integrable ((X,A,u) \(\times\) (Y,B,v)) f \(\land\)
+(AE y::(Y,B,v). integrable (X,A,u) (\(\lambda\)x. f (x,y))) \(\land\)
+(AE x::(X,A,u). integrable (Y,B,v) (\(\lambda\)y. f (x,y))) \(\land\)
+integrable (X,A,u) (\(\lambda\)x. \(\int\) (Y,B,v) (\(\lambda\)y. f (x,y))) \(\land\)
+integrable (Y,B,v) (\(\lambda\)y. \(\int\) (X,A,u) (\(\lambda\)x. f (x,y))) \(\land\)
+\(\int\) ((X,A,u) \(\times\) (Y,B,v)) f =
+\(\int\) (Y,B,v) (\(\lambda\)y. \(\int\) (X,A,u) (\(\lambda\)x. f (x,y))) \(\land\)
+\(\int\) ((X,A,u) \(\times\) (Y,B,v)) f =
+\(\int\) (X,A,u) (\(\lambda\)x. \(\int\) (Y,B,v) (\(\lambda\)y. f (x,y)))
+\end{alltt}
+\end{hol}
+
+In this thesis, these theorems are applied in constructing auxiliary sequences of independent random variables from product spaces and evaluating their expectations.
+Their use guarantees that swapping integration order is mathematically and formally valid in our HOL4 development.
+
+\subsection{$\mathcal{L}^p$ Spaces and Fundamental Inequalities}
+
+In HOL4, the $\mathcal{L}^p$ space over a measure space $m$ is defined as the set of (extended) real-valued Borel-measurable functions whose $p$-th absolute power is integrable:
+\begin{hol}
+\begin{alltt}
+Theorem lp\_space\_alt\_finite :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p m f.
+0 < p \(\land\) p \(\ne\) +\(\infty\) \(\Rightarrow\)
+(f \(\in\) lp\_space p m \({\Leftrightarrow}\)
+ f \(\in\) Borel\_measurable (measurable\_space m) \(\land\)
+ \(\int\)\(\sp{+}\) m (\(\lambda\)x. abs (f x) powr p) \(\ne\) +\(\infty\))
+\end{alltt}
+\end{hol}
+
+The associated seminorm is defined as:
+
+\[
+\|f\|_p := \left( \int |f(x)|^p \, \mathrm{d}\mu(x) \right)^{1/p}
+\]
+
+\begin{hol}
+\begin{alltt}
+Theorem seminorm\_normal :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p m f.
+0 < p \(\land\) p \(\ne\) +\(\infty\) \(\Rightarrow\)
+seminorm p m f = \(\int\)\(\sp{+}\) m (\(\lambda\)x. abs (f x) powr p) powr p\(\sp{-}\)\(\sp{1}\)
+\end{alltt}
+\end{hol}
+
+This seminorm forms the basis for all moment bounds and continuity results used in the CLT proof.
+In particular, the finiteness of third moments used in Lyapunov's condition corresponds to membership in $\mathcal{L}^3$.
+
+\paragraph{Hölder’s Inequality.}
+For functions $u \in \mathcal{L}^p(m)$ and $v \in \mathcal{L}^q(m)$ where $1/p + 1/q = 1$, Hölder’s inequality bounds their product:
+\begin{hol}
+\begin{alltt}
+Theorem Hoelder\_inequality :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)m u v p q.
+measure\_space m \(\land\) 0 < p \(\land\) 0 < q \(\land\)
+p\(\sp{-}\)\(\sp{1}\) + q\(\sp{-}\)\(\sp{1}\) = 1 \(\land\)
+u \(\in\) lp\_space p m \(\land\) v \(\in\) lp\_space q m \(\Rightarrow\)
+integrable m (\(\lambda\)x. u x * v x) \(\land\)
+\(\int\) m (\(\lambda\)x. abs (u x * v x)) \(\le\) seminorm p m u * seminorm q m v
+\end{alltt}
+\end{hol}
+
+\paragraph{Minkowski’s Inequality.}
+Also known as the triangle inequality for $\mathcal{L}^p$ norms:
+\begin{hol}
+\begin{alltt}
+Theorem Minkowski\_inequality :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p m u v.
+measure\_space m \(\land\) 1 \(\le\) p \(\land\)
+u \(\in\) lp\_space p m \(\land\) v \(\in\) lp\_space p m \(\Rightarrow\)
+(\(\lambda\)x. u x + v x) \(\in\) lp\_space p m \(\land\)
+seminorm p m (\(\lambda\)x. u x + v x) \(\le\)
+seminorm p m u + seminorm p m v
+\end{alltt}
+\end{hol}
+
+\paragraph{Cauchy–Schwarz Inequality.}
+This is Hölder's inequality for the particular case $p = q = 2$, which assures us that the inner products of square-integrable functions are bounded:
+\begin{hol}
+\begin{alltt}
+Theorem Cauchy\_Schwarz\_inequality :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)m u v.
+measure\_space m \(\land\) u \(\in\) L2\_space m \(\land\) v \(\in\) L2\_space m \(\Rightarrow\)
+integrable m (\(\lambda\)x. u x * v x) \(\land\)
+\(\int\) m (\(\lambda\)x. abs (u x * v x)) \(\le\) seminorm 2 m u * seminorm 2 m v
+\end{alltt}
+\end{hol}
+
+These inequalities are foundational tools in functional analysis.
+In this thesis, only Hölder’s inequality is explicitly used—namely, in the proof of the Lyapunov inequality (`liapounov\_ineq\_lemma`) to relate $L^p$ norms.
+While Minkowski’s and Cauchy–Schwarz inequalities are stated here for completeness, they were not directly invoked in the formalisation.
+The Taylor approximation error in the CLT proof is instead bounded via a custom lemma using Lyapunov’s inequality and third moment control.
+
 \section{Probability Theory}
+
+This section reviews the essential formal components of probability theory used throughout the formalisation of the Central Limit Theorem (CLT) in HOL4.
+We focus on probability spaces, random variables, independence, expectation, variance, and convergence of random sequences—only including results directly used or closely related to our proof.
+
+\subsection{Probability Spaces}
+
+A probability space is a triple $(\Omega, \mathscr{F}, \mathscr{P})$, where $\Omega$ is the sample space, $\mathscr{F}$ is a $\sigma$-algebra of measurable events, and $\mathscr{P}$ is a probability measure satisfying $\mathscr{P}(\Omega) = 1$.
+In HOL4, this is encoded as:
+
+\begin{hol}
+\begin{alltt}
+Definition prob\_space\_def :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p.
+prob\_space p \({\Leftrightarrow}\)
+measure\_space p \(\land\) measure p (m\_space p) = 1
+\end{alltt}
+\end{hol}
+
+All the fundamental components of a probability space—event sets, probability values, and sample spaces—are accessed via:
+
+\[
+\texttt{events p} = \texttt{measurable\_sets p},\quad
+\texttt{prob p} = \texttt{measure p},\quad
+\texttt{p\_space p} = \texttt{m\_space p}
+\]
+
+\subsection{Random Variables and Distributions}
+
+A random variable is a measurable function $X : \Omega \rightarrow \mathbb{R}$ from a probability space to the Borel $\sigma$-algebra. HOL4 defines real-valued random variables as:
+
+\begin{hol}
+\begin{alltt}
+Definition real\_random\_variable\_def :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)X p.
+real\_random\_variable X p \({\Leftrightarrow}\)
+random\_variable X p Borel \(\land\)
+\(\forall\)x. x \(\in\) p\_space p \(\Rightarrow\) X x \(\ne\) -\(\infty\) \(\land\) X x \(\ne\) +\(\infty\)
+\end{alltt}
+\end{hol}
+
+The distribution of a random variable is the pushforward measure of $X$, and the corresponding cumulative distribution function (cdf) is:
+
+\[
+\texttt{distribution\_function p X t} = \texttt{prob p} \{x \in p\_space p | X x \leq t\}
+\]
+
+\subsection{Mathematical Expectation and Moments}
+
+Expectation is defined as Lebesgue integration over a probability space:
+
+\begin{hol}
+\begin{alltt}
+Definition expectation\_def :
+\(\!\!\!{\turn}\!\!\!\!\) expectation = \(\int\)
+\end{alltt}
+\end{hol}
+
+A variable is integrable if its absolute value has finite integral. Variance is defined as the second central moment:
+
+\begin{hol}
+\begin{alltt}
+Theorem variance\_eq :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p X.
+prob\_space p \(\land\) real\_random\_variable X p \(\land\)
+integrable p (\(\lambda\)x. (X x)\(^2\)) \(\Rightarrow\)
+variance p X =
+expectation p (\(\lambda\)x. (X x)\(^2\)) - (expectation p X)\(^2\)
+\end{alltt}
+\end{hol}
+
+This thesis frequently verifies integrability and bounds on moments, e.g., when applying Lyapunov's condition or Taylor approximation.
+
+\subsection{Independence}
+
+Random variables $X_i$ are independent if for all measurable sets $A_i$, their inverse images under $X_i$ form independent events. HOL4 formalises total independence of a sequence via:
+
+\begin{hol}
+\begin{alltt}
+Definition indep\_vars\_def :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p X A J.
+indep\_vars p X A J \({\Leftrightarrow}\)
+\(\forall\)E N.
+  N \(\subseteq\) J \(\land\) N \(\ne\) \(\emptyset\) \(\land\) FINITE N \(\land\)
+  E \(\in\) N \(\Rightarrow\) subsets \(\circ\) A \(\Rightarrow\)
+  prob p (BIGINTER (IMAGE (\(\lambda\)n. PREIMAGE (X n) (E n) \(\cap\) p\_space p) N)) =
+  \(\prod\) (prob p \(\circ\) (\(\lambda\)n. PREIMAGE (X n) (E n) \(\cap\) p\_space p)) N
+\end{alltt}
+\end{hol}
+
+\begin{hol}
+\begin{alltt}
+Definition indep\_rv\_def :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p X Y s t.
+indep\_vars p X Y s t \({\Leftrightarrow}\)
+\(\forall\)a b.
+  a \(\in\) subsets s \(\land\) b \(\in\) subsets t \(\Rightarrow\)
+  indep p (PREIMAGE X a \(\cap\) p\_space p) (PREIMAGE Y b \(\cap\) p\_space p)
+\end{alltt}
+\end{hol}
+
+We use this extensively when constructing auxiliary sequences for the Lindeberg method.
+
+\subsection{Convergence of Random Sequences}
+
+In our formalisation of the Central Limit Theorem, we primarily rely on \emph{convergence in distribution}, which captures the idea that the sequence of random variables $X_n$ becomes increasingly close to a limiting variable $Y$ in terms of their probability distributions.
+
+Formally, convergence in distribution is defined in HOL4 as:
+\begin{hol}
+\begin{alltt}
+Theorem converge\_in\_dist\_alt :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p X Y.
+prob\_space p \(\land\) (\(\forall\)n. real\_random\_variable (X n) p) \(\land\)
+real\_random\_variable Y p \(\Rightarrow\)
+((X \(\longrightarrow\) Y) (in\_distribution p) \({\Leftrightarrow}\)
+ (\(\lambda\)n. distribution p (X n)) \(\longrightarrow\) distribution p Y)
+\end{alltt}
+\end{hol}
+
+This equivalence allows us to reduce convergence in distribution to pointwise convergence of the associated distributions.
+
+A more analytic characterisation—useful in proofs using Taylor approximations and moment estimates—is given by:
+\begin{hol}
+\begin{alltt}
+Theorem converge\_in\_dist\_alt' :
+\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p X Y.
+prob\_space p \(\land\) (\(\forall\)n. real\_random\_variable (X n) p) \(\land\)
+real\_random\_variable Y p \(\Rightarrow\)
+((X \(\longrightarrow\) Y) (in\_distribution p) \({\Leftrightarrow}\)
+ \(\forall\)f. bounded (IMAGE f \(\mathbb{U}\)(:real)) \(\land\)
+        f continuous\_on \(\mathbb{U}\)(:real) \(\Rightarrow\)
+   ((\(\lambda\)n. expectation p (Normal \(\circ\) f \(\circ\) real \(\circ\) X n)) \(\longrightarrow\)
+    expectation p (Normal \(\circ\) f \(\circ\) real \(\circ\) Y)) sequentially)
+\end{alltt}
+\end{hol}
+
+This version states that convergence in distribution is equivalent to the convergence of expectations for all bounded continuous test functions $ f $, which aligns closely with the Portmanteau theorem.
+
+In our CLT proof, this form enables us to focus on expectations of composed functions and verify the convergence using Taylor expansion and moment bounds, without needing pointwise or $ L^p $ convergence of random variables.
+
+\subsection{Summary}
+
+This section outlines only the essential building blocks from HOL4's formalisation of probability theory that are directly used in the formalisation of the CLT using the Lindeberg method.
+In particular, it sets up the notion of independence, variance control, and convergence that underpin the proof strategy.

### 2025-05-22 acknowledgments.tex
Add Abstract and Acknowledgments

\section*{Acknowledgments}

The author want to give special thanks to Dr. \emph{Chun Tian} for his supervision of this thesis and for suggesting the topic of formalizing the Central Limit Theorem. Without Dr. Tian’s generous support, deep insights, and continuous guidance, the author wouldn't have been able to pursue this work within the HOL4 theorem prover and complete the requirements for graduation.

The author also want to thanks to her family for their unconditional love, belief in her, and emotional support, even from afar. Their presence—whether near or distant—gave her strength through every challenge.

To dear friend \emph{Le Manh Tai}, thank you for always being there—for every late night, every deadline, and every quiet moment of doubt. Your support made a difference more than words can say.

Finally, the author would like to thank all her friends in Vietnam, Australia, and around the world who have supported her in different ways. Without all of you, the author wouldn’t have made it this far.

### 2025-05-22 abstract.tex
Minor changes

\chapter*{Abstract}
\label{abstract}

This thesis presents a formalisation of Lyapunov’s Central Limit Theorem (CLT) in the HOL4 theorem prover. The proof follows the Lindeberg replacement method and employs Taylor expansion techniques to quantify the difference between the distribution of normalised partial sums and the standard normal distribution.

The formalisation enriches HOL4’s existing \texttt{ProbabilityTheory} by introducing additional infrastructure necessary for the proof, including higher-order derivatives, bounds on smooth functions, and properties of moment conditions.

A key outcome of this work is a modular and rigorous development of the Lindeberg-based proof of the CLT, which complements existing formal efforts in \texttt{ProbabilityTheory} in HOL4. The approach demonstrates that classical analytical techniques can be accurately mechanised using modern theorem proving tools.

### 2025-05-22 futurework.tex
Add informal explanations and end of chapter conclusion

@@
 \item \textbf{Multivariate CLT.} Extend the proof to sequences of random vectors in \( \mathbb{R}^d \), a standard result in multivariate statistics.
-\item \textbf{Berry–Esseen Bounds.} Provide a quantitative rate of convergence, leveraging the moment bounds and error control machinery developed in this thesis. \cite{berry1941accuracy}
+\item \textbf{Berry–Esseen Bounds.} Provide a quantitative rate of convergence, leveraging the moment bounds and error control machinery developed in this thesis \cite{berry1941accuracy}.
 \end{itemize}

### 2025-05-22 acknowledgments.tex
Minor changes

@@
-\section*{Acknowledgments}
+\chapter*{Acknowledgments}
+\label{ack}
 
@@
 
-The author also want to thanks to her family for their unconditional love, belief in her, and emotional support, even from afar.
+The author also wants to thanks to her family for their unconditional love, belief in her, and emotional support, even from afar.
 Their presence—whether near or distant—gave her strength through every challenge.
 
-To dear friend \emph{Le Manh Tai}, thank you for always being there—for every late night, every deadline, and every quiet moment of doubt. Your support made a difference more than words can say.
+To her dear friend \emph{Le Manh Tai}, thank you for always being there—for every late night, every deadline, and every quiet moment of doubt. Your support made a difference more than words can say.
 

### 2025-05-23 conclusions.tex
Some fixes in conclusions

@@
 This thesis has formalised a general version of the Central Limit Theorem (CLT) in HOL4 under Lyapunov's condition, using the Lindeberg replacement method.
-The version presented here is far more general than the classical i.i.d.
-case: it handles independent (but not identically distributed) summands with varying variances and third absolute moments.
+The version presented here is far more general than the classical I.I.D.~case: it handles independent (but not identically distributed) summands with varying variances and third absolute moments.
 Our development closely follows Chung's analytical presentation~\cite{Chung:2001}, and surpasses earlier formal work such as the CLT in Isabelle/HOL~\cite{serafin2015formally} in both generality and technical depth.
 
-To our knowledge, this is the first mechanised proof of Lyapunov’s CLT in HOL4 for general independent sequences, employing the Lindeberg replacement trick, Taylor expansion, and Big-O reasoning.
-Unlike earlier approaches that assume identical distributions, our formalisation accommodates inhomogeneous sequences, which introduces substantial technical overhead—particularly in bounding variable-specific replacement errors.
+To the best of our knowledge, this is the first mechanised proof of CLT beyond the I.I.D.~cases.
+This version, usually known as the \emph{Lyapunov's form of the CLT}, employs the Lindeberg's method of replacement trick, Taylor expansion, and Big-O reasoning.
+Unlike earlier approaches that assume identical distributions, our formalisation accommodates inhomogeneous sequences, which introduces substantial technical overhead--particularly in bound variables--specific replacement errors.
 However, this effort yields a more widely applicable theorem under Lyapunov’s condition.
 
-Over 6000 lines of HOL4 code were developed to support this formalisation, including infrastructure for moment inequalities, summability, Taylor approximation, and asymptotic analysis.
+Over 6000 lines of HOL4 proof scripts were developed to support this formalisation, including infrastructure for moment inequalities, summability, Taylor approximation, and asymptotic analysis.
 Much of this was written from scratch or adapted from existing theories to suit the strict requirements of probabilistic reasoning in HOL4.
@@
 \begin{itemize}
-    \item Construction of auxiliary sequences of Gaussian random variables;
+    \item Construction of auxiliary sequences of Gaussian (normal) random variables;
     \item Moment and variance control via Lyapunov-type inequalities;

### 2025-05-23 conclusions.tex
More fixes eliminating some unnecessary "fake" math formulas

@@
 This version, usually known as the \emph{Lyapunov's form of the CLT}, employs the Lindeberg's method of replacement trick, Taylor expansion, and Big-O reasoning.
-Unlike earlier approaches that assume identical distributions, our formalisation accommodates inhomogeneous sequences, which introduces substantial technical overhead--particularly in bound variables--specific replacement errors.
+Unlike earlier approaches that assume identical distributions, our formalisation accommodates inhomogeneous sequences, which introduces substantial technical overhead--particularly in bounding variables--specific replacement errors.
 However, this effort yields a more widely applicable theorem under Lyapunov’s condition.
